---
title: "Climate & community"
author: "Volker Rudolf"
date: "2025-06-12"
output:
  html_document: default
  pdf_document: default
---
---
title: "Climate_community"
output: html_document
date: "2025-05-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Goals
We will examine how climate variables are linked to different aspects of community state transitions. THis is focused on frog calling data from 8 different ponds across 22 years. Climate data is taking from local weather stations, and frog calling data was converted to transition matrix indicating how community states (i.e. a specific combination of species) transition over time within a year.


## Load packages & data
```{r, warning=FALSE, message=FALSE}
 #Load packages
  #list of packages that are needed
  pack<-c("tidyverse",
          "knitr",
          "here",
          "car",
          "lme4",
          "flextable",
          "stringr",
          "ggpubr",
          "broom.mixed",
          "lmerTest",
          "rlang",
          "ggrepel"
          )
  #check whether packages are installed and load them, if not get and then load
 package.check <- lapply(pack, FUN = function(x){
    if (!require(x, character.only = TRUE)) 
      {
        install.packages(x, dependencies = TRUE)
        library(x, character.only = TRUE)
      }
    })
   #search() #verify that packages are loaded

```

Let's load the community matrix and climate data (see "Community transition code.Rdm" and "Climate data preparation.Rdm")
```{r get data, message=F}
#get summary metrics of the transition matrix
state_connectance_metrics<-read_rds(here("./output/frog_transition_matrix_metrics.rds"))#raw data transition matrix 
#we have 3 additional version of the transition matrix that ignore a gap of 1, 2, or 3 day in the calling of a species and treat it as a continuous period. 
state_connectance_metrics_g1<-read_rds(here("output","frog_transition_matrix_metrics_g1.rds")) 
state_connectance_metrics_g2<-read_rds(here("output","frog_transition_matrix_metrics_g2.rds"))
state_connectance_metrics_g3<-read_rds(here("output","frog_transition_matrix_metrics_g3.rds"))

#get annual climate data summary metrics 
sumclima<-read_csv(here("output","frog_climate_data_annual_scaled.csv"))%>%subset(cycle<25)

#combine climate and network statistics
netclim <- reduce(list(sumclima, state_connectance_metrics), full_join) %>%
  select(pond, cycle, Pond_Cycle, everything())
netclim_g1 <- reduce(list(sumclima, state_connectance_metrics_g1), full_join) %>%
  select(pond, cycle, Pond_Cycle, everything())
netclim_g2 <- reduce(list(sumclima, state_connectance_metrics_g2), full_join) %>%
  select(pond, cycle, Pond_Cycle, everything())
netclim_g3 <- reduce(list(sumclima, state_connectance_metrics_g3), full_join) %>%
  select(pond, cycle, Pond_Cycle, everything())

```


before we get into analysis lets look at correlation between network metrics
```{r metric correlations}
library(ggcorrplot)
metrics<-state_connectance_metrics%>%select(-pond,-cycle,-Pond_Cycle, -markov_memory)
metrics_g1<-state_connectance_metrics_g1%>%select(-pond,-cycle,-Pond_Cycle, -markov_memory)
metrics_g2<-state_connectance_metrics_g2%>%select(-pond,-cycle,-Pond_Cycle, -markov_memory)
metrics_g3<-state_connectance_metrics_g3%>%select(-pond,-cycle,-Pond_Cycle, -markov_memory)

corr<-round(cor(metrics, use = "complete.obs"),1)
p.mat<-cor_pmat(metrics)
ggcorrplot(corr, hc.order=TRUE, type="upper",lab="TRUE", p.mat=p.mat)+
  ggtitle("Correlation matrix raw data, without ignoring gaps")

corrg1<-round(cor(metrics_g1, use = "complete.obs"),1)
p.mat<-cor_pmat(metrics_g1)
ggcorrplot(corr, hc.order=TRUE, type="upper", lab="TRUE",p.mat=p.mat)+
  ggtitle("Correlation matrix raw ignoring 1 day gaps")

corrg3<-round(cor(metrics_g3, use = "complete.obs"),1)
p.mat<-cor_pmat(metrics_g3)
ggcorrplot(corr, hc.order=TRUE, type="upper", lab="TRUE",p.mat=p.mat)+
  ggtitle("Correlation matrix raw without ignoring 3 day gaps")

#look at some of the strongest correlations
ggplot(metrics, aes(x=Unique_Nodes, y=TotalEdges))+geom_point()+stat_smooth(method="lm")+theme_bw()
ggplot(metrics, aes(x=Weighted_connectance, y=Connectance))+geom_point()+stat_smooth(method="lm")+theme_bw()
ggplot(metrics, aes(x=Weighted_connectance, y=harm.centrality))+geom_point()+stat_smooth(method="lm")+theme_bw()

ggplot(metrics, aes(y=log(Unique_Nodes), x=No_Spec))+geom_point()+stat_smooth(method="lm")+theme_bw()

```
So whether we ignore small gaps has  no effect on correlation between metrics even though metrics differ between those choices. 
We also see metrics we can drop: weighted_connectance and connectance are perfectly correlated, so let's drop weighted_connectance, sd_t and mean_t are highly correlated as well, so we can drop sd_t which is partially included in B. Unique nodes and unique edges are also almost perfectly correlated, but we'll keep it for now since that's interesting.

Let's see how those metrics fall out in a multidimensional space in a PCA to better understand they relationship
```{r networkPCA, message=F}
library("factoextra")
library("FactoMineR")
metricpca<- metrics%>%select(-meanDayOnset)%>%prcomp(scale = TRUE)
print(metricpca)

metricpca_g1<-metrics_g1 %>%select(-meanDayOnset)%>% prcomp(scale = TRUE)
metricpca_g3<- metrics_g3%>% select(-meanDayOnset)%>%prcomp( scale = TRUE)

#get corresponding ponds and year
pca_data <- as.data.frame(metricpca$x)
pca_data$cycle<-state_connectance_metrics$cycle
pca_data$pond<-state_connectance_metrics$pond

#plot PC %explanation which is almost identical for all three data sets.
#fviz_eig(metricpca, col.var="blue")
#fviz_eig(metricpca_g1, col.var="blue")
#fviz_eig(metricpca_g3, col.var="blue")

#plot variables for first 2 PCA
fviz_pca_var(metricpca,
             repel = TRUE
             )+ggtitle("PCA loading not ignoring gaps")

#plot both loading and year values
fviz_pca_biplot(metricpca,
                label="var",
                geom_ind="point",
,                col.var = "black",               
                gradient.cols = c("blue", "green"),
                repel = TRUE,
) +ggtitle("PCA not ignoring gaps")
fviz_pca_biplot(metricpca_g1,
                label="var",
                geom_ind="point",
,                col.var = "black",               
                gradient.cols = c("blue", "green"),
                repel = TRUE,
)+ggtitle("PCA ignoring 1 day gaps")
fviz_pca_biplot(metricpca_g3,
                label="var",
                geom_ind="point",
,                col.var = "black",               
                gradient.cols = c("blue", "green"),
                repel = TRUE,
)+ggtitle("PCA ignoring 3 day gaps")



metricPCA<-cbind(netclim[,1:2],metricpca$x[,1:2])
metricPCA_g1<-cbind(netclim[,1:2],metricpca_g1$x[,1:2])
metricPCA_g3<-cbind(netclim[,1:2],metricpca_g3$x[,1:2])

netPCAclim<-full_join(sumclima,metricPCA)%>%select(pond, cycle, everything())
netPCAclim_g1<-full_join(sumclima, metricPCA_g1)%>%select(pond, cycle, everything())
netPCAclim_g3<-full_join(sumclima, metricPCA_g3)%>%select(pond, cycle, everything())

```
So two PCA are enough to explain a lot of variation in all these variables, and the first explains by far the most. It's also interesting that the correlation between variables are largely robust and don't change a lot if we ignore small 1, - 3 day gaps in calling. 
Based on these relationship, we definitely can drop unique nodes since it's almost perfectly correlated with uniqueEdges, and uniqueEdges seems to be more different from No_species than unique nodes and bit more informative on other aspects of the network. We also don't need  mean_t and st_t which are part of memory and burstiness. Connectance and Weighted_connectance are also highly correlated, and weighted_betweenness is also almost perfectly correlated with  harmonic centrality, so let's drop that one.


#Network structure & Weather 
Instead of using all the different weather variables, we will  use the PCA of climate to avoid colinearity issues among weather variables and reduce the number of preditors. Fortunately only two PCA were really important, which makes this a little bit easier

##PCA analysis
```{r metric_WeatherPCA}
climatePCA<-read_csv(here("./output/frog_WeatherPCA.csv"))
netclimPC<-left_join(netclim,climatePCA)
netclimPC_g1<-left_join(netclim_g1,climatePCA)
netclimPC_g2<-left_join(netclim_g2,climatePCA)
netclimPC_g3<-left_join(netclim_g3,climatePCA)

currentPCdata<-netclimPC_g1

#note, I could test for random pond effect using ranova function of lmerTest package
#i can also use the performance package to quantify effect size as explained variance for random effect.


#get variables to analyze
variable.for.extanalysis<-c("Unique_Nodes","TotalEdges","Connectance","betweenness", "modularity", "mean_Distance", "skewEdge", "ave_Edge_weight","B", "meanDurAv", "freqStateAv", "skewDur", "event.rate", "time_memory")

#scale data to be comparable but first log-transform two variables to improve lmer model fit
currentPCdataS<-currentPCdata%>% 
  mutate(across(c("Connectance", "mean_Distance"), log))

currentPCdataS[variable.for.extanalysis] <- lapply(currentPCdata[variable.for.extanalysis], scale)
#analyze the data

resPCA <- map_dfr(variable.for.extanalysis, function(response) {
  # Fit the mixed model for this response
  model <- lmerTest::lmer(as.formula(paste(response,"~ PC1+PC2 +(1 | pond)")), data = currentPCdataS)
  
  # Tidy model output: fixed effects only, with confidence intervals
  broom.mixed::tidy(model, effects = "fixed", conf.int = TRUE) %>%
    mutate(response = response)%>%
    filter(!str_detect(term,"(Intercept)"))
  }
  )

# Create significance pattern table
sig_tablePCA <- resPCA %>%
  mutate(response=variable_labels[response])%>%
  select(response, term,p.value) %>%
  mutate(p.value = round(p.value, 4)) %>%
  pivot_wider(
    names_from = term, 
    values_from = p.value)

# Print formatted flextable for Word output
sig_tablePCA %>%
  flextable() %>%
  set_caption("Table: Significance Patterns") %>%
  theme_booktabs() %>%        # Adds horizontal lines under header and at bottom
  bold(part = "header") %>%   # Make header bold
  align(align = "center", part = "all") %>%  # Center all text
  autofit()  

```

##PCA analysis effect size plot
```{r}
resPCA_prep <- resPCA %>%
  mutate(significant = p.value < 0.05)

# Add a category column to your data based on your grouping
resPCA_prep <- resPCA_prep %>%
  mutate(category = case_when(
    response %in% c("Unique_Nodes",  "Connectance", 
                    "betweenness", "mean_Distance", "modularity") ~ "Network",
    response %in% c("meanDurAv", "freqStateAv", "skewDur") ~ "Node",
    response %in% c("ave_Edge_weight","TotalEdges", "skewEdge") ~ "Edge",
    response %in% c("event.rate", "B", "time_memory") ~ "Spacing",
    TRUE ~ "Other"
  )) %>%
  # Set category order (top to bottom as they appear in your list)
  mutate(category = factor(category, 
                          levels = c("Network", "Node", "Edge", "Spacing"))) %>%
  # Set response variable order within categories
  mutate(response = factor(response, levels = my_grouped_y_order))


# Create background shading by category (spans entire category height)
background_data <- data.frame(
  category = factor(c("Network", "Node", "Edge", "Spacing"),
                   levels = c("Network", "Node", "Edge", "Spacing")),
  fill_color = c("white", "grey95", "white", "grey95")
) %>%
  # Expand to include both terms
  tidyr::crossing(term = unique(resPCA_prep$term))

# Create the plot with faceting by category
coef.plotPCA <- resPCA_prep %>%
  ggplot(aes(x = estimate, y = response, color = term)) +
  # Add background rectangles that span entire category
  geom_rect(data = background_data,
            aes(fill = fill_color),
            xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf,
            inherit.aes = FALSE, alpha = 1) + 
  scale_fill_identity() +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray50", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high, alpha = significant), 
                 height = 0.0, linewidth = 1) +
  geom_point(aes(alpha = significant), size = 3) +
  scale_alpha_manual(values = c("FALSE" = 0.3, "TRUE" = 1), guide = "none") +
  # Facet by both category (rows) and term (columns)
  facet_grid(category ~ term, 
             scales = "free_y",      # Free y-axis for each category
             space = "free_y",       # Adjust panel height based on number of items
             switch = "y",           # Put category labels on the left
             labeller = labeller(
               term = c("PC1" = "Weather PCA1", "PC2" = "Weather PCA2"),
               category = label_value)) +
  scale_y_discrete(
    drop = TRUE,  # Drop unused factor levels in each facet
    labels = variable_labels) +
  labs(x = "Standardized coefficients and 95% CI", y = "") + 
  scale_color_manual(
    values = c("PC1" = "#33B8CC", "PC2" = "#CCAA33"), 
    guide = "none") +
  theme_pubr(base_size = 16) +
  theme(
    strip.background = element_blank(),  # Remove box around facet labels
    strip.text.x = element_text(size = 18, face = "bold"),
    strip.text.y.left = element_text(size = 16, face = "bold", angle = 90),  # Rotate category labels
    strip.placement = "outside",
    panel.spacing = unit(0.3, "lines"),      # Space between categories
    panel.border = element_blank(),  # Remove boxes around panels
    axis.text.x = element_text(size = 14), 
    axis.text.y = element_text(size = 16), 
    axis.title = element_text(size = 18)
  )

coef.plotPCA
ggsave(here("Figures","PCAWeather.effect.png"),plot=coef.plotPCA,dpi=600,width = 30, height = 15, units = "cm")
```



# Extreme climate
Let's look at differences in community structure metrics between extreme years and normal years.
```{r extremeClimateAnalysis}
#get raw climate data
sumclima.raw<-read.csv(here("./output/frog_climate_data_annualsum.csv"))%>%
  mutate(exthighTemp=factor(exthighTemp), extlongHeat=factor(extlongHeat), extDry=factor(extDry),extDryT=factor(extDryT), extHot=factor(extHot))%>%
  select(-burstinessRain)

netclimCur<-netclim #can be netclim, netclime_g1, _g2, _g3 based on gap data set
#join data
netclimR<-left_join(netclimCur,sumclima.raw, by=c("cycle"))


netclimR[variable.for.extanalysis] <- lapply(netclimR[variable.for.extanalysis], scale)

#analyze the data

res <- map_dfr(variable.for.extanalysis, function(response) {
  # Fit the mixed model for this response
  model <- lmer(as.formula(paste(response,"~ extHot+extDryT + (1 | pond)")), data = netclimR)
  
  # Tidy model output: fixed effects only, with confidence intervals
  broom.mixed::tidy(model, effects = "fixed", conf.int = TRUE) %>%
    mutate(response = response)%>%filter(!str_detect(term,"(Intercept)"))
  }
  )

```

Plot effect sizes of extreme drought and temperature
```{r}
# Add a category column to your data based on your grouping
res_prepE <- res %>%
  filter(!response %in% c("No_Spec")) %>%  # Explicitly drop species richness
   mutate(significant = p.value < 0.05)%>%
   mutate(category = case_when(
    response %in% c("Unique_Nodes", "Connectance", 
                    "betweenness", "mean_Distance", "modularity") ~ "Network",
    response %in% c("meanDurAv", "freqStateAv", "skewDur") ~ "Node",
    response %in% c("ave_Edge_weight",  "TotalEdges","skewEdge") ~ "Edge",
    response %in% c("event.rate", "B", "time_memory") ~ "Spacing",
    TRUE ~ "Other"
  )) %>%
  # Set category order (top to bottom as they appear in your list)
  mutate(category = factor(category, 
                          levels = c("Network", "Node", "Edge", "Spacing"))) %>%
  # Set response variable order within categories
  mutate(response = factor(response, levels = my_grouped_y_order))


# Create background shading by category (spans entire category height)
background_data <- data.frame(
  category = factor(c("Network", "Node", "Edge", "Spacing"),
                   levels = c("Network", "Node", "Edge", "Spacing")),
  fill_color = c("white", "grey95", "white", "grey95")
) %>%
  # Expand to include both terms
  tidyr::crossing(term = unique(res_prepE$term))
library(ggtext)
# Create the plot with faceting by category
coef.plotE <- res_prepE %>%
  ggplot(aes(x = estimate, y = response, color = term)) +
  # Add background rectangles that span entire category
  geom_rect(data = background_data,
            aes(fill = fill_color),
            xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf,
            inherit.aes = FALSE, alpha = 1) + 
  scale_fill_identity() +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray50", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high, alpha = significant), 
                 height = 0.0, linewidth = 1) +
  geom_point(aes(alpha = significant), size = 3) +
  scale_alpha_manual(values = c("FALSE" = 0.3, "TRUE" = 1), guide = "none") +
  # Facet by both category (rows) and term (columns)
  facet_grid(category ~ term, 
             scales = "free_y", 
             space = "free_y", 
             switch = "y", 
             labeller=labeller(term = c(
               "extDryTYes" = "Extreme Dry",
               "extHotYes" = "Extreme Hot"),
               category = label_value)) +
   scale_y_discrete(
    drop = TRUE,  # Drop unused factor levels in each facet
    labels = variable_labels) +
  labs(x = "Standardized coefficients and 95% CI", y = "") + 
  scale_color_manual(
    values = c("extDryTYes" = "#A65F32", 
               "extHotYes" = "#CC3333"), 
    guide = "none") +
  theme_pubr(base_size = 16) +
 theme(
    strip.background = element_blank(),
    # Modify strip.text.x to underline column (term) labels
    strip.text.x = element_text(size = 18, face = "bold"),
    # Modify strip.text.y.left to underline row (category) labels
    strip.text.y.left = element_text(size = 16, face = "bold", angle = 90), 
    strip.placement = "outside",
    panel.spacing = unit(0.3, "lines"),
    panel.border = element_blank(),
    axis.text.x = element_text(size = 14), 
    axis.text.y = element_text(size = 16), 
    axis.title = element_text(size = 18)
  )

coef.plotE
ggsave(here("Figures","ExtremeClim.effect.png"),plot=coef.plotE,dpi=600,width = 30, height = 15, units = "cm")

```