---
title: "Climate data preparation"
author: "Volker Rudolf"
date: "2025-06-12"
output: html_document
---
---
title: "Climate"
output: html_document
date: "2025-05-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Goals
The code here prepares the climate data from our frog pond location for general analysis, and looks at some climate patterns 

# Load packages
```{r, warning=FALSE, message=FALSE}
 #Load packages
  #list of packages that are needed
  pack<-c("tidyverse",
          "knitr",
          "here",
          "lubridate",
          "paletteer"
          )
  #check whether packages are installed and load them, if not get and then load
 package.check <- lapply(pack, FUN = function(x){
    if (!require(x, character.only = TRUE)) 
      {
        install.packages(x, dependencies = TRUE)
        library(x, character.only = TRUE)
      }
    })
   #search() #verify that packages are loaded

```

# Temperature and precipitation data
```{r formatdata}
#get and organize raw daily climate observations from weather station
climate<-read_csv(here("data","Climatedata2011_2024.csv"))%>%select(DATE,PRCP,TAVG,TMAX,TMIN)
climate$DATE<-as.Date(climate$DATE,format="%m/%d/%Y")
climate<-climate%>%mutate(year = year(DATE), month=month(DATE), day=day(DATE),doy=as.numeric(format(DATE, "%j")))%>%subset(year<2025)

#now this is usually on annual basis, but we need to align it with the "frog cycle" we used for the community analysis
climate<-climate%>%mutate(cycle=case_when(
  doy>340~year-1999,
  .default=year-2000
))
climate<-climate%>%mutate(cycle_day=case_when(
  doy>=339~doy-338,
  .default=doy+27
))

sumclimaY<-climate%>%group_by(cycle)%>%
  summarize(meanMaxT=mean(TMAX,na.rm = TRUE), 
            meanMinT=mean(TMIN,na.rm = TRUE),
            day100=sum(TMAX > 35,na.rm = TRUE),
            meanPrec=mean(PRCP,na.rm=TRUE),
            cumPrec=sum(PRCP, na.rm=TRUE))
```

Look at some of this weather data plotted
```{r}
ggplot(subset(sumclimaY,cycle<25),aes(x=cycle,y=meanMaxT))+
  geom_line()+
  geom_point()+
  theme_bw()+
  geom_hline(aes(yintercept=25.975),linetype="dashed")

ggplot(subset(sumclimaY,cycle<25),aes(x=cycle,y=day100))+
  geom_line()+
  geom_point()+
  theme_bw()+
  geom_hline(aes(yintercept=37.6),linetype="dashed")

ggplot(subset(sumclimaY,cycle<25),aes(x=cycle,y=meanMinT))+
  geom_line()+
  geom_point()+
  theme_bw()+
  geom_hline(aes(yintercept=13.678),linetype="dashed")

ggplot(subset(sumclimaY,cycle<25),aes(x=cycle,y=meanPrec))+
  geom_line()+
  geom_point()+
  theme_bw()+
  geom_hline(aes(yintercept=3.545489),linetype="dashed")


sumclimaM<-climate%>%group_by(cycle, month)%>%
  summarize(MmeanMaxT=mean(TMAX,na.rm = TRUE), 
            MmeanMinT=mean(TMIN,na.rm = TRUE), 
            Mday100=sum(TMAX > 35,na.rm = TRUE),
            MmeanPrec=mean(PRCP,na.rm=TRUE),
            McumPrec=sum(PRCP, na.rm=TRUE))

```

Lets claculate how bursty rain fall is.
```{r burstiness}
# Define a function to calculate burstiness
calculate_burstiness <- function(dates) {
  if (length(dates) < 2) return(NA)  # Not enough rainy days
  inter_event_times <- as.numeric(diff(sort(dates)))
  mu <- mean(inter_event_times)
  sigma <- sd(inter_event_times)
  B<-if ((mu + sigma) == 0) NA else (sigma - mu) / (sigma + mu)
  data.frame(
    burstinessRain=B,
    Rain_inter_event_time=mu
  )
}

# Filter to only rainy days
rainy_days <- climate %>% filter(PRCP > 0)

# Group by year and compute burstiness
burstiness_by_cycle <- rainy_days %>%
  group_by(cycle) %>%
  summarise(
    stats = list(calculate_burstiness(DATE)))%>%
  unnest(cols=c(stats))

ggplot(burstiness_by_cycle, aes(x=cycle,y=burstinessRain))+geom_point()+theme_bw()
ggplot(burstiness_by_cycle, aes(x=cycle,y=Rain_inter_event_time))+geom_point()+theme_bw()
ggplot(burstiness_by_cycle, aes(x=burstinessRain,y=Rain_inter_event_time))+geom_point()+geom_smooth()+theme_bw()


#add it to yearly climate data
sumclimaYB<-left_join(sumclimaY,burstiness_by_cycle)
```
##drought data
This is data on the drought index, which is calculated based on "missing" rain. It comes in range of severeness categories, so not as continuous values.
```{r drought}

drought<-read_csv(here("data","DroughtDataDCForest.csv"))%>%select(ValidStart,None,D0,D1,D2,D3,D4)
drought$DATE<-as.Date(drought$ValidStart,"%m/%d/%Y")
drought<-drought%>%mutate(year = year(DATE), month=month(DATE), day=day(DATE),doy=as.numeric(format(DATE, "%j")))%>%subset(year<2025)
drought<-drought%>%mutate(cycle=case_when(
  doy>340~year-1999,
  .default=year-2000
))
drought<-drought%>%mutate(cycle_day=case_when(
  doy>=339~doy-338,
  .default=doy+27
))

#combine the severity indexes into one drought severity metric. We do this by giving more weight to more sever drought indexes
drought<-drought%>%mutate(Dseverity=D0+D1*2+D2*3+D3*4+D4*4)
sumdrought<-drought%>%group_by(cycle)%>%
  summarize(Dsevmean=mean(Dseverity),Dext=max(Dseverity))
ggplot(sumdrought, aes(x=cycle, y=Dext))+geom_line()+geom_point()+theme_bw()


#add it to climate data
sumclimaYBD<-left_join(sumclimaYB,sumdrought)%>%subset(cycle<25)

#normalize the climate data since they are all on different scales.
scaledsumclimaY <- data.frame(
  cycle=sumclimaYBD[1],                      # keep the first column untouched
  scale(sumclimaYBD[, -1])              # scale all other columns
)

write_csv(sumclimaYBD, here("output","frog_climate_data_annualsum.csv"))
write_csv(scaledsumclimaY, here("output","frog_climate_data_annual_scaled.csv"))

```

PCA of weather data
```{r}
sumclimaY<-read.csv(here("output","frog_climate_data_annualsum.csv"))
scaledsumclimaY<-read.csv(here("output","frog_climate_data_annual_scaled.csv"))

library("factoextra")

weather.pca <- prcomp(scaledsumclimaY[,-1], scale = TRUE)
print(weather.pca)

#plot PC %explanation
fviz_eig(weather.pca, col.var="blue")

#plot variables for first 2 PCA
fviz_pca_var(weather.pca,
             repel = TRUE
             )

#plot cycles in PCA space
cycle_values <- scaledsumclimaY[, 1]
cycl <- get_pca_ind(weather.pca)
cycl

pcavalues<-fviz_pca_ind(weather.pca,
             col.ind = 1:nrow(scaledsumclimaY),  # Just test indexing
             gradient.cols = c("blue", "green"),
             repel = TRUE
)

#plot both loading and year values
fviz_pca_biplot(weather.pca,
                habillage = scaledsumclimaY[, 1],
                col.var = "black",               
                gradient.cols = c("blue", "green"),
                repel = TRUE,
)


#since PC1 and PC2 explain most, let's save both in new file for later analysis
#Extract PC1 and PC2 scores
pc_scores <- as.data.frame(weather.pca$x[, 1:2])
colnames(pc_scores) <- c("PC1", "PC2")            # Rename for clarity

# Add cycle information back
pc_scores$cycle <- scaledsumclimaY$cycle

#Reorder columns: cycle, PC1, PC2
pc_output <- pc_scores %>%select(cycle, PC1, PC2)

# Save to CSV
write.csv(pc_output, here("output","frog_WeatherPCA.csv"), row.names = FALSE)
```

So based on this, the most unusual cycles are 11 and 23 and 12, 22 (driest & hottest in order), 15 (dry but cold). 4 was the wettest cycle, and 5 is borderline but  shows up in several "extreme hot" categories


Now finally let's look at some correlations of climate variables
```{r climate metric correlations}
library(ggcorrplot)
sumclimaYNoC<-scaledsumclimaY%>%subset(cycle<24)%>%select(-cycle)
corr<-round(cor(sumclimaYNoC),1)
p.mat<-cor_pmat(sumclimaYNoC)
ggcorrplot(corr, hc.order=TRUE, type="lower", p.mat=p.mat)

ggplot(sumclimaYNoC, aes(x=meanMaxT, y=meanMinT))+geom_point()+stat_smooth(method="lm")+theme_bw()
ggplot(sumclimaYNoC, aes(x=cumPrec, y=meanPrec))+geom_point()+stat_smooth(method="lm")+theme_bw()
ggplot(sumclimaYNoC, aes(x=meanMaxT, y=meanPrec))+geom_point()+stat_smooth(method="lm")+theme_bw()
ggplot(sumclimaYNoC, aes(x=meanMinT, y=meanPrec))+geom_point()+stat_smooth(method="lm")+theme_bw()
ggplot(sumclimaYNoC, aes(x=Dext, y=meanPrec))+geom_point()+stat_smooth(method="lm")+theme_bw()
ggplot(sumclimaYNoC, aes(x=Dsevmean, y=meanPrec))+geom_point()+stat_smooth(method="lm")+theme_bw()
ggplot(sumclimaYNoC, aes(x=Dext, y=Dsevmean))+geom_point()+stat_smooth(method="lm")+theme_bw()
ggplot(sumclimaYNoC, aes(x=meanMaxT, y=Dext))+geom_point()+stat_smooth(method="lm")+theme_bw()
ggplot(sumclimaYNoC, aes(x=meanMaxT, y=Dsevmean))+geom_point()+stat_smooth(method="lm")+theme_bw()

ggplot(sumclimaYNoC, aes(x=meanMaxT, y=burstinessRain, color=meanMinT))+geom_point()+stat_smooth(method="lm")+theme_bw()
ggplot(sumclimaYNoC, aes(x=meanMinT, y=burstinessRain))+geom_point()+stat_smooth(method="lm")+theme_bw()
ggplot(sumclimaYNoC, aes(x=meanPrec, y=burstinessRain, color=meanMinT))+geom_point()+stat_smooth(method="lm")+theme_bw()

```
so based on this there are some unsurprisingly strong correlations, and we should not include comPrec & meanPrec, but pick one. Also meanMaxT and day100 are highly correlated. And meanMaxT has some correlation with drought and precipitation but those are different enough that we can't pick one. So let's stick with this list: meanMaxT, meanMinT, meanPrec, burstinessRain

```

