---
title: "Climate & community"
author: "Volker Rudolf"
date: "2025-06-12"
output: html_document
---
---
title: "Climate_community"
output: html_document
date: "2025-05-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Goals
We will examine how climate variables are linked to different aspects of community state transitions. THis is focused on frog calling data from 8 different ponds across 22 years. Climate data is taking from local weather stations, and frog calling data was converted to transition matrix indicating how community states (i.e. a specific combination of species) transition over time within a year.


# Load packages
```{r, warning=FALSE, message=FALSE}
 #Load packages
  #list of packages that are needed
  pack<-c("tidyverse",
          "knitr",
          "here",
          "car",
          "lme4"
          )
  #check whether packages are installed and load them, if not get and then load
 package.check <- lapply(pack, FUN = function(x){
    if (!require(x, character.only = TRUE)) 
      {
        install.packages(x, dependencies = TRUE)
        library(x, character.only = TRUE)
      }
    })
   #search() #verify that packages are loaded

```

Let's load the community matrix and climate data (see "Community transition code.Rdm" and "Climate data preparation.Rdm")
```{r get data, message=F}
#get summary metrics of the transition matrix
state_connectance_w<-read_csv("output/frog_transition_matrix_metrics.csv") #raw data transition matrix 
#we have 3 additional version of the transition matrix that ignore a gap of 1, 2, or 3 day in the calling of a species and treat it as a continuous period. 
state_connectance_w_g1<-read_csv("output/frog_transition_matrix_metrics_g1.csv") 
state_connectance_w_g2<-read_csv("output/frog_transition_matrix_metrics_g2.csv")
state_connectance_w_g3<-read_csv("output/frog_transition_matrix_metrics_g3.csv")

#get annual climate data summary metrics 
sumclima<-read_csv("output/frog_climate_data_annual_scaled.csv")%>%subset(cycle<23)

#combine climate and network statistics
netclim <- reduce(list(sumclima, state_connectance_w), full_join) %>%
  select(pond, cycle, Pond_Cycle, everything())
netclim_g1 <- reduce(list(sumclima, state_connectance_w_g1), full_join) %>%
  select(pond, cycle, Pond_Cycle, everything())
netclim_g2 <- reduce(list(sumclima, state_connectance_w_g2), full_join) %>%
  select(pond, cycle, Pond_Cycle, everything())
netclim_g3 <- reduce(list(sumclima, state_connectance_w_g3), full_join) %>%
  select(pond, cycle, Pond_Cycle, everything())

```


before we get into analysis lets look at correlation between network metrics
```{r metric correlations}
library(ggcorrplot)
metrics<-state_connectance_w%>%select(-pond,-cycle,-Pond_Cycle, -markov_memory)
metrics_g1<-state_connectance_w_g1%>%select(-pond,-cycle,-Pond_Cycle, -markov_memory)
metrics_g2<-state_connectance_w_g2%>%select(-pond,-cycle,-Pond_Cycle, -markov_memory)
metrics_g3<-state_connectance_w_g3%>%select(-pond,-cycle,-Pond_Cycle, -markov_memory)

corr<-round(cor(metrics, use = "complete.obs"),1)
p.mat<-cor_pmat(metrics)
ggcorrplot(corr, hc.order=TRUE, type="upper",lab="TRUE", p.mat=p.mat)

corrg1<-round(cor(metrics_g1, use = "complete.obs"),1)
p.mat<-cor_pmat(metrics_g1)
ggcorrplot(corr, hc.order=TRUE, type="upper", lab="TRUE",p.mat=p.mat)

corrg3<-round(cor(metrics_g3, use = "complete.obs"),1)
p.mat<-cor_pmat(metrics_g3)
ggcorrplot(corr, hc.order=TRUE, type="upper", lab="TRUE",p.mat=p.mat)

ggplot(metrics_g1, aes(x=No_Spec, y=Unique_Nodes))+geom_point()+stat_smooth(method="lm")+theme_bw()
ggplot(metrics_g1, aes(x=Unique_Nodes, y=totalEdges))+geom_point()+stat_smooth(method="lm")+theme_bw()

```
 
let's see how those metrics fall out in a multidimensional space in a PCA to better understand they relationship
```{r networkPCA}
library("factoextra")
library("FactoMineR")
metricpca<- prcomp(metrics, scale = TRUE)
print(metricpca)

metricpca_g1<- prcomp(metrics_g1, scale = TRUE)
metricpca_g3<- prcomp(metrics_g3, scale = TRUE)


#plot PC %explanation
fviz_eig(metricpca, col.var="blue")
fviz_eig(metricpca_g1, col.var="blue")
fviz_eig(metricpca_g3, col.var="blue")

#plot variables for first 2 PCA
fviz_pca_var(metricpca,
             repel = TRUE
             )

#plot both loading and year values
fviz_pca_biplot(metricpca,
                label="var",
                geom_ind="point",
,                col.var = "black",               
                gradient.cols = c("blue", "green"),
                repel = TRUE,
)
fviz_pca_biplot(metricpca_g1,
                label="var",
                geom_ind="point",
,                col.var = "black",               
                gradient.cols = c("blue", "green"),
                repel = TRUE,
)
fviz_pca_biplot(metricpca_g3,
                label="var",
                geom_ind="point",
,                col.var = "black",               
                gradient.cols = c("blue", "green"),
                repel = TRUE,
)



metricPCA<-cbind(netclim[,1:2],metricpca$x[,1:2])
metricPCA_g1<-cbind(netclim[,1:2],metricpca_g1$x[,1:2])
metricPCA_g3<-cbind(netclim[,1:2],metricpca_g3$x[,1:2])

dfs_pca<-list(sumclima, metricPCA, burst_mem)
netPCAclim<-reduce(dfs_pca, full_join)%>%select(pond, cycle, everything())

dfs_pcag1<-list(sumclima, metricPCA_g1, burst_mem)
netPCAclim_g1<-reduce(dfs_pcag1, full_join)%>%select(pond, cycle, everything())

dfs_pcag3<-list(sumclima, metricPCA_g3, burst_mem)
netPCAclim_g3<-reduce(dfs_pcag3, full_join)%>%select(pond, cycle, everything())


```
So two PCA are enough to explain a lot of variation in all these variables, and the first explains by far the most. It's also interesting that the correlation between variables are robust and don't change if we ignore small 1, 2, or 3 day gaps. 
Based on these relationship, we definitely want to keep No_Spec as covariate when we analyze the other metrics, and there is no point in including totalEdges and  Unique nodes since both are so correlated, and total Edges seems to be more different from No species than unique nodes and bit more informative on other aspects of the network.


#Network structure & climate analysis
Ok so let's analyze the network metrics and how they are linked to climate variables. We have to options: (1) use PC1 and PC2 in the analysis which strongly reduces the number of analyses we need. But it also is hard to interpete, and can miss more nuanced relationship of individual variables. So we will do both, PC based GLMM and individual network metric GLMM analyses
```{r, GLMM network metric climate analysis}
#GLMM models
currentdata=netclimPC_g1

mS <- lmer(No_Spec~meanMinT*meanMaxT*meanPrec*burstinessRain+(1|pond)+cycle, data=currentdata)
mb <- lmer(betweenness~meanMinT*meanMaxT*meanPrec*burstinessRain+No_Spec+(1|pond)+cycle, data=currentdata)
mc <- lmer(Connectance~meanMinT*meanMaxT*meanPrec*burstinessRain+No_Spec+(1|pond)+cycle, data=currentdata)
mE <- lmer(log(totalEdges)~meanMinT*meanMaxT*meanPrec*burstinessRain+No_Spec+(1|pond)+cycle, data=currentdata)
mm <- lmer(modularity~meanMinT*meanMaxT*meanPrec*burstinessRain+No_Spec+(1|pond)+cycle, data=currentdata)
mD <- lmer((mean_Distance)~meanMinT*meanMaxT*meanPrec*burstinessRain+No_Spec+(1|pond)+cycle, data=currentdata)
mcen <- lmer(harm.centrality~meanMinT*meanMaxT*meanPrec*burstinessRain+No_Spec+(1|pond)+cycle, data=currentdata)
mBurst <- lmer(B~meanMinT*meanMaxT*meanPrec*burstinessRain+No_Spec+(1|pond)+cycle, data=currentdata)
mMem <- lmer(time_memory~meanMinT*meanMaxT*meanPrec*burstinessRain+No_Spec+(1|pond)+cycle, data=currentdata)

# Create a list of models with descriptive names
models <- list(
  "No_Species" = mS,
  "Betweenness" = mb,
  "Connectance" = mc,
  "Total_Edges" = mE,
  "Modularity" = mm,
  "Mean_Distance" = mD,
  "Harmonic_Centrality" = mcen,
  "Burstiness" = mBurst,
  "Memory"= mMem
)

# Function to extract ANOVA results and coefficient information
extract_anova_results <- function(model, model_name) {
  anova_result <- Anova(model)
  coef_summary <- summary(model)$coefficients
  
  # Get fixed effects names (excluding random effects)
  fixed_effects <- rownames(coef_summary)
  anova_effects <- rownames(anova_result)
  
  # Initialize vectors for effect sizes and standard errors
  effect_sizes <- rep(NA, length(anova_effects))
  std_errors <- rep(NA, length(anova_effects))
  
  # For each ANOVA effect, find corresponding coefficient(s)
  for(i in 1:length(anova_effects)) {
    effect_name <- anova_effects[i]
    
    # Find matching coefficients
    matching_coefs <- grep(effect_name, fixed_effects, value = TRUE)
    
    # If no direct match, try alternative matching patterns
    if(length(matching_coefs) == 0) {
      if(grepl(":", effect_name)) {
        # Handle interactions
        terms <- strsplit(effect_name, ":")[[1]]
        matching_coefs <- fixed_effects[sapply(fixed_effects, function(x) {
          all(sapply(terms, function(term) grepl(term, x)))
        })]
      } else {
        # Single term
        matching_coefs <- fixed_effects[grepl(effect_name, fixed_effects)]
      }
    }
    
    # Take the first matching coefficient if multiple found
    if(length(matching_coefs) > 0) {
      primary_coef <- matching_coefs[1]
      coef_index <- which(fixed_effects == primary_coef)
      if(length(coef_index) > 0) {
        effect_sizes[i] <- round(coef_summary[coef_index, "Estimate"], 4)
        std_errors[i] <- round(coef_summary[coef_index, "Std. Error"], 4)
      }
    }
  }
  
  # Create results data frame
  results_df <- data.frame(
    Model = model_name,
    Effect = anova_effects,
    Effect_Size = effect_sizes,
    Std_Error = std_errors,
    Chisq = round(anova_result$Chisq, 3),
    Df = anova_result$Df,
    P_value = round(anova_result$`Pr(>Chisq)`, 4),
    Significance = ifelse(anova_result$`Pr(>Chisq)` < 0.001, "***",
                         ifelse(anova_result$`Pr(>Chisq)` < 0.01, "**",
                               ifelse(anova_result$`Pr(>Chisq)` < 0.05, "*",
                                     ifelse(anova_result$`Pr(>Chisq)` < 0.1, ".", ""))))
  )
  
  return(results_df)
}

# Apply the function to all models
all_results <- do.call(rbind, mapply(extract_anova_results, models, names(models), SIMPLIFY = FALSE))

# Reset row names
rownames(all_results) <- NULL

# Display the complete results table
print("Complete ANOVA Results for All Models:")
print(all_results)

# Save to CSV
#write.csv(all_results, "combined_anova_results.csv", row.names = FALSE)

# Create a wide format table showing p-values for each effect across models
pivot_table <- all_results %>%
  select(Model, Effect, P_value) %>%
  pivot_wider(names_from = Model, values_from = P_value) %>%
  arrange(Effect)

print("P-values across all models:")
print(pivot_table)

# Create effect size comparison table
effect_size_table <- all_results %>%
  select(Model, Effect, Effect_Size) %>%
  pivot_wider(names_from = Model, values_from = Effect_Size) %>%
  arrange(Effect)

print("Effect sizes across all models:")
print(effect_size_table)

# Create standard error comparison table
std_error_table <- all_results %>%
  select(Model, Effect, Std_Error) %>%
  pivot_wider(names_from = Model, values_from = Std_Error) %>%
  arrange(Effect)

print("Standard errors across all models:")
print(std_error_table)

# Create significance pattern table with proper sorting
sig_table <- all_results %>%
  select(Model, Effect, Significance) %>%
  # Add interaction level column (count colons in Effect names)
  mutate(interaction_level = str_count(Effect, ":")) %>%
  # Sort by interaction level first, then alphabetically within each level
  arrange(interaction_level, Effect) %>%
  # Remove the helper column and create wide format
  select(-interaction_level) %>%
  pivot_wider(names_from = Model, values_from = Significance, values_fill = "")

print(sig_table)
```
Now let's do it with the PC instead 
```{r, PCmetric climate analysis}
pca1<-lmer(PC1~meanMinT+
             meanMaxT+
             meanPrec+
             burstinessRain+
             meanMinT:meanMaxT+
             meanMinT:meanPrec+
             meanMinT:burstinessRain+
             meanMaxT:burstinessRain+
             meanPrec:burstinessRain+
             meanMinT:meanMaxT:burstinessRain+
             #meanMinT:meanMaxT:burstinessRain:meanPrec+
             (1|pond)+cycle, 
           data=netPCAclim)
pca1_g1<-lmer(PC1~meanMinT+
             meanMaxT+
             meanPrec+
             burstinessRain+
             meanMinT:meanMaxT+
             meanMinT:meanPrec+
             meanMinT:burstinessRain+
             meanMaxT:burstinessRain+
             meanPrec:burstinessRain+
             meanMinT:meanMaxT:burstinessRain+
             #meanMinT:meanMaxT:burstinessRain:meanPrec+
             +(1|pond)+cycle,
             data=netPCAclim_g1)
pca1_g3<-lmer(PC1~meanMinT+
             meanMaxT+
             meanPrec+
             burstinessRain+
             meanMinT:meanMaxT+
             meanMinT:meanPrec+
             meanMinT:burstinessRain+
             meanMaxT:burstinessRain+
             meanPrec:burstinessRain+
             meanMinT:meanMaxT:burstinessRain+
             #meanMinT:meanMaxT:burstinessRain:meanPrec+
             (1|pond)+cycle,
             data=netPCAclim_g3)

Anova(pca1)
Anova(pca1_g1)
Anova(pca1_g3)


pca2<-lmer(PC2~meanMinT*meanMaxT*meanPrec*burstinessRain+(1|pond)+cycle, data=netPCAclim)
pca2_g1<-lmer(PC2~meanMinT*meanMaxT*meanPrec*burstinessRain+(1|pond), data=netPCAclim_g1)
pca2_g3<-lmer(PC2~meanMinT*meanMaxT*meanPrec*burstinessRain+(1|pond)+cycle, data=netPCAclim_g3)
Anova(pca2)
Anova(pca2_g1)
Anova(pca2_g3)

```
Let's make some plots
```{r}
# Prepare data for plotting - remove rows with missing effect sizes
plot_data <- all_results %>%
  filter(!is.na(Effect_Size) & !is.na(Std_Error)) %>%
  mutate(
    # Create confidence intervals (mean ± 1 SE)
    CI_lower = Effect_Size - Std_Error,
    CI_upper = Effect_Size + Std_Error,
    # Create a combined label for better readability
    Effect_Model = paste(Effect, "-", Model, sep = " "),
    # Add significance indicator for point shapes
    Significant = ifelse(P_value < 0.05, "Significant", "Non-significant"),
    # Count interaction level and create ordering
    interaction_level = str_count(Effect, ":"),
    # Create ordering factor for effects
    effect_order = case_when(
      interaction_level == 0 ~ paste("1_Main:", Effect),
      interaction_level == 1 ~ paste("2_Two-way:", Effect),
      interaction_level == 2 ~ paste("3_Three-way:", Effect),
      interaction_level == 3 ~ paste("4_Four-way:", Effect),
      TRUE ~ paste("5_Other:", Effect)
  )
  ) %>%
  # Reorder Effect factor based on interaction level, then alphabetically within level
  arrange(-interaction_level, Effect) %>%
  mutate(Effect = factor(Effect, levels = unique(Effect)))


# Create the main effect size plot
effect_plot <- ggplot(plot_data, aes(x = Effect_Size, y = Effect, color = Model)) +
  # Add vertical line at zero for reference
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", alpha = 0.7) +
  
  # Add error bars (whiskers)
  geom_errorbarh(aes(xmin = CI_lower, xmax = CI_upper), 
                 height = 0.3, alpha = 0.7, size = 0.8) +
  
  # Add points for effect sizes
  geom_point(aes(shape = Significant), size = 3, alpha = 0.8) +
  
  # Customize scales
  scale_shape_manual(values = c("Significant" = 16, "Non-significant" = 1)) +
  scale_color_viridis_d(name = "Model") +
  
  # Labels and theme
  labs(
    title = "Effect Sizes Across Models",
    subtitle = "Points show coefficient estimates, whiskers show ± 1 Standard Error",
    x = "Effect Size (Coefficient Estimate)",
    y = "Fixed Effects",
    shape = "Significance (p < 0.05)"
  ) +
  
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "right",
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  ) +
  
  # Add better spacing
  theme(plot.margin = margin(20, 20, 20, 20))

# Display the plot
print(effect_plot)

# Alternative version: Faceted by model for clearer comparison
faceted_plot <- ggplot(plot_data, aes(x = Effect_Size, y = Effect)) +
  # Add vertical line at zero
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", alpha = 0.7) +
  
  # Add error bars
  geom_errorbarh(aes(xmin = CI_lower, xmax = CI_upper, color = Significant), 
                 height = 0.2, size = 0.8) +
  
  # Add points
  geom_point(aes(color = Significant, shape = Significant), size = 2.5, alpha = 0.8) +
  
  # Facet by model
  facet_wrap(~ Model, scales = "free_x", ncol = 3) +
  
  # Customize scales
  scale_color_manual(values = c("Significant" = "#E31A1C", "Non-significant" = "#377EB8")) +
  scale_shape_manual(values = c("Significant" = 16, "Non-significant" = 1)) +
  
  # Labels and theme
  labs(
    title = "Effect Sizes by Model",
    subtitle = "Points show coefficient estimates, whiskers show ± 1 Standard Error",
    x = "Effect Size (Coefficient Estimate)",
    y = "Fixed Effects",
    color = "Significance",
    shape = "Significance"
  ) +
  
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 10),
    legend.position = "bottom",
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(size = 9),
    axis.title = element_text(size = 11)
  )

print("Faceted plot:")
print(faceted_plot)

# Create a summary table of effect sizes
summary_stats <- plot_data %>%
  group_by(Effect) %>%
  summarise(
    n_models = n(),
    mean_effect = round(mean(Effect_Size, na.rm = TRUE), 4),
    median_effect = round(median(Effect_Size, na.rm = TRUE), 4),
    min_effect = round(min(Effect_Size, na.rm = TRUE), 4),
    max_effect = round(max(Effect_Size, na.rm = TRUE), 4),
    n_significant = sum(Significant == "Significant"),
    prop_significant = round(n_significant / n_models, 2),
    .groups = "drop"
  ) %>%
  arrange(desc(abs(mean_effect)))

print("Summary of effect sizes across models:")
print(summary_stats)

# Save plots
#ggsave("effect_sizes_plot.png", effect_plot, width = 12, height = 8, dpi = 300)
#ggsave("effect_sizes_faceted.png", faceted_plot, width = 14, height = 10, dpi = 300)


```

Look at extreme years
```{r}
#get raw climate data
sumclimaY<-read.csv("output/frog_climate_data_annualsum.csv")

#define extreme years see "climate data preparation" code for cutoffs
#hot years
sumclimaY$exthighTemp<-ifelse(sumclimaY$meanMaxT> 26.79, "Yes", "No") #this is cycle 5, 11, 12, 22, 23

#years with long hot periods
sumclimaY$extLongHot<-ifelse(sumclimaY$day100> 64, "Yes", "No") #this is cycle 5, 11, 22, 23

#cold years
sumclimaY$extlowTemp<-ifelse(sumclimaY$meanMinT< 13.3, "Yes", "No") #this is cycle 2, 8, 10

#years with long droughts
sumclimaY$extDry<-ifelse(sumclimaY$Dext> 750, "Yes", "No") #this is cycle 11,12,13,15,23

#years with high annual precipitation
sumclimaY$Wet<-ifelse(sumclimaY$meanPrec> 4, "Yes", "No") #this is cycle 2,4,15,18

#years with low annual precipitation
sumclimaY$LowPrec<-ifelse(sumclimaY$meanPrec< 2.5, "Yes", "No") #this is cycle 5,10,11

#years with bursty rainfall
sumclimaY$Bursty<-ifelse(sumclimaY$burstinessRain> 0.09, "Yes", "No") #this is cycle 1 15 22 23

#join data
sumclimaYSub<-sumclimaY%>%select(cycle, exthighTemp,extLongHot, extDry,Wet,LowPrec,Bursty)
netclimR<-left_join(netclim,sumclimaYSub)


ggplot(netclimR,aes(x=pond,y=betweenness,color=exthighTemp))+
  geom_point(position=position_dodge(width=0.5))+stat_summary(fun.data=mean_se, position=position_dodge(width=0.5))+theme_bw()

ggplot(netclimR,aes(x=pond,y=Connectance,color=Bursty))+
  geom_point(position=position_dodge(width=0.5))+stat_summary(fun.data=mean_se, position=position_dodge(width=0.5))+theme_bw()


```

#Network vs Weather PCA analysis
Instead of using all the different weather variables, we can also use the PCA of climate. Only two were important, which makes this a little bit easier
```{r}
climatePCA<-read_csv("output/frog_WeatherPCA.csv")
netclimPC<-left_join(netclim,climatePCA)
netclimPC_g1<-left_join(netclim_g1,climatePCA)
netclimPC_g2<-left_join(netclim_g2,climatePCA)
netclimPC_g3<-left_join(netclim_g3,climatePCA)

currentPCdata=netclimPC_g3

mSPC <- lmer(No_Spec~PC1+PC2+(1|pond)+cycle, data=currentPCdata)
mbPC <- lmer(betweenness~PC1+PC2+(1|pond)+cycle, data=currentPCdata)
mcPC <- lmer(Connectance~PC1+PC2+(1|pond)+cycle, data=currentPCdata)
mEPC <- lmer(log(totalEdges)~PC1+PC2+(1|pond)+cycle, data=currentPCdata)
mmPC <- lmer(modularity~PC1+PC2+(1|pond)+cycle, data=currentPCdata)
mDPC <- lmer((mean_Distance)~PC1+PC2+(1|pond)+cycle, data=currentPCdata)
mcenPC <- lmer(harm.centrality~PC1+PC2+(1|pond)+cycle, data=currentPCdata)
mBurstPC <- lmer(B~PC1+PC2+(1|pond)+cycle, data=currentPCdata)
mMemPC <- lmer(time_memory~PC1+PC2+(1|pond)+cycle, data=currentPCdata)



# Create a list of models with descriptive names
models <- list(
  "No_Species" = mSPC,
  "Betweenness" = mbPC,
  "Connectance" = mcPC,
  "Total_Edges" = mEPC,
  "Modularity" = mmPC,
  "Mean_Distance" = mDPC,
  "Harmonic_Centrality" = mcenPC,
  "Burstiness" = mBurstPC,
  "Memory"= mMemPC
)

# Function to extract ANOVA results and coefficient information
extract_anova_results <- function(model, model_name) {
  anova_result <- Anova(model)
  coef_summary <- summary(model)$coefficients
  
  # Get fixed effects names (excluding random effects)
  fixed_effects <- rownames(coef_summary)
  anova_effects <- rownames(anova_result)
  
  # Initialize vectors for effect sizes and standard errors
  effect_sizes <- rep(NA, length(anova_effects))
  std_errors <- rep(NA, length(anova_effects))
  
  # For each ANOVA effect, find corresponding coefficient(s)
  for(i in 1:length(anova_effects)) {
    effect_name <- anova_effects[i]
    
    # Find matching coefficients
    matching_coefs <- grep(effect_name, fixed_effects, value = TRUE)
    
    # If no direct match, try alternative matching patterns
    if(length(matching_coefs) == 0) {
      if(grepl(":", effect_name)) {
        # Handle interactions
        terms <- strsplit(effect_name, ":")[[1]]
        matching_coefs <- fixed_effects[sapply(fixed_effects, function(x) {
          all(sapply(terms, function(term) grepl(term, x)))
        })]
      } else {
        # Single term
        matching_coefs <- fixed_effects[grepl(effect_name, fixed_effects)]
      }
    }
    
    # Take the first matching coefficient if multiple found
    if(length(matching_coefs) > 0) {
      primary_coef <- matching_coefs[1]
      coef_index <- which(fixed_effects == primary_coef)
      if(length(coef_index) > 0) {
        effect_sizes[i] <- round(coef_summary[coef_index, "Estimate"], 4)
        std_errors[i] <- round(coef_summary[coef_index, "Std. Error"], 4)
      }
    }
  }
  
  # Create results data frame
  results_df <- data.frame(
    Model = model_name,
    Effect = anova_effects,
    Effect_Size = effect_sizes,
    Std_Error = std_errors,
    Chisq = round(anova_result$Chisq, 3),
    Df = anova_result$Df,
    P_value = round(anova_result$`Pr(>Chisq)`, 4),
    Significance = ifelse(anova_result$`Pr(>Chisq)` < 0.001, "***",
                         ifelse(anova_result$`Pr(>Chisq)` < 0.01, "**",
                               ifelse(anova_result$`Pr(>Chisq)` < 0.05, "*",
                                     ifelse(anova_result$`Pr(>Chisq)` < 0.1, ".", ""))))
  )
  
  return(results_df)
}

# Apply the function to all models
all_results <- do.call(rbind, mapply(extract_anova_results, models, names(models), SIMPLIFY = FALSE))

# Reset row names
rownames(all_results) <- NULL

# Display the complete results table
print("Complete ANOVA Results for All Models:")
print(all_results)

# Save to CSV
#write.csv(all_results, "combined_anova_results.csv", row.names = FALSE)

# Create a wide format table showing p-values for each effect across models
pivot_table <- all_results %>%
  select(Model, Effect, P_value) %>%
  pivot_wider(names_from = Model, values_from = P_value) %>%
  arrange(Effect)

print("P-values across all models:")
print(pivot_table)

# Create effect size comparison table
effect_size_table <- all_results %>%
  select(Model, Effect, Effect_Size) %>%
  pivot_wider(names_from = Model, values_from = Effect_Size) %>%
  arrange(Effect)

print("Effect sizes across all models:")
print(effect_size_table)

# Create standard error comparison table
std_error_table <- all_results %>%
  select(Model, Effect, Std_Error) %>%
  pivot_wider(names_from = Model, values_from = Std_Error) %>%
  arrange(Effect)

print("Standard errors across all models:")
print(std_error_table)

# Create significance pattern table with proper sorting
sig_table <- all_results %>%
  select(Model, Effect, Significance) %>%
  # Add interaction level column (count colons in Effect names)
  mutate(interaction_level = str_count(Effect, ":")) %>%
  # Sort by interaction level first, then alphabetically within each level
  arrange(interaction_level, Effect) %>%
  # Remove the helper column and create wide format
  select(-interaction_level) %>%
  pivot_wider(names_from = Model, values_from = Significance, values_fill = "")

print(sig_table)
```

