---
title: "Climate data preparation"
author: "Volker Rudolf"
date: "2025-06-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Goals
The code here prepares the climate data from our frog pond location for general analysis, and looks at some climate patterns 

# Load packages
```{r, warning=FALSE, message=FALSE}
 #Load packages
  #list of packages that are needed
  pack<-c("tidyverse",
          "knitr",
          "here",
          "lubridate",
          "paletteer",
          "ggpubr",
          "patchwork",
          "factoextra"
          )
  #check whether packages are installed and load them, if not get and then load
 package.check <- lapply(pack, FUN = function(x){
    if (!require(x, character.only = TRUE)) 
      {
        install.packages(x, dependencies = TRUE)
        library(x, character.only = TRUE)
      }
    })
   #search() #verify that packages are loaded

```

# Temperature and precipitation data
```{r formatdata}
#get and organize raw daily climate observations from weather station
climate<-read_csv(here("data","Climatedata2011_2024.csv"))%>%select(DATE,PRCP,TAVG,TMAX,TMIN)
climate$DATE<-as.Date(climate$DATE,format="%m/%d/%Y")
climate<-climate%>%mutate(year = year(DATE), month=month(DATE), day=day(DATE),doy=as.numeric(format(DATE, "%j")))%>%subset(year<2025)

#now this is usually on annual basis, but we need to align it with the "frog cycle" we used for the community analysis
climate<-climate%>%subset(year<2025)%>%mutate(cycle=case_when(
  doy>340~year-1999,
  .default=year-2000
))
climate<-climate%>%mutate(cycle_day=case_when(
  doy>=339~doy-338,
  .default=doy+27
))

sumclimaY<-climate%>%group_by(cycle)%>%
  summarize(meanMaxT=mean(TMAX,na.rm = TRUE), 
            meanMinT=mean(TMIN,na.rm = TRUE),
            hotdays=sum(TMAX > 35,na.rm = TRUE),
            meanPrec=mean(PRCP,na.rm=TRUE),
            noPrec=sum(PRCP==0,na.rm=TRUE))
```

# Extreme climate year
We will identify extreme climate events based on the deviation from the average and plot it for reference. We'll focus on extreme hot and dry for now. 

```{r ExtremeYears}
sumclimaYE<-subset(sumclimaY, cycle<25)%>%
  mutate(
    mMaxTY=mean(meanMaxT),
    mMinTY=mean(meanMinT),
    mPrecY=mean(meanPrec),
    mhotdaysY=mean(hotdays),
    mNoPrec=mean(noPrec),
    eMaxTY=mean(meanMaxT)+sd(meanMaxT),
    eMinTY=mean(meanMinT)-sd(meanMinT),
    ePrecY=mean(meanPrec)-sd(meanPrec),
    eNoPrec=mean(mNoPrec)+sd(noPrec),
    ehotdaysY=mean(hotdays)+sd(hotdays),
    )%>%
  mutate(exthighTemp=ifelse(meanMaxT>eMaxTY, "Yes", "No"),
         extlowTemp=ifelse(meanMinT<eMinTY, "Yes", "No"),
         extlongHeat=ifelse(hotdays>ehotdaysY, "Yes", "No"),
         extDry=ifelse(meanPrec<ePrecY, "Yes", "No"),
         extDryDays=ifelse(noPrec>eNoPrec, "Yes", "No"),
         extHot=ifelse(exthighTemp=="Yes" | extlongHeat=="Yes", "Yes", "No"),
         extDryT=ifelse(extDry=="Yes" | extDryDays=="Yes", "Yes", "No"))

#save extreme categories
write.csv(sumclimaYE, here("output","sumclimaYE.csv"), row.names = FALSE)

#make plots
ETmaxplot<-ggplot(subset(sumclimaYE),aes(x=cycle,y=meanMaxT))+
  geom_hline(aes(yintercept=mMaxTY),linetype="dashed")+
  geom_line()+
  geom_point(aes(color=exthighTemp), size=3)+
  geom_ribbon(aes(ymin = mMaxTY-sd(meanMaxT), ymax = eMaxTY), alpha = 0.3, fill = "grey") +
  labs(y="Daily high temperature (\u00b0C)", x="Annual Cycle")+
  scale_color_manual(name = "Extreme event",
    values = c("No" = "black", "Yes" = "red"))+
  scale_x_continuous(
    limits = c(1, NA), 
    breaks = seq(from = 1, to = max(sumclimaYE$cycle), by = 1))+
  theme_classic()+
  theme(legend.position = c(0.15, 0.9),
        axis.text.x = element_text(size = 10), 
        axis.text.y = element_text(size = 10), 
        axis.title = element_text(size = 14),)+theme(legend.position = "none")

Ehotdayplot<-ggplot(subset(sumclimaYE),aes(x=cycle,y=hotdays))+
  geom_hline(aes(yintercept=mhotdaysY),linetype="dashed")+
  geom_line()+
  geom_point(aes(color=extlongHeat), size=3)+
  geom_ribbon(aes(ymin = mhotdaysY-sd(hotdays), ymax = ehotdaysY), alpha = 0.3, fill = "grey") +
  labs(y="Total hot days (>35\u00b0C)", x="Annual Cycle")+
  scale_color_manual(name = "Extreme event",
    values = c("No" = "black", "Yes" = "red"))+
  scale_x_continuous(limits = c(1, NA), breaks = seq(from = 1, to = max(sumclimaYE$cycle), by = 1))+
  theme_classic()+
  theme(legend.position = "none")+
  theme(legend.position = c(0.15, 0.9),
        axis.text.x = element_text(size = 10), 
        axis.text.y = element_text(size = 10), 
        axis.title = element_text(size = 14))
  
Eprecplot<-ggplot(subset(sumclimaYE),aes(x=cycle,y=meanPrec))+
  geom_hline(aes(yintercept=mPrecY),linetype="dashed")+
  geom_line()+
  geom_point(aes(color=extDry), size=3)+
  geom_ribbon(aes(ymax = mPrecY+sd(meanPrec), ymin = ePrecY), alpha = 0.3, fill = "grey") +
  labs(y="Precipitation (mm)", x="Annual Cycle")+
  scale_color_manual(name = "Extreme event",
    values = c("No" = "black", "Yes" = "red"))+
  scale_x_continuous(limits = c(1, NA), breaks = seq(from = 1, to = max(sumclimaYE$cycle), by = 1))+
  theme_classic()+
  theme(legend.position = "none")+
  theme(legend.position = c(0.15, 0.9),
        axis.text.x = element_text(size = 10), 
        axis.text.y = element_text(size = 10), 
        axis.title = element_text(size = 14))
  
Enorainplot<-ggplot(subset(sumclimaYE),aes(x=cycle,y=noPrec))+
  geom_hline(aes(yintercept=mNoPrec),linetype="dashed")+
  geom_line()+
  geom_point(aes(color=extDryDays), size=3)+
  geom_ribbon(aes(ymin = mNoPrec-sd(noPrec), ymax = mNoPrec+sd(noPrec)), alpha = 0.3, fill = "grey")+
  labs(y="Total days without rain", x="Annual Cycle")+
  scale_color_manual(name = "Extreme event",
    values = c("No" = "black", "Yes" = "red"))+
  scale_x_continuous(limits = c(1, NA), breaks = seq(from = 1, to = max(sumclimaYE$cycle), by = 1))+
  theme_classic()+
  theme(legend.position = "none")+
  theme(legend.position = c(0.15, 0.9),
        axis.text.x = element_text(size = 10), 
        axis.text.y = element_text(size = 10), 
        axis.title = element_text(size = 14))
  
combined_plot <- (ETmaxplot + Ehotdayplot) / (Eprecplot + Enorainplot)+
  plot_layout(guides="collect")& 
  theme(
    legend.position = "none",
    #axis.text = element_text(size = 8), 
    #axis.title = element_text(size = 9, face="bold")
  )+theme(legend.position="top")

combined_plot
ggsave(here("Figures","extremeYears.png"),plot=combined_plot,dpi=600,width = 30, height = 15, units = "cm")

```

#Burstiness of rainfall
Lets calculate how bursty rain fall is.
```{r burstiness}
# Define a function to calculate burstiness
calculate_burstiness <- function(dates) {
  if (length(dates) < 2) return(NA)  # Not enough rainy days
  inter_event_times <- as.numeric(diff(sort(dates)))
  mu <- mean(inter_event_times)
  sigma <- sd(inter_event_times)
  B<-if ((mu + sigma) == 0) NA else (sigma - mu) / (sigma + mu)
  data.frame(
    burstinessRain=B,
    Rain_inter_event_time=mu
  )
}

# Filter to only rainy days
rainy_days <- climate %>% filter(PRCP > 0)

# Group by year and compute burstiness
burstiness_by_cycle <- rainy_days %>%
  group_by(cycle) %>%
  summarise(
    stats = list(calculate_burstiness(DATE)))%>%
  unnest(cols=c(stats))

#add it to yearly climate data
sumclimaYB<-left_join(sumclimaY,burstiness_by_cycle)
```

##drought index data
This is data on the drought index, which is calculated based on "missing" rain and is a composite integrative index. It comes in range of severeness categories, so not as continuous values. To use it in the analysis we modify it and combine it into two single indeces, one indicating the mean drought severity, and one indicating most extreme drought status
```{r drought}
#load data and organize data & convert it to "frog" annual cycles
drought<-read_csv(here("data","DroughtDataDCForest.csv"))%>%select(ValidStart,None,D0,D1,D2,D3,D4)
drought$DATE<-as.Date(drought$ValidStart,"%m/%d/%Y")
drought<-drought%>%mutate(year = year(DATE), month=month(DATE), day=day(DATE),doy=as.numeric(format(DATE, "%j")))%>%subset(year<2025)
drought<-drought%>%mutate(cycle=case_when(
  doy>340~year-1999,
  .default=year-2000
))
drought<-drought%>%mutate(cycle_day=case_when(
  doy>=339~doy-338,
  .default=doy+27
))

#combine the severity indexes into one drought severity metric. We do this by giving more weight to more sever drought indexes
drought<-drought%>%mutate(Dseverity=D0+D1*2+D2*3+D3*4+D4*4)
sumdrought<-drought%>%group_by(cycle)%>%
  summarize(Dsevmean=mean(Dseverity),Dext=max(Dseverity))
ggplot(sumdrought, aes(x=cycle, y=Dext))+geom_line()+geom_point()+theme_bw()
ggplot(sumdrought, aes(x=cycle, y=Dsevmean))+geom_line()+geom_point()+theme_bw()


#add it to climate data
sumclimaYBD<-left_join(sumclimaYB,sumdrought)%>%subset(cycle<25)

```

Now that we have all weather data, we save it and we scale the variables so we can compare them
```{r scaleweatherdata}
#save raw data with extreme years
sumclimaYE<-left_join(sumclimaYBD, sumclimaYE)
write_csv(sumclimaYE, here("output","frog_climate_data_annualsum.csv"))

#normalize the climate data since they are all on different scales.
scaledsumclimaY <- data.frame(
  cycle=sumclimaYBD[1],                      # keep the first column untouched
  scale(sumclimaYBD[, -1])              # scale all other columns
)

#save scaled data
write_csv(scaledsumclimaY, here("output","frog_climate_data_annual_scaled.csv"))
```

# PCA of weather data
Now since there are several climate variables and some are correlated, we will simplify the complex data set using PCA analysis
```{r PCA.weather}
#get normalzed weather data
scaledsumclimaY<-read.csv(here("output","frog_climate_data_annual_scaled.csv"))

#rename variables for plotting
new_var_names <- c("Max.Temp", "Min.Temp", "Hot.Days", "Mean.Prec", "Zero.Day.Prec","Burst.Rain","Days.btw.Rain", "Mean.Drougth", "Max.Drought")
colnames(scaledsumclimaY)[-1] <- new_var_names

#run pca
weather.pca <- prcomp(scaledsumclimaY[,-1], scale = TRUE)
print(weather.pca)

#plot PC %explanation
fviz_eig(weather.pca, col.var="blue")

#plot vector loading for first 2 PCA
pca_loading<-fviz_pca_var(weather.pca,
             labelsize = 4,
             repel = TRUE
             )+ 
  labs(title = "",              # Removes the plot title
   x = "PC1 (55.7%)",               # Relabels X-axis
   y = "PC2 (22%)"                # Relabels Y-axis
    ) +
  theme_pubr(base_size = 18)
pca_loading
ggsave(here("Figures","WeatherPCAloading.png"),plot=pca_loading,dpi=600,width = 15, height = 15, units = "cm")

  
#plot cycles in PCA space
cycle_values <- scaledsumclimaY[, 1]
cycl <- get_pca_ind(weather.pca)
cycl

#plot both loading and year values
weather_pca_biPlot<-fviz_pca_biplot(weather.pca,
    habillage = scaledsumclimaY[, 1],
    pointshape=16,
    geom.var="none",
    col.var = "black", 
    # alpha.var = 0.8,
    repel = TRUE, 
    label = "all",
    labelsize = 4
    ) + 
  labs(title = "",              # Removes the plot title
   x = "PC1",               # Relabels X-axis
   y = "PC2"                # Relabels Y-axis
    ) +
  scale_colour_viridis_d(option="I")+
  theme_pubr(base_size = 18)+
  theme(legend.position = "none")

weather_pca_biPlot

ggsave(here("Figures","WeatherPCABiplot.png"),plot=weather_pca_biPlot,dpi=600,width = 15, height = 15, units = "cm")


#since PC1 and PC2 explain most, let's save both in new file for later analysis
#Extract PC1 and PC2 scores
pc_scores <- as.data.frame(weather.pca$x[, 1:2])
colnames(pc_scores) <- c("PC1", "PC2")            # Rename for clarity

# Add cycle information back
pc_scores$cycle <- scaledsumclimaY$cycle

#Reorder columns: cycle, PC1, PC2
pc_output <- pc_scores %>%select(cycle, PC1, PC2)

# Save to CSV
write.csv(pc_output, here("output","frog_WeatherPCA.csv"), row.names = FALSE)
```