---
title: "Climate & community"
author: "Volker Rudolf"
date: "2025-06-12"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Goals
We will examine how climate variables are linked to different aspects of community state transitions. THis is focused on frog calling data from 8 different ponds across 22 years. Climate data is taking from local weather stations, and frog calling data was converted to transition matrix indicating how community states (i.e. a specific combination of species) transition over time within a year.

## Load packages & data
```{r, warning=FALSE, message=FALSE}
 #Load packages
  #list of packages that are needed
  pack<-c("tidyverse",
          "knitr",
          "here",
          "car",
          "lme4",
          "flextable",
          "ggpubr",
          "broom.mixed",
          "lmerTest",
          "ggrepel", 
          "cowplot",
          "factoextra",
          "FactoMineR"
          )
  #check whether packages are installed and load them, if not get and then load
 package.check <- lapply(pack, FUN = function(x){
    if (!require(x, character.only = TRUE)) 
      {
        install.packages(x, dependencies = TRUE)
        library(x, character.only = TRUE)
      }
    })
   #search() #verify that packages are loaded

```

Let's load the community matrix and climate data (see "Community transition code.Rdm" and "Climate data preparation.Rdm")
```{r get data, message=F}
#get summary metrics of the transition matrix
state_connectance_metrics<-read_rds(here("./output/frog_transition_matrix_metrics.rds"))#raw data transition matrix 
#we have 3 additional version of the transition matrix that ignore a gap of 1, 2, or 3 day in the calling of a species and treat it as a continuous period. 
state_connectance_metrics_g1<-read_rds(here("output","frog_transition_matrix_metrics_g1.rds")) 
state_connectance_metrics_g2<-read_rds(here("output","frog_transition_matrix_metrics_g2.rds"))
state_connectance_metrics_g3<-read_rds(here("output","frog_transition_matrix_metrics_g3.rds"))

#get annual climate data summary metrics 
sumclima<-read_csv(here("output","frog_climate_data_annual_scaled.csv"))%>%subset(cycle<25)

#combine climate and network statistics
netclim <- reduce(list(sumclima, state_connectance_metrics), full_join) %>%
  select(pond, cycle, Pond_Cycle, everything())
netclim_g1 <- reduce(list(sumclima, state_connectance_metrics_g1), full_join) %>%
  select(pond, cycle, Pond_Cycle, everything())
netclim_g2 <- reduce(list(sumclima, state_connectance_metrics_g2), full_join) %>%
  select(pond, cycle, Pond_Cycle, everything())
netclim_g3 <- reduce(list(sumclima, state_connectance_metrics_g3), full_join) %>%
  select(pond, cycle, Pond_Cycle, everything())

```
##metric PCA (For general info, not included in manuscript) 
let's see how those metrics fall out in a multidimensional space in a PCA to better understand they relationship (note, correlations are in Community_transition code)
```{r networkPCA, message=F}
library("factoextra")
library("FactoMineR")
metricpca<- metrics%>%select(-meanDayOnset)%>%prcomp(scale = TRUE)
print(metricpca)

metricpca_g1<-metrics_g1 %>%select(-meanDayOnset)%>% prcomp(scale = TRUE)
metricpca_g3<- metrics_g3%>% select(-meanDayOnset)%>%prcomp( scale = TRUE)

#get corresponding ponds and year
pca_data <- as.data.frame(metricpca$x)
pca_data$cycle<-state_connectance_metrics$cycle
pca_data$pond<-state_connectance_metrics$pond

#plot PC %explanation which is almost identical for all three data sets.
#fviz_eig(metricpca, col.var="blue")
#fviz_eig(metricpca_g1, col.var="blue")
#fviz_eig(metricpca_g3, col.var="blue")

#plot variables for first 2 PCA
fviz_pca_var(metricpca,
             repel = TRUE
             )+ggtitle("PCA loading not ignoring gaps")

#plot both loading and year values
fviz_pca_biplot(metricpca,
                label="var",
                geom_ind="point",
,                col.var = "black",               
                gradient.cols = c("blue", "green"),
                repel = TRUE,
) +ggtitle("PCA not ignoring gaps")
fviz_pca_biplot(metricpca_g1,
                label="var",
                geom_ind="point",
,                col.var = "black",               
                gradient.cols = c("blue", "green"),
                repel = TRUE,
)+ggtitle("PCA ignoring 1 day gaps")
fviz_pca_biplot(metricpca_g3,
                label="var",
                geom_ind="point",
,                col.var = "black",               
                gradient.cols = c("blue", "green"),
                repel = TRUE,
)+ggtitle("PCA ignoring 3 day gaps")



metricPCA<-cbind(netclim[,1:2],metricpca$x[,1:2])
metricPCA_g1<-cbind(netclim[,1:2],metricpca_g1$x[,1:2])
metricPCA_g3<-cbind(netclim[,1:2],metricpca_g3$x[,1:2])

netPCAclim<-full_join(sumclima,metricPCA)%>%select(pond, cycle, everything())
netPCAclim_g1<-full_join(sumclima, metricPCA_g1)%>%select(pond, cycle, everything())
netPCAclim_g3<-full_join(sumclima, metricPCA_g3)%>%select(pond, cycle, everything())

```
It's also interesting that the correlation between variables are largely robust and don't change a lot if we ignore small 1, - 3 day gaps in calling. Based on these relationship, we definitely can drop unique nodes since it's almost perfectly correlated with uniqueEdges, and uniqueEdges seems to be more different from No_species than unique nodes and bit more informative on other aspects of the network. We also don't need  mean_t and st_t which are part of memory and burstiness. Connectance and Weighted_connectance are also highly correlated.


#Network structure & Weather 
Instead of using all the different weather variables, we will  use the PCA of climate to avoid co-linearity issues among weather variables and reduce the number of predictors. Fortunately only two PCA were really important, which makes this a little bit easier

##PCA analysis
```{r metric_WeatherPCA}
climatePCA<-read_csv(here("./output/frog_WeatherPCA.csv"))
netclimPC<-left_join(netclim,climatePCA)
netclimPC_g1<-left_join(netclim_g1,climatePCA)
netclimPC_g2<-left_join(netclim_g2,climatePCA)
netclimPC_g3<-left_join(netclim_g3,climatePCA)

currentPCdata<-netclimPC

variable_labels <- c(
  "Unique_Nodes" = "Unique nodes",
  "Connectance" = "Connectance",
  "modularity" = "Modularity",
  "betweenness" = "Betweenness",
  "mean_Distance" = "Mean distance",
  "meanDurAv" = "Node duration",
  "freqStateAv" = "Node frequency",
  "skewDur" = "Duration skewness",
  "TotalEdges" = "Total edges",
  "ave_Edge_weight"="Edge frequency",
  "skewEdge" = "Skewness edge freq",
  "time_memory" = "Memory",
  "B" = "Burstiness",
  "event.rate" = "Event rate"
)

#get variables to analyze
variable.for.extanalysis<-c("Unique_Nodes","Connectance", "modularity","betweenness", "mean_Distance", "freqStateAv",    "meanDurAv","skewDur","TotalEdges","ave_Edge_weight","skewEdge", "time_memory","B","event.rate")

#scale data to be comparable but first log-transform two variables to improve lmer model fit
currentPCdataS<-currentPCdata%>% 
  mutate(across(c("Connectance", "mean_Distance"), log))

currentPCdataS[variable.for.extanalysis] <- lapply(currentPCdata[variable.for.extanalysis], scale)
#analyze the data

resPCA <- map_dfr(variable.for.extanalysis, function(response) {
  # Fit the mixed model for this response
  model <- lmerTest::lmer(as.formula(paste(response,"~ PC1+PC2 +(1 | pond)")), data = currentPCdataS)
  

  # Tidy model output: fixed effects only, with confidence intervals
  broom.mixed::tidy(model, effects = "fixed", conf.int = TRUE) %>%
    mutate(response = response)%>%
    filter(!str_detect(term,"(Intercept)"))
  }
  )

# Create significance pattern table
sig_tablePCA <- resPCA %>%
  mutate(response=variable_labels[response])%>%
  select(response, term,p.value, statistic) %>%
  mutate(
    p.value = round(p.value, 4),
    statistic=round(statistic,2)
    ) %>%
  pivot_wider(
    names_from = term, 
    values_from = c(statistic,p.value),
    names_glue="{term}_{.value}",
    names_vary="slowest"
  )



# Print formatted flextable for Word output
sig_table_PCA<-sig_tablePCA %>%
  flextable() %>%
  set_caption("Table: Significance Patterns") %>%
  theme_booktabs() %>%        # Adds horizontal lines under header and at bottom
  bold(part = "header") %>%   # Make header bold
  align(align = "center", part = "all") %>%  # Center all text
  autofit()  

#save table
save_as_docx(sig_table_PCA, path = here("Output", "sig_table_PCA.docx"))


```

##PCA analysis effect size plot
```{r}
resPCA_prep <- resPCA %>%
  mutate(significant = p.value < 0.05)

# Add a category column to your data based on your grouping
resPCA_prep <- resPCA_prep %>%
  mutate(category = case_when(
    response %in% c("Unique_Nodes",  "Connectance", 
                    "betweenness", "mean_Distance", "modularity") ~ "Network",
    response %in% c("meanDurAv", "freqStateAv", "skewDur") ~ "Node",
    response %in% c("ave_Edge_weight","TotalEdges", "skewEdge") ~ "Edge",
    response %in% c("event.rate", "B", "time_memory") ~ "Spacing",
    TRUE ~ "Other"
  )) %>%
  # Set category order (top to bottom as they appear in your list)
  mutate(category = factor(category, 
                          levels = c("Network", "Node", "Edge", "Spacing"))) %>%
  # Set response variable order within categories
  mutate(response = factor(response, levels = my_grouped_y_order))


# Create background shading by category (spans entire category height)
background_data <- data.frame(
  category = factor(c("Network", "Node", "Edge", "Spacing"),
                   levels = c("Network", "Node", "Edge", "Spacing")),
  fill_color = c("white", "grey95", "white", "grey95")
) %>%
  # Expand to include both terms
  tidyr::crossing(term = unique(resPCA_prep$term))

# Create the plot with faceting by category
coef.plotPCA <- resPCA_prep %>%
  ggplot(aes(x = estimate, y = response, color = term)) +
  # Add background rectangles that span entire category
  geom_rect(data = background_data,
            aes(fill = fill_color),
            xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf,
            inherit.aes = FALSE, alpha = 1) + 
  scale_fill_identity() +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray50", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high, alpha = significant), 
                 height = 0.0, linewidth = 1) +
  geom_point(aes(alpha = significant), size = 3) +
  scale_alpha_manual(values = c("FALSE" = 0.3, "TRUE" = 1), guide = "none") +
  # Facet by both category (rows) and term (columns)
  facet_grid(category ~ term, 
             scales = "free_y",      # Free y-axis for each category
             space = "free_y",       # Adjust panel height based on number of items
             switch = "y",           # Put category labels on the left
             labeller = labeller(
               term = c("PC1" = "Weather PCA1", "PC2" = "Weather PCA2"),
               category = label_value)) +
  scale_y_discrete(
    drop = TRUE,  # Drop unused factor levels in each facet
    labels = variable_labels) +
  labs(x = "Standardized coefficients and 95% CI", y = "") + 
  scale_color_manual(
    values = c("PC1" = "#33B8CC", "PC2" = "#CCAA33"), 
    guide = "none") +
  theme_pubr(base_size = 16) +
  theme(
    strip.background = element_blank(),  # Remove box around facet labels
    strip.text.x = element_text(size = 18, face = "bold"),
    strip.text.y.left = element_text(size = 16, face = "bold", angle = 90),  # Rotate category labels
    strip.placement = "outside",
    panel.spacing = unit(0.3, "lines"),      # Space between categories
    panel.border = element_blank(),  # Remove boxes around panels
    axis.text.x = element_text(size = 14), 
    axis.text.y = element_text(size = 16), 
    axis.title = element_text(size = 18)
  )

coef.plotPCA
ggsave(here("Figures","PCAWeather.effect.png"),plot=coef.plotPCA,dpi=600,width = 30, height = 15, units = "cm")
```

# Extreme climate
Let's look at differences in community structure metrics between extreme years and normal years.
```{r extremeClimateAnalysis}
#get raw climate data
sumclima.raw<-read.csv(here("./output/frog_climate_data_annualsum.csv"))%>%
  mutate(exthighTemp=factor(exthighTemp), extlongHeat=factor(extlongHeat),extDry=factor(extDry),extDryT=factor(extDryT), extHot=factor(extHot),extYear=if_else(extDryT=="Yes" & extHot=="Yes","Yes","No"))%>%
  select(-burstinessRain)

netclimCur<-netclim #can be netclim, netclime_g1, _g2, _g3 based on gap data set
#join data
netclimR<-left_join(netclimCur,sumclima.raw, by=c("cycle"))


netclimR[variable.for.extanalysis] <- lapply(netclimR[variable.for.extanalysis], scale)

#analyze the data

res <- map_dfr(variable.for.extanalysis, function(response) {
  # Fit the mixed model for this response
  model <- lmer(as.formula(paste(response,"~ extHot*extDryT + (1 | pond)")), data = netclimR)
  
  # Tidy model output: fixed effects only, with confidence intervals
  broom.mixed::tidy(model, effects = "fixed", conf.int = TRUE) %>%
    mutate(response = response)%>%filter(!str_detect(term,"(Intercept)"))
  }
  )

# Create significance pattern table
sig_tableExt <- res %>%
  mutate(response=variable_labels[response])%>%
  select(response, term,p.value, statistic) %>%
  mutate(
    p.value = round(p.value, 4),
    statistic=round(statistic,2)
    ) %>%
  pivot_wider(
    names_from = term, 
    values_from = c(statistic,p.value),
    names_glue="{term}_{.value}",
    names_vary="slowest"
  )

# Print formatted flextable for Word output
sig_table_Ex<-sig_tableExt %>%
  flextable() %>%
  set_caption("Table: Significance Patterns") %>%
  theme_booktabs() %>%        # Adds horizontal lines under header and at bottom
  bold(part = "header") %>%   # Make header bold
  align(align = "center", part = "all") %>%  # Center all text
  autofit()  

#save table
save_as_docx(sig_table_Ex, path = here("Output", "sig_table_Ext.docx"))

```

test if species diversity differs between extreme years
```{r}
sp_Ex<-lmer(No_Spec~ extHot+extDryT + (1 | pond), data = netclimR)
plot(sp_Ex)
Anova(sp_Ex)

```


Plot effect sizes of extreme drought and temperature
```{r}
# Add a category column to your data based on your grouping
res_prepE <- res %>%
  filter(!response %in% c("No_Spec")) %>%  # Explicitly drop species richness
   mutate(significant = p.value < 0.05)%>%
   mutate(category = case_when(
    response %in% c("Unique_Nodes", "Connectance", 
                    "betweenness", "mean_Distance", "modularity") ~ "Network",
    response %in% c("meanDurAv", "freqStateAv", "skewDur") ~ "Node",
    response %in% c("ave_Edge_weight",  "TotalEdges","skewEdge") ~ "Edge",
    response %in% c("event.rate", "B", "time_memory") ~ "Spacing",
    TRUE ~ "Other"
  )) %>%
  # Set category order (top to bottom as they appear in your list)
  mutate(category = factor(category, 
                          levels = c("Network", "Node", "Edge", "Spacing"))) %>%
  # Set response variable order within categories
  mutate(response = factor(response, levels = my_grouped_y_order))


# Create background shading by category (spans entire category height)
background_data <- data.frame(
  category = factor(c("Network", "Node", "Edge", "Spacing"),
                   levels = c("Network", "Node", "Edge", "Spacing")),
  fill_color = c("white", "grey95", "white", "grey95")
) %>%
  # Expand to include both terms
  tidyr::crossing(term = unique(res_prepE$term))
library(ggtext)
# Create the plot with faceting by category
coef.plotE <- res_prepE %>%
  ggplot(aes(x = estimate, y = response, color = term)) +
  # Add background rectangles that span entire category
  geom_rect(data = background_data,
            aes(fill = fill_color),
            xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf,
            inherit.aes = FALSE, alpha = 1) + 
  scale_fill_identity() +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray50", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high, alpha = significant), 
                 height = 0.0, linewidth = 1) +
  geom_point(aes(alpha = significant), size = 3) +
  scale_alpha_manual(values = c("FALSE" = 0.3, "TRUE" = 1), guide = "none") +
  # Facet by both category (rows) and term (columns)
  facet_grid(category ~ term, 
             scales = "free_y", 
             space = "free_y", 
             switch = "y", 
             labeller=labeller(term = c(
               "extDryTYes" = "Extreme Dry",
               "extHotYes" = "Extreme Hot"),
               category = label_value)) +
   scale_y_discrete(
    drop = TRUE,  # Drop unused factor levels in each facet
    labels = variable_labels) +
  labs(x = "Standardized coefficients and 95% CI", y = "") + 
  scale_color_manual(
    values = c("extDryTYes" = "#A65F32", 
               "extHotYes" = "#CC3333"), 
    guide = "none") +
  theme_pubr(base_size = 16) +
 theme(
    strip.background = element_blank(),
    # Modify strip.text.x to underline column (term) labels
    strip.text.x = element_text(size = 18, face = "bold"),
    # Modify strip.text.y.left to underline row (category) labels
    strip.text.y.left = element_text(size = 16, face = "bold", angle = 90), 
    strip.placement = "outside",
    panel.spacing = unit(0.3, "lines"),
    panel.border = element_blank(),
    axis.text.x = element_text(size = 14), 
    axis.text.y = element_text(size = 16), 
    axis.title = element_text(size = 18)
  )

coef.plotE
ggsave(here("Figures","ExtremeClim.effect.png"),plot=coef.plotE,dpi=600,width = 30, height = 15, units = "cm")

```
combine plots into one figure
```{r}
finalEffP<-plot_grid(coef.plotPCA,coef.plotE,
                     labels=c("A","B"),
                     ncol=1,
                     rel_heights = c(1, 1)
                     )
ggsave(here("Figures","ClimateExt_Effect.png"),plot=finalEffP,dpi=600,width = 30, height = 30, units = "cm")
```


lets repeat the analysis with No_species as covariate
```{r}
resPCA_sp <- map_dfr(variable.for.extanalysis, function(response) {
  # Fit the mixed model for this response
  model <- lmerTest::lmer(as.formula(paste(response,"~ PC1+PC2 +No_Spec+(1 | pond)")), data = currentPCdataS)
  

  # Tidy model output: fixed effects only, with confidence intervals
  broom.mixed::tidy(model, effects = "fixed", conf.int = TRUE) %>%
    mutate(response = response)%>%
    filter(!str_detect(term,"(Intercept)"))
  }
  )

# Create significance pattern table
sig_tablePCA_sp <- resPCA_sp %>%
  mutate(response=variable_labels[response])%>%
  select(response, term,p.value, statistic) %>%
  mutate(
    p.value = round(p.value, 4),
    statistic=round(statistic,2)
    ) %>%
  pivot_wider(
    names_from = term, 
    values_from = c(statistic,p.value),
    names_glue="{term}_{.value}",
    names_vary="slowest"
  )



# Print formatted flextable for Word output
sig_table_PCA_sp<-sig_tablePCA_sp %>%
  flextable() %>%
  set_caption("Table: Significance Patterns") %>%
  theme_booktabs() %>%        # Adds horizontal lines under header and at bottom
  bold(part = "header") %>%   # Make header bold
  align(align = "center", part = "all") %>%  # Center all text
  autofit()  

#save table
save_as_docx(sig_table_PCA_sp, path = here("Output", "sig_table_PCA_sp.docx"))
```
now same for extreme
```{r}
res_sp <- map_dfr(variable.for.extanalysis, function(response) {
  # Fit the mixed model for this response
  model <- lmer(as.formula(paste(response,"~ extHot+extDryT + No_Spec+(1 | pond)")), data = netclimR)
  
  # Tidy model output: fixed effects only, with confidence intervals
  broom.mixed::tidy(model, effects = "fixed", conf.int = TRUE) %>%
    mutate(response = response)%>%filter(!str_detect(term,"(Intercept)"))
  }
  )

# Create significance pattern table
sig_tableExt_sp <- res_sp %>%
  mutate(response=variable_labels[response])%>%
  select(response, term,p.value, statistic) %>%
  mutate(
    p.value = round(p.value, 4),
    statistic=round(statistic,2)
    ) %>%
  pivot_wider(
    names_from = term, 
    values_from = c(statistic,p.value),
    names_glue="{term}_{.value}",
    names_vary="slowest"
  )

# Print formatted flextable for Word output
sig_table_Ex_sp<-sig_tableExt_sp %>%
  flextable() %>%
  set_caption("Table: Significance Patterns") %>%
  theme_booktabs() %>%        # Adds horizontal lines under header and at bottom
  bold(part = "header") %>%   # Make header bold
  align(align = "center", part = "all") %>%  # Center all text
  autofit()  

#save table
save_as_docx(sig_table_Ex_sp, path = here("Output", "sig_table_Ext_sp.docx"))
```


