---
title: "Sequential network states"
author: "Volker Rudolf"
date: "2025-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
#Goals
The goal of this analysis is to turn species phenologies into a network of community state transitions. When species co-occur, they form a community state, and each species combination is a unique state, which can occur more than once. Each state has a start and ending date, and transitions into a new state. This creates a network that we will analyze to better understand how communities change

### Load packages
```{r, warning=FALSE, message=FALSE}
 #Load packages
  #list of packages that are needed
  pack<-c("tidyverse",
          "tsna",
          "ndtv",
          "knitr",
          "here",
          "igraph",
          "ggraph",
          "beeswarm",
          "ggcorrplot",
          "broom",
          "moments",
          "car",
          "ggbeeswarm",
          "lme4"
          )
  #check whether packages are installed and load them, if not get and then load
 package.check <- lapply(pack, FUN = function(x){
    if (!require(x, character.only = TRUE)) 
      {
        install.packages(x, dependencies = TRUE)
        library(x, character.only = TRUE)
      }
    })
   #search() #verify that packages are loaded

```

## Get & transform data into a transition network

### get the data
We assume the data comes in the form of Dynamic Nodes from the "frog data preparation" code, i.e. each observation (row) indicates when (e.g. the day) a species appears (onset) and disappears (terminus). This can happen multiple times for a given species. So there are at least 3 columns: 1) species ID, 2) onset, 3) terminus
```{r getdata, message=FALSE}
#get previously curated data
#raw data
node_dynamic<-read_csv(here("./output/frogs_node_dynamic.csv"))%>%select("pond", "cycle", "onset", "terminus", "species")%>%relocate(species,.before=onset)%>%subset(cycle<25)%>%mutate(pond = as.factor(pond))

#get number of species per pond x cycle

sp_num<-node_dynamic%>%
  group_by(pond, cycle)%>%
  summarize(No_Spec=n_distinct(species),.groups="drop")
write_rds(sp_num, here("./output/frog_pond_cycle_NoSpec.rds"))  

#same data but ignoring gaps of either 1, 2, or 3 days between calling events.
node_dynamic_g1<-read_csv(here("./output/frogs_node_dynamic_gap1.csv"))%>%select("pond", "cycle", "onset", "terminus", "species")%>%relocate(species,.before=onset)%>%subset(cycle<25)%>%mutate(pond = as.factor(pond))
node_dynamic_g2<-read_csv(here("./output/frogs_node_dynamic_gap2.csv"))%>%select("pond", "cycle", "onset", "terminus", "species")%>%relocate(species,.before=onset)%>%subset(cycle<25)%>%mutate(pond = as.factor(pond))
node_dynamic_g3<-read_csv(here("./output/frogs_node_dynamic_gap3.csv"))%>%select("pond", "cycle", "onset", "terminus", "species")%>%relocate(species,.before=onset)%>%subset(cycle<25)%>%mutate(pond = as.factor(pond))

```

### organize data into community states
We will create a new data frame, that includes column for each unique species combination, and additional columns that indicate when the state starts and ends. Note, in same cases this might include state without species, which we call "none" here.
```{r community_states, message=FALSE}
generate_state_transitions <- function(df) {
    # Validate input
  if (nrow(df) == 0) {
    return(data.frame(state = character(), onset = integer(), terminus = integer()))
  }
  
  required_cols <- c("onset", "terminus", "species")
  if (!all(required_cols %in% names(df))) {
    stop(paste("Missing required columns:", paste(setdiff(required_cols, names(df)), collapse = ", ")))
  }
  
  # Get all unique transition points (both onset and terminus)
  transition_points <- sort(unique(c(min(df$onset), df$onset, df$terminus + 1)))
  
  # Initialize results list
  results <- vector("list", length(transition_points) - 1)
  
  # Process each time window between transition points
  for (i in seq_along(transition_points)[-length(transition_points)]) {
    current_day <- transition_points[i]
    next_transition <- transition_points[i + 1]
    
    # Find species present during this window
    present_species <- df$species[df$onset <= current_day & df$terminus >= current_day]
    
    # Create state string - if no species present, use "none"
    state <- if (length(present_species) == 0) "none" else paste(sort(present_species), collapse = ",")
    
    results[[i]] <- data.frame(
      state = state,
      onset = current_day,
      terminus = next_transition - 1
    )
  }
  
  # Combine results
  result_df <- do.call(rbind, results)
  
  # Add final "none" state if needed
  max_terminus <- max(df$terminus)
  if (tail(result_df$terminus, 1) < max_terminus) {
    final_none <- data.frame(
      state = "none",
      onset = tail(result_df$terminus, 1) + 1,
      terminus = max_terminus
    )
    result_df <- rbind(result_df, final_none)
  }
  
  return(result_df)
}
# Main function to process community states for any dataset
process_community_states <- function(node_dynamic_data, 
                                   grouping_vars = c("pond", "cycle"),
                                   output_file = NULL,
                                   add_duration = TRUE,
                                   add_metrics = TRUE) {
  
  # Validate inputs
  if (nrow(node_dynamic_data) == 0) {
    stop("Input data is empty")
  }
  
  required_cols <- c("onset", "terminus", "species")
  missing_cols <- setdiff(required_cols, names(node_dynamic_data))
  if (length(missing_cols) > 0) {
    stop(paste("Missing required columns:", paste(missing_cols, collapse = ", ")))
  }
  
  missing_group_cols <- setdiff(grouping_vars, names(node_dynamic_data))
  if (length(missing_group_cols) > 0) {
    stop(paste("Missing grouping columns:", paste(missing_group_cols, collapse = ", ")))
  }
  
  cat("Processing community states...\n")
  cat(paste("Grouping by:", paste(grouping_vars, collapse = ", "), "\n"))
  
  # Process all group combinations at once
  community_states <- node_dynamic_data %>%
    group_by(across(all_of(grouping_vars))) %>%
    group_split() %>%
    lapply(function(group) {
      transitions <- generate_state_transitions(group)
      
      # Add grouping variables to the result
      for (var in grouping_vars) {
        transitions[[var]] <- unique(group[[var]])
      }
      
      transitions
    }) %>%
    do.call(rbind, .)
  
  # Reset row names
  rownames(community_states) <- NULL
  
  # Add duration and other metrics if requested
  if (add_duration) {
    community_states <- community_states %>%
      mutate(duration = terminus - onset + 1)
  }
  
  if (add_metrics) {
    community_states <- community_states %>%
      mutate(
        n_species = ifelse(state == "none", 0, lengths(strsplit(state, ","))),
        state_id = as.numeric(factor(state))
      )
  }
  
# Reorder columns to put grouping variables first
  col_order <- c(grouping_vars, setdiff(names(community_states), grouping_vars))
  community_states <- community_states[, col_order]
  
  # Sort the results
  community_states <- community_states %>%
    arrange(across(all_of(c(grouping_vars, "onset"))))
  
  # Save to file if specified
  if (!is.null(output_file)) {
    write_rds(community_states, output_file)
    cat(paste("Results saved to:", output_file, "\n"))
  }
  
  return(community_states)
}

# Convenience wrapper for the most common use case (pond + cycle grouping)
process_pond_cycle_states <- function(node_dynamic_data, 
                                    output_suffix = NULL,
                                    output_dir = here("output")) {
  
  # Create output filename if suffix provided
  output_file <- if (!is.null(output_suffix)) {
    file.path(output_dir, paste0("frog_CommunityStates", output_suffix, ".rds"))
  } else {
    NULL
  }
  
  # Process with pond and cycle grouping
  result <- process_community_states(
    node_dynamic_data = node_dynamic_data,
    grouping_vars = c("pond", "cycle"),
    output_file = output_file,
    add_duration = TRUE,
    add_metrics = TRUE
  )
  
  return(result)
}

#run the function for all 4 versions of the data
community_states <- process_pond_cycle_states(node_dynamic, output_suffix = "")

community_states_g1 <- process_pond_cycle_states(node_dynamic_g1,output_suffix = "_g1")

community_states_g2 <- process_pond_cycle_states(node_dynamic_g2, output_suffix = "_g2")

community_states_g3 <- process_pond_cycle_states(node_dynamic_g3, output_suffix = "_g3")
```

## create transition links between states
Now that we know when states are present, we can connect states. 
```{r state_transitions, message=FALSE}
#get data
community_states<-readRDS(here("./output/frog_CommunityStates.rds"))
community_states_g1<-readRDS(here("./output/frog_CommunityStates_g1.rds"))
community_states_g2<-readRDS(here("./output/frog_CommunityStates_g3.rds"))
community_states_g3<-readRDS(here("./output/frog_CommunityStates_g3.rds"))


na_summary <- community_states %>%
  group_by(pond, cycle) %>%
  summarise(any_na = any(is.na(across(everything()))), .groups = "drop") %>%
  filter(any_na == TRUE)

# Create transitions data frame function
get_community_transitions <- function(data) {
  data %>%
    group_by(pond, cycle) %>%
    arrange(pond, cycle, onset) %>%
    mutate(
      starting_state = lag(state),
      day = onset
    ) %>%
    filter(!is.na(starting_state)) %>%
    select(pond, cycle, starting_state, end_state = state, day) %>%
    ungroup()
}

#get transitions
community_transitions<-get_community_transitions(community_states)
community_transitions_g1<-get_community_transitions(community_states_g1)
community_transitions_g2<-get_community_transitions(community_states_g2)
community_transitions_g3<-get_community_transitions(community_states_g3)

#save results
saveRDS(community_transitions, here("./output/frog_transit_states.rds"))
saveRDS(community_transitions_g1, here("./output/frog_transit_states_g1.rds"))
saveRDS(community_transitions_g2, here("./output/frog_transit_states_g2.rds"))
saveRDS(community_transitions_g3, here("./output/frog_transit_states_g3.rds"))
```

## create transition networks
Let's create network graphs and visualize some using the igraph package
```{r transition_networks, warning=FALSE}
#load data if needed
community_transitions<-readRDS(here("./output/frog_transit_states.rds"))
community_transitions_g1<-readRDS(here("./output/frog_transit_states_g1.rds"))
community_transitions_g2<-readRDS(here("./output/frog_transit_states_g2.rds"))
community_transitions_g3<-readRDS(here("./output/frog_transit_states_g3.rds"))

# Create a list of networks, one for each pond x cycle combination
community_networks_fun <- function(data){
  # Get the grouping keys for naming
  group_keys <- data %>%
    group_by(pond, cycle) %>%
    group_keys()
  
  # Create networks and names simultaneously
  networks <- data %>%
    group_by(pond, cycle) %>%
    group_split() %>%
    map(function(group_data) {
      edges <- group_data %>%
        select(starting_state, end_state)
      graph_from_data_frame(d = edges, directed = TRUE)
    }) %>%
    set_names(paste0("pond_", group_keys$pond, "_cycle_", group_keys$cycle))
  
  return(networks)
  }

#run function for different version of the data
community_networks<- community_networks_fun(community_transitions)
community_networks_g1<- community_networks_fun(community_transitions_g1)
community_networks_g2<- community_networks_fun(community_transitions_g2)
community_networks_g3<- community_networks_fun(community_transitions_g3)

#save data
saveRDS(community_networks,file=here("./output/frog_community_networks.rds"))
saveRDS(community_networks_g1,file=here("./output/frog_community_networks_g1.rds"))
saveRDS(community_networks_g2,file=here("./output/frog_community_networks_g2.rds"))
saveRDS(community_networks_g3,file=here("./output/frog_community_networks_g3.rds"))

```

# Analysis
Now that we have state transition networks, we can analyze the structure in various ways.

## Community states
First, we look at community states,unique states, the average duration of a given state, and their frequency.
```{r state_stats, warning=FALSE}
#average number of species per community state
ggplot(subset(community_states, n_species!=0),
       aes(x=n_species, fill=as.factor(cycle)))+
  geom_histogram()+
  facet_wrap(~pond)+
  scale_x_continuous(breaks = seq(0, 14, by = 1))+
  labs(x="Species per state", y="Frequency")+
  theme_bw()
  
#look at average and standard deviation of duration of individual states, how often a state occurs, and we do this for the raw data sets or those that  ignore short activity gaps (g1-g3)

#calculate mean duration of state in a given pond year
com_state_Ave<-community_states%>%
  group_by(pond, cycle,state)%>%
  summarize(meanDur=mean(duration),
            varDur=sd(duration),
            freqState=n()
            )
#repeat for all version of data set
com_state_Ave_g1<-community_states_g1%>%
  group_by(pond, cycle,state)%>%
  summarize(meanDur=mean(duration),
            varDur=sd(duration),
            freqState=n()
            )

com_state_Ave_g2<-community_states_g3%>%
  group_by(pond, cycle,state)%>%
  summarize(meanDur=mean(duration),
            varDur=sd(duration),
           freqState=n()
           )
com_state_Ave_g3<-community_states_g3%>%
  group_by(pond, cycle,state)%>%
  summarize(meanDur=mean(duration),
            varDur=sd(duration),
           freqState=n()
           )

#now let's get network annual averages across nodes 
com_state_Network<-com_state_Ave%>%
  group_by(pond, cycle)%>%
  summarize(meanDurAv=mean(meanDur),
            freqStateAv=mean(freqState),
            skewDur=ifelse(is.na(skewness(meanDur, na.rm = TRUE)), 0, skewness(meanDur, na.rm = TRUE))
            )

com_state_Network_g1<-com_state_Ave_g1%>%
  group_by(pond, cycle)%>%
  summarize(meanDurAv=mean(meanDur),
            freqStateAv=mean(freqState),
            skewDur=ifelse(is.na(skewness(meanDur, na.rm = TRUE)), 0, skewness(meanDur, na.rm = TRUE))
            )
com_state_Network_g2<-com_state_Ave_g2%>%
  group_by(pond, cycle)%>%
  summarize(meanDurAv=mean(meanDur),
            freqStateAv=mean(freqState),
            skewDur=ifelse(is.na(skewness(meanDur, na.rm = TRUE)), 0, skewness(meanDur, na.rm = TRUE))
            )

com_state_Network_g3<-com_state_Ave_g3%>%
  group_by(pond, cycle)%>%
  summarize(meanDurAv=mean(meanDur),
            freqStateAv=mean(freqState),
            skewDur=ifelse(is.na(skewness(meanDur, na.rm = TRUE)), 0, skewness(meanDur, na.rm = TRUE))
            )

# compare frequency to exponential distribution most of the observations are below 70 days, with some extreme outliers
lambda_df <- community_states %>%
  subset(duration<70) %>%
  summarise(
    lambda = 1 / mean(duration),
    .groups = "drop"
  )
#or by pond
lambda_df <- community_states %>%
  subset(duration<70) %>%
  group_by(pond) %>%
  summarise(
    lambda = 1 / mean(duration),
    .groups = "drop"
  )



# Choose a max x-value (duration range)
max_duration <- max(community_states$duration)

# Create curve points for each pond
exp_curve <- lambda_df %>%
  mutate(data = map(lambda, ~ {
    tibble(
      # Using 0.2 or 0.5 instead of 0.1 prevents the 'vertical bar' aliasing
      duration = seq(0, 60, by = 0.5), 
      density = dexp(seq(0, 60, by = 0.5), rate = .x)
    )
  })) %>%
  unnest(data)
#plot data with exponential distribution for comparison
Duration_probD_plot<-ggplot() +
  geom_area(data = exp_curve, aes(x = duration, y = density),
            fill = "#D55E00", alpha = 0.3) +
  geom_histogram(data = community_states, 
                 aes(x = duration, y = ..density..),
                 binwidth = 1, fill = "#2c3e50", alpha=0.8) +
  geom_path(data = exp_curve, aes(x = duration, y = density),
            color = "#D55E00", linewidth = 1) +
  #stat_function(fun = dexp, args = list(rate = 0.8), 
  #              geom = "area", fill = "#D55E00", alpha = 0.2)+
  #stat_function(fun = dexp, args = list(rate = 0.8), 
  #              color = "#D55E00", linewidth = 1) +
  facet_wrap(~pond,  nrow=2) +
  labs(x="Node duration (days)", y="Porbability density")+
  theme_classic()+
  theme(strip.background = element_rect(fill = "white"), 
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(), 
    axis.title = element_text(size = 12))+
  coord_cartesian(xlim = c(0, 60))

ggsave(here("Figures","Durationprop.png"),plot=Duration_probD_plot,dpi=300,width = 8, height = 8, units = "in")

```

let's calculate if duration and frequency of occurrence of a state are correlated
```{r freq_dur_state_corr, warning=FALSE}

#first calculate the correlation coefficient for different data sets
corStat_dur_freq <- com_state_Ave %>%
  group_by(pond, cycle) %>%
  summarise(correlation = cor(meanDur, freqState, method = "kendall",),
            obs=n(),
            skewDur=ifelse(is.na(skewness(meanDur, na.rm = TRUE)), 0, skewness(meanDur, na.rm = TRUE)),
            .groups = "drop")

corStat_dur_freq_g1 <- com_state_Ave_g1 %>%
  group_by(pond, cycle) %>%
  summarise(correlation = cor(meanDur, freqState, method = "kendall",),
            obs=n(),
            .groups = "drop")

corStat_dur_freq_g3 <- com_state_Ave_g3 %>%
  group_by(pond, cycle) %>%
  summarise(correlation = cor(meanDur, freqState, method = "kendall",),
            obs=n(),
            .groups = "drop")

#plot the coefficients for each pond & cycle
library(ggbeeswarm)


freq_dur_plot<-ggplot(corStat_dur_freq, aes(x=pond, y=correlation, color=cycle, size=obs))+
 geom_beeswarm(side=-1L,cex=1, priority="density", alpha=0.6)+
  stat_summary(fun.data=mean_se, show.legend = FALSE, size=1, position=position_nudge(x=0.2))+
  scale_size(range = c(1, 5)) +
    scale_color_gradient(low = "#56B4E9", high = "#E69F00") +
  labs(y="Correlation between state duration & frequency", x="Pond community")+
  theme_bw()


ggplot(corStat_dur_freq_g1, aes(x=pond, y=correlation, color=cycle, size=obs))+
  geom_beeswarm(cex=1.5, priority="density", alpha=0.6)+
  scale_size(range = c(1, 5)) +
  stat_summary(fun.data=mean_se, show.legend = FALSE, size=1)+
  scale_color_gradient(low = "#56B4E9", high = "#E69F00") +
  labs(y="Correlation between state duration & frequency", x="Pond community")+
  theme_bw()

ggplot(corStat_dur_freq_g3, aes(x=pond, y=correlation, color=cycle, size=obs))+
  geom_beeswarm(cex=1.5, priority="density", alpha=0.6)+
  scale_size(range = c(1, 5)) +
  stat_summary(fun.data=mean_se, show.legend = FALSE, size=1)+
  scale_color_gradient(low = "#56B4E9", high = "#E69F00") +
  labs(y="Correlation between state duration & frequency", x="Pond community")+
  theme_bw()

ggsave(here("Figures","StateDur_Freq.png"),plot=freq_dur_plot,dpi=300,width = 8, height = 4, units = "in")

```

## timing of transitions
There is likely to be a seasonal timing of transition between states, and we can look at how this varies across ponds and seasonal cycles. Since transitions in this type of data all have a single day duration, we can simple look at the distribution of start of nodes, which indicate when transitions happen. We are really interested in the day of the cycle when they happen. Note that we look at their temporal distribution using burstiness and memory below. For now we are really focused on the timing within a cycle. 

```{r transitiontiming}
#we need the community_states data for this which has all the info on onset and terminus
#let's start by calculating mean and median values for duration and onset
sum_start<-community_states%>%group_by(pond, cycle)%>%
  summarize(meanDayOnset=mean(onset))

sum_start_g1<-community_states_g1%>%group_by(pond, cycle)%>%
  summarize(meanDayOnset=mean(onset))
sum_start_g2<-community_states_g2%>%group_by(pond, cycle)%>%
  summarize(meanDayOnset=mean(onset))
sum_start_g3<-community_states_g3%>%group_by(pond, cycle)%>%
  summarize(meanDayOnset=mean(onset))

ggplot(sum_start,aes(x=pond, y=meanDayOnset, color=cycle))+
  geom_beeswarm(cex=1.5, priority="density", alpha=0.6)+
  stat_summary(fun.data=mean_se,size=1)+
  theme_bw()

meanDy<-lm(meanDayOnset~pond+as.factor(cycle), data=sum_start_g1)
Anova(meanDy)
```

##Event rate
we can also look at cumulative events over time in a year to see how that differs across ponds and cycles. we will define the event rate is the number of changes in community state per unit time and we calculate that by fitting a linear model to cummulative number of events.
```{r eventrate}

daily_cumulative_onsets <- community_states %>%
  count(pond, cycle, day = onset) %>%  # count onsets per day per group
  group_by(pond, cycle) %>%
  complete(day = full_seq(day, 1), fill = list(n = 0)) %>%  # fill missing days
  arrange(day) %>%
  mutate(cum_onsets = cumsum(n)) %>%
  mutate(scaled_events=cum_onsets/(max(cum_onsets)))%>%
  ungroup()

daily_cumulative_onsets_g1 <- community_states_g1 %>%
  count(pond, cycle, day = onset) %>%  # count onsets per day per group
  group_by(pond, cycle) %>%
  complete(day = full_seq(day, 1), fill = list(n = 0)) %>%  # fill missing days
  arrange(day) %>%
  mutate(cum_onsets = cumsum(n)) %>%
  mutate(scaled_events=cum_onsets/(max(cum_onsets)))%>%
  ungroup()
daily_cumulative_onsets_g2 <- community_states_g2 %>%
  count(pond, cycle, day = onset) %>%  # count onsets per day per group
  group_by(pond, cycle) %>%
  complete(day = full_seq(day, 1), fill = list(n = 0)) %>%  # fill missing days
  arrange(day) %>%
  mutate(cum_onsets = cumsum(n)) %>%
  mutate(scaled_events=cum_onsets/(max(cum_onsets)))%>%
  ungroup()
daily_cumulative_onsets_g3 <- community_states_g3 %>%
  count(pond, cycle, day = onset) %>%  # count onsets per day per group
  group_by(pond, cycle) %>%
  complete(day = full_seq(day, 1), fill = list(n = 0)) %>%  # fill missing days
  arrange(day) %>%
  mutate(cum_onsets = cumsum(n)) %>%
  mutate(scaled_events=cum_onsets/(max(cum_onsets)))%>%
  ungroup()

#plot cumulative events over time
ggplot(subset(daily_cumulative_onsets, pond==4) , aes(x=day,y=scaled_events, group=cycle, color=cycle))+
  geom_line(size=2)+
  stat_smooth(method="lm",se=F, color="gray")+
  facet_grid(pond~cycle)+
  theme_bw()+
   scale_color_viridis_c(option = "plasma", end = 0.9)

#let's fit simple linear model to this data which fits impressively well (see plot above)
eventvstime_regression <- daily_cumulative_onsets %>%
  group_by(pond, cycle) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(scaled_events ~ day, data = .x)),
    tidy_model = map(model, tidy)
  ) %>%
  unnest(tidy_model) %>%
  filter(term == "day") %>%  # get only the slope term
  select(pond, cycle, estimate, std.error, p.value) %>%
  rename(
    event.rate = estimate,
    se = std.error,
    p_value = p.value
  )

eventvstime_regression_g1 <- daily_cumulative_onsets_g1 %>%
  group_by(pond, cycle) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(scaled_events ~ day, data = .x)),
    tidy_model = map(model, tidy)
  ) %>%
  unnest(tidy_model) %>%
  filter(term == "day") %>%  # get only the slope term
  select(pond, cycle, estimate, std.error, p.value) %>%
  rename(
    event.rate = estimate,
    se = std.error,
    p_value = p.value
  )
eventvstime_regression_g2 <- daily_cumulative_onsets_g2 %>%
  group_by(pond, cycle) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(scaled_events ~ day, data = .x)),
    tidy_model = map(model, tidy)
  ) %>%
  unnest(tidy_model) %>%
  filter(term == "day") %>%  # get only the slope term
  select(pond, cycle, estimate, std.error, p.value) %>%
  rename(
    event.rate = estimate,
    se = std.error,
    p_value = p.value
  )
eventvstime_regression_g3 <- daily_cumulative_onsets_g3 %>%
  group_by(pond, cycle) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(scaled_events ~ day, data = .x)),
    tidy_model = map(model, tidy)
  ) %>%
  unnest(tidy_model) %>%
  filter(term == "day") %>%  # get only the slope term
  select(pond, cycle, estimate, std.error, p.value) %>%
  rename(
    event.rate = estimate,
    se = std.error,
    p_value = p.value
  )


#save the estimate and add it to the response variable list
eventvs.time<-eventvstime_regression%>%select(pond, cycle,event.rate)
eventvs.time_g1<-eventvstime_regression_g1%>%select(pond, cycle,event.rate)
eventvs.time_g2<-eventvstime_regression_g2%>%select(pond, cycle,event.rate)
eventvs.time_g3<-eventvstime_regression_g3%>%select(pond, cycle,event.rate)

timing_sum<-left_join(sum_start,eventvs.time)
timing_sum_g1<-left_join(sum_start_g1,eventvs.time_g1)
timing_sum_g2<-left_join(sum_start_g2,eventvs.time_g1)
timing_sum_g3<-left_join(sum_start_g3,eventvs.time_g1)

```

##Network Structure Analysis
Finally, we can look at the whole structure of the transition network. We can do this as a temporal network, or transform it into a static Markov-chain network.
The second option essentially ways the transition between nodes by the frequency at which the transitions occur. Note this option looses all temporal information except for pairwise transitions.

###Tansition graphs
```{r transitiongraph}
#function to turn transition matrix into a markov transition matrix that gives frequency of transitions between states
trans_markov<-function(data){
  data%>%
  group_by(pond, cycle,starting_state,end_state)%>%
  summarize(weight=n())%>%
  ungroup()
}

trans_markov_matrix<-trans_markov(community_transitions)
trans_markov_matrix_g1<-trans_markov(community_transitions_g1)
trans_markov_matrix_g2<-trans_markov(community_transitions_g2)
trans_markov_matrix_g3<-trans_markov(community_transitions_g3)


# save results
write_rds(trans_markov_matrix, here("./output/frog_trans_markov.rds"))
write_rds(trans_markov_matrix_g1, here("./output/frog_trans_markov_g1.rds"))
write_rds(trans_markov_matrix_g2, here("./output/frog_trans_markov_g2.rds"))
write_rds(trans_markov_matrix_g3, here("./output/frog_trans_markov_g3.rds"))

#calculated weighed networks using iGraph package
weighted_networks <- function(data){
  grouped_data<-data%>%group_by(pond, cycle) %>%group_split()
  
  graphs<-lapply(grouped_data,function(group_data) {
    # Create edges from starting_state to end_state
    edges <- group_data %>%
      select(starting_state, end_state, weight)
    # Create the graph
    graph_from_data_frame(d = edges, directed = TRUE)
    })
  
  # Name each network in the list with "pond_cycle"
  names(graphs) <- data%>%
  group_by(pond, cycle) %>%
  group_keys() %>%
  mutate(name = paste0("pond_", pond, "_cycle_", cycle)) %>%
  pull(name)
  return(graphs)
  }

community_networks_weighted<- weighted_networks(trans_markov_matrix)
community_networks_weighted_g1<- weighted_networks(trans_markov_matrix_g1)
community_networks_weighted_g2<- weighted_networks(trans_markov_matrix_g2)
community_networks_weighted_g3<- weighted_networks(trans_markov_matrix_g3)

#save data
write_rds(community_networks_weighted,file=here("./output/frog_community_networks_weighted.rds"))
write_rds(community_networks_weighted_g1,file=here("./output/frog_community_networks_weighted_g1.rds"))
write_rds(community_networks_weighted_g2,file=here("./output/frog_community_networks_weighted_g2.rds"))
write_rds(community_networks_weighted_g3,file=here("./output/frog_community_networks_weighted_g3.rds"))
```

# Network statistics on weighted transition graphs
We can now calculate a number of statistics on basic "static" weighted transition graphs, with or without weights using the basic igraph package. Statistics like betweeness(), closseness(), strength(sums up edge weights), degree(), centrality (e.g. harmonic_centrality()), etc.

## Edge diversity  
Transition between certain state pairs can occur more common, indicating a back and forth, while other transitions may be rare. We can think of this in term of either distributions of edges, weights, or evenness. If all transitions occur equally, then eveness of edges should be high, if some transitions are dominant and occur much more frequently, than eveness is low. So in simplest case, we can calculate this with Pielou's eveness index.
```{r edge_diversity}
#load edge frequency matrix
trans_freq<-read_rds(here("./output/frog_trans_markov.rds"))
trans_freq_g1<-read_rds(here("./output/frog_trans_markov_g1.rds"))
trans_freq_g2<-read_rds(here("./output/frog_trans_markov_g2.rds"))
trans_freq_g3<-read_rds(here("./output/frog_trans_markov_g3.rds"))

#calculate skenwess and frequency of transitions
calc_skew<-function(df){
  df%>%
    group_by(pond, cycle)%>%
    summarize(
      skewEdge=ifelse(is.na(skewness(weight, na.rm = TRUE)), 0, skewness(weight, na.rm = TRUE)),
      total_weight=sum(weight),
      ave_Edge_weight=mean(weight),
      .groups="drop"
    )
}
Trans_evenness <- calc_skew(trans_freq)%>%select(pond, cycle,skewEdge,ave_Edge_weight)
print(Trans_evenness)

Trans_evenness_g1 <- calc_skew(trans_freq_g1)%>%select(pond, cycle,skewEdge,ave_Edge_weight)
Trans_evenness_g2 <- calc_skew(trans_freq_g2)%>%select(pond, cycle,skewEdge,ave_Edge_weight)
Trans_evenness_g3 <- calc_skew(trans_freq_g3)%>%select(pond, cycle,skewEdge,ave_Edge_weight)

```

##Network metrics
Let's calcuate more "typical" network metrics
```{r transition_metrics}
#load data
community_networks_weighted<-read_rds(here("./output/frog_community_networks_weighted.rds"))
community_networks_weighted_g1<-read_rds(here("./output/frog_community_networks_weighted_g1.rds"))
community_networks_weighted_g2<-read_rds(here("./output/frog_community_networks_weighted_g2.rds"))
community_networks_weighted_g3<-read_rds(here("./output/frog_community_networks_weighted_g3.rds"))

#note that we have weighted graphs, but in igraph weights indicate a "cost". In our case, we assume connections are stronger with higher weights, so we need to use the inverse of weights for weighted metrics.


#This function calculates various network metrics for a list of igraph objects
analyze_network_metrics <- function(network_list, join_data = NULL) {
  # Get network count
  network_count <- length(network_list)
  
  # Initialize results dataframe
  results <- data.frame(
    Pond_Cycle = character(network_count),
    Connectance = numeric(network_count),
    Weighted_connectance = numeric(network_count),
    Unique_Nodes = numeric(network_count),
    UniqueEdges = numeric(network_count),
    TotalEdges = numeric(network_count),
    modularity = numeric(network_count),
    mean_Distance = numeric(network_count),
    harm.centrality = numeric(network_count),
    betweenness = numeric(network_count),
    betweenness_weighted = numeric(network_count),
    stringsAsFactors = FALSE
  )
  
  # Loop through and fill the data frame by index
  for(i in seq_len(network_count)) {
    name <- names(network_list)[i]
    graph <- network_list[[name]]
    
    # Skip if there's any problem with the graph
    if(!is(graph, "igraph")) {
      results[i, ] <- NA
      next
    }
    
     # Check for trivial or empty graphs
    if (gorder(graph) < 2 || gsize(graph) == 0) {
      warning("Graph too small or empty: ", name)
      results[i, ] <- NA
      next
    }
    
    # Fill in the values for this row
    results$Pond_Cycle[i] <- name
    results$Unique_Nodes[i] <- gorder(graph) #indicates unique community states
    results$UniqueEdges[i] <- ecount(graph) #indicates the unique number of edges (transitions)
    results$TotalEdges[i]<-sum(E(graph)$weight) #this sums all the edges (transitions) based on their weight.
    
     # Calculate connectance (directed graph formula, excludes self loops not including weights)
    results$Connectance[i] <- results$UniqueEdges[i] / 
    (results$Unique_Nodes[i] * (results$Unique_Nodes[i] - 1))
    #same but with weights
    results$Weighted_connectance[i] <- sum(E(graph)$weight) / (gorder(graph) * (gorder(graph) - 1) / 2)
    
    # Calculate modularity using walktrap algorithm
    results$modularity[i] <- modularity(cluster_walktrap(graph))
    
    # Calculate mean distance weighing inversely by frequency
    results$mean_Distance[i] <- mean_distance(graph,weights = 1/E(graph)$weight)
    
    # Calculate centralization measures
    results$harm.centrality[i] <- mean(harmonic_centrality(graph, mode = "all", weights = 1/E(graph)$weight, normalized = TRUE)) #harmonic sets disconnect =0 not Inf
    results$betweenness[i] <- centr_betw(graph)$centralization #Freeman version
    
    # Weighted Betweenness centralization formula (Freeman 1977)
    btw <- betweenness(graph, weights = 1/E(graph)$weight, normalized = FALSE)

    # Number of nodes
    n <- gorder(graph)
    # Weighted Betweenness centralization formula (Freeman 1977)
    if (n > 2) {
      max_diff_sum <- sum(max(btw) - btw)
      theoretical_max <- (n - 1) * (n - 2)
      central_betweenness <- max_diff_sum / theoretical_max
    } else {
      central_betweenness <- NA
    }
    results$betweenness_weighted[i]=central_betweenness

  }
  
  # Extract pond and cycle information from names
  results <- results %>%
    mutate(
      pond = as.factor(sub("pond_([^_]+)_cycle_.*", "\\1", Pond_Cycle)),
      cycle = as.numeric(sub(".*_cycle_([^_]+)", "\\1", Pond_Cycle))
    )
  
  # Join with additional data if provided
  if(!is.null(join_data)) {
    results <- inner_join(results, join_data, by = intersect(names(results), names(join_data)))
  }
  
  # Remove any rows with NA values (from skipped networks)
  #results <- results%>%filter(!is.na)
  results <- results %>%
    mutate(across(everything(), ~replace_na(., NA)))
  return(results)
}

sp_num<-read_rds(here("./output/frog_pond_cycle_NoSpec.rds"))

state_connectance_w <- analyze_network_metrics(community_networks_weighted, sp_num)
state_connectance_w_g1 <- analyze_network_metrics(community_networks_weighted_g1, sp_num)
state_connectance_w_g2 <- analyze_network_metrics(community_networks_weighted_g2, sp_num)
state_connectance_w_g3 <- analyze_network_metrics(community_networks_weighted_g3, sp_num)%>%mutate(pond=as.factor(pond))
```

## Burstiness
We will calculate temporal distribution of events, in this case when a new state appears. This gives us a good idea of when and how changes are clumped in time or not. Values=1 indicate perfect (periodic) spacing, = 0 random, >0 clumping. Note, we need to decide if we include empty states in this calculation
```{r burstiness}
#get data, we'll just use basic community state data for simplicity but could do this with network data too
community_states<-read_rds(here("./output/frog_CommunityStates.rds"))
community_statesNonone<-read_rds(here("./output/frog_CommunityStates.rds"))%>%subset(state!="none") #this is a subset of the data where we exclude periods without any species being active (so no "empty" states)
community_states_g1<-read_rds(here("./output/frog_CommunityStates_g1.rds"))
community_states_g2<-read_rds(here("./output/frog_CommunityStates_g2.rds"))
community_states_g3<-read_rds(here("./output/frog_CommunityStates_g3.rds"))

# Function to compute burstiness coefficient B
calculate_burstiness <- function(df) {
  df %>%
    group_by(pond, cycle) %>%  # Group by pond and cycle
    arrange(onset) %>%  # Ensure events are ordered by time
    summarise(
      mean_t = mean(diff(onset), na.rm = TRUE),
      sd_t = sd(diff(onset), na.rm = TRUE),
      B = ifelse(sd_t + mean_t > 0, (sd_t - mean_t) / (sd_t + mean_t), NA_real_),  # Burstiness formula
      .groups = "drop"
    )
}

# Calculate it for all data sets with different gap rules
burstiness_results<-calculate_burstiness(community_states) 
burstiness_resultsNonone<-calculate_burstiness(community_statesNonone)
burstiness_results_g1<-calculate_burstiness(community_states_g1)
burstiness_results_g2<-calculate_burstiness(community_states_g2)
burstiness_results_g3<-calculate_burstiness(community_states_g3)

#let's check if including the empty state or ignoring gaps alters general patterns (it doesn't, just shifts curves down a bit)
ggplot(burstiness_results, aes(x=cycle,y=B, color=as.factor(pond)))+
  stat_smooth(data=burstiness_results, aes(x=cycle,y=B, color=as.factor(pond)), linetype=1, se=F)+
  geom_point(alpha=0.5)+
  stat_smooth(data=burstiness_resultsNonone, aes(x=cycle,y=B, color=as.factor(pond)), linetype=2, se=F)+
  stat_smooth(data=burstiness_results_g1, aes(x=cycle,y=B),color="blue",  se=F)+
  stat_smooth(data=burstiness_results_g3, aes(x=cycle,y=B),color="darkgrey",  se=F)+
  labs(x="Pond", y="Burstiness, B")+
  facet_wrap(~pond)+
  theme_bw()

```

## Memory
Now we calculate memory, which is complementary to burstiness and examines how correlated events are. >0 indicates that short interevent follow short events, =0 means it's random, <0 means long follow short and vice versa. Basically a 1-lag autocorrelation coefficient
```{r memory}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(purrr)

# Calculate memory metrics for a given pond and cycle
calculate_memory <- function(data, pond_val, cycle_val) {
  # Sort by onset time
  data <- data %>% arrange(onset)
  
  # Calculate time gaps between consecutive states
  data <- data %>%
    mutate(
      prev_state = lag(state),
      prev_terminus = lag(terminus),
      time_gap = onset - lag(terminus),
      state_transition = paste(lag(state), "->", state)
    )
  
  # Calculate state frequencies
  state_counts <- table(data$state)
  state_freq <- state_counts / nrow(data)
  
  # Calculate autocorrelation for durations
  ac <- tryCatch({
    acf_result <- acf(data$duration, lag.max = min(10, nrow(data)-1), plot = FALSE)
    ac_values <- as.numeric(acf_result$acf)
    names(ac_values) <- paste0("ac_lag_", 0:(length(ac_values)-1))
    ac_values
  }, error = function(e) {
    ac_values <- rep(NA, 11)
    names(ac_values) <- paste0("ac_lag_", 0:10)
    ac_values
  })
  
  # Calculate memory metrics
  # Markov memory: How well previous state predicts current state
  state_repeats <- sum(data$state == data$prev_state, na.rm = TRUE) / max(1, (nrow(data) - 1))
  
  # Time memory: Correlation between consecutive durations
  time_memory <- tryCatch({
    if(nrow(data) > 1 && sd(data$duration, na.rm = TRUE) > 0) {
      durations1 <- data$duration[-length(data$duration)]
      durations2 <- data$duration[-1]
      if(sd(durations1, na.rm = TRUE) > 0 && sd(durations2, na.rm = TRUE) > 0) {
        cor(durations1, durations2, use = "complete.obs")
      } else {
        NA_real_
      }
    } else {
      NA_real_
    }
  }, error = function(e) {
    NA_real_
  })
  
  # Mean return time (how long until a state recurs)
  return_times_df <- tryCatch({
    data %>%
      group_by(state) %>%
      mutate(
        next_occurrence = lead(onset),
        return_time = ifelse(!is.na(next_occurrence), next_occurrence - onset, NA)
      ) %>%
      summarise(mean_return_time = mean(return_time, na.rm = TRUE), .groups = "drop")
  }, error = function(e) {
    data.frame(state = character(), mean_return_time = numeric())
  })
  
  # Convert return times to a more compact representation
  return_times_str <- if(nrow(return_times_df) > 0) {
    paste(return_times_df$state, round(return_times_df$mean_return_time, 2), 
          sep = ":", collapse = "; ")
  } else {
    NA_character_
  }
  
  # Return as a single-row data frame with all metrics
  result_df <- data.frame(
    # Don't include pond and cycle here - they're already in the outer data frame
    markov_memory = state_repeats,
    time_memory = time_memory
    )
  return(result_df)
}

# Apply memory calculation to each pond and cycle combination
memory_analysis <- function(community_states) {
  results <- community_states %>%
    group_by(pond, cycle) %>%
    nest() %>%
    mutate(
      memory_metrics = pmap(
        list(data = data, pond_val = pond, cycle_val = cycle),
        function(data, pond_val, cycle_val) {
          calculate_memory(data, pond_val, cycle_val)
        }
      )
    )
  
  #Unnest the results
  results_df <- results %>%
    select(-data) %>%
    unnest(memory_metrics)
  
  return(results_df)
}

# Summarize memory metrics across ponds and cycles
summarize_memory <- function(memory_results) {
  summary_stats <- memory_results %>%
    summarise(across(where(is.numeric), 
                    list(mean = ~mean(.x, na.rm = TRUE),
                         sd = ~sd(.x, na.rm = TRUE),
                         min = ~min(.x, na.rm = TRUE),
                         max = ~max(.x, na.rm = TRUE)),
                    .names = "{.col}_{.fn}"))
  
  return(summary_stats)
}

memory_results <- memory_analysis(community_states) 
memory_results_g1 <- memory_analysis(community_states_g1) 
memory_results_g2 <- memory_analysis(community_states_g2)
memory_results_g3 <- memory_analysis(community_states_g3)


#compare memory across data sets using different gap cutoffs
ggplot(memory_results, aes(x=cycle, y=time_memory, group=pond,color=pond))+
  geom_point()+
  geom_line(se=F)+
  geom_line(data=memory_results_g1, aes(x=cycle, y=time_memory, group=pond,color=pond),color="black", se=F)+
  geom_line(data=memory_results_g2, aes(x=cycle, y=time_memory, group=pond,color=pond),color="darkgrey",linetype=2, se=F)+
  geom_line(data=memory_results_g3, aes(x=cycle, y=time_memory, group=pond,color=pond),color="lightgrey", se=F)+
  theme_bw()+
  facet_wrap(~pond)

```

Lets combine both with all the other network metrics into one data file
```{r combine_metrics, warning=FALSE}

# Assume list of data frames
df_list <- list(state_connectance_w,Trans_evenness,burstiness_results,memory_results, com_state_Network,timing_sum)
df_list_g1 <- list(state_connectance_w_g1,Trans_evenness_g1,burstiness_results_g1,memory_results_g1, com_state_Network_g1,timing_sum_g1)
df_list_g2 <- list(state_connectance_w_g2,Trans_evenness_g2,burstiness_results_g2,memory_results_g2,com_state_Network_g2,timing_sum_g2)
df_list_g3 <- list(state_connectance_w_g3,Trans_evenness_g3,burstiness_results_g3,memory_results_g3,com_state_Network_g3, timing_sum_g3)

# Join all by "id" using left_join
Transition_metrics <- reduce(df_list, left_join, by = c("pond", "cycle")) %>%
  select(pond, cycle,Pond_Cycle,No_Spec, everything())
Transition_metrics_g1 <- reduce(df_list_g1, left_join, by = c("pond", "cycle")) %>%
  select(pond, cycle,Pond_Cycle,No_Spec, everything())
Transition_metrics_g2 <- reduce(df_list_g2, left_join, by = c("pond", "cycle")) %>%
  select(pond, cycle,Pond_Cycle,No_Spec, everything())
Transition_metrics_g3 <- reduce(df_list_g3, left_join, by = c("pond", "cycle")) %>%
  select(pond, cycle,Pond_Cycle,No_Spec, everything())

#save the data
write_rds(Transition_metrics, here("./output/frog_transition_matrix_metrics.rds"))
write_rds(Transition_metrics_g1, here("./output/frog_transition_matrix_metrics_g1.rds"))
write_rds(Transition_metrics_g2, here("./output/frog_transition_matrix_metrics_g2.rds"))
write_rds(Transition_metrics_g3, here("./output/frog_transition_matrix_metrics_g3.rds"))
```

#lets plot all response variables by pond and year
```{r}
Transition_metrics<-readRDS(here("./output/frog_transition_matrix_metrics.rds"))

library(tidyverse)
library(ggbeeswarm)
library(patchwork)
library(here)

# 1. Define your standard response list outside the function
response_list <- c(
  "Unique_Nodes" = "Unique nodes",
  "Connectance" = "Connectance",
  "betweenness" = "Betweenness",
  "mean_Distance" = "Mean distance",
  "modularity" = "Modularity",
  "freqStateAv" = "Node frequency",
  "meanDurAv" = "Node duration",
  "skewDur" = "Duration skewness",
  "TotalEdges" = "Total edges",
  "ave_Edge_weight" = "Edge frequency",
  "skewEdge" = "Skewness edge freq",
  "event.rate" = "Event rate",
  "B" = "Burstiness",
  "time_memory" = "Memory"
)

generate_metric_panel <- function(data, file_name) {
  
  # Internal helper to create individual plots
  make_individual_plot <- function(col_name, label_name) {
    ggplot(data, aes(x = factor(pond), y = .data[[col_name]], color = cycle)) +
      geom_beeswarm(side = -1L, cex = 1.5, alpha = 0.7, priority = "random") +
      stat_summary(fun.data = mean_se, size = 0.8, color = "black", position = position_nudge(x = 0.2)) +
      scale_color_gradient(low = "#56B4E9", high = "#E69F00") +
      labs(x = "Pond", y = label_name) +
      theme_bw() +
      theme(axis.title.y = element_text(size = 9),
            axis.text = element_text(size = 8))
  }
  
  # 2. Map through the response list to create all plots
  # imap creates a loop over both names (col_name) and values (label_name)
  plot_list <- imap(response_list, ~make_individual_plot(.y, .x))
  
  # 3. Combine with Patchwork
  combined_plot <- wrap_plots(plot_list) +
    guide_area() +
    plot_layout(ncol = 5, guides = "collect") &
    theme(
      legend.key.size = unit(0.4, "cm"), 
      legend.text = element_text(size = 8),
      legend.title = element_text(size = 9),
      legend.margin = margin(0, 0, 0, 0),
      legend.box.margin = margin(-5, -5, -5, -5)
    )
  
  # 4. Save the output
  full_path <- here("Figures", paste0(file_name, ".png"))
  ggsave(full_path, plot = combined_plot, dpi = 300, width = 12, height = 8, units = "in")
  
  return(combined_plot)
}

#run function for different data sets
generate_metric_panel(Transition_metrics, "MetricbyPondYear")
generate_metric_panel(Transition_metrics_g1, "MetricbyPondYear_g1")
generate_metric_panel(Transition_metrics_g2, "MetricbyPondYear_g2")
generate_metric_panel(Transition_metrics_g3, "MetricbyPondYear_g3")


```
Now let's look at how node transitions is linked to species diversity and possible number of states. The goal is to calculate all possible states for a certain number of species, ignoring temporal partitioning, and see how this scales with number of species and realized number of community states we observe
```{r}
#look at unique node vs species with predicted expectation
nodvsspecDiv_plot<-ggplot(Transition_metrics, aes(x=(No_Spec), y=(Unique_Nodes)))+
  #plot points
  geom_beeswarm(size=3, alpha=0.6, color="#1b7837")+
  labs(x="Annual species richness ", y="Node diversity")+
  #fit simple linear model (2nd order polynomial did not fit better)
  stat_smooth(method=lm,
              color="#1b7837",
              fill="#1b7837", 
              alpha=0.2, 
              size=2)+
  #calculate possible number of states for a given number of species ignoring temporal separation
  stat_function(fun = function(x) 2^x - 1, color = "#762a83", size = 2)+
  scale_y_log10() +
  scale_x_continuous(breaks = seq(1, max(Transition_metrics$No_Spec, na.rm = TRUE), by = 1)) +
  theme_classic()+
  theme(axis.text = element_text(size = 10),
    axis.title = element_text(size = 16))

#safe figure
ggsave(here("Figures","NodevsSpecDiv.png"),plot=nodvsspecDiv_plot,dpi=600,width = 8, height = 6, units = "in")

```


##Metric averages
Let's get averages across pond and time to desribe general patterns
```{r}
#  SE function 
se <- function(x, na.rm = FALSE) {
  if (na.rm) x <- x[!is.na(x)]
  sd(x) / sqrt(length(x))
}

# Now run the summary
summary_table <- Transition_metrics %>%
  select(-pond, -cycle, -Pond_Cycle) %>% 
  summarise(across(everything(), list(mean = mean, se = se), na.rm = TRUE)) %>%
  pivot_longer(everything(), 
               names_to = c("metric", ".value"), 
               names_pattern = "(.*)_(mean|se)")

print(summary_table)
```

## test for pond and cycle differences
Here we simply want to test if there are significant differences among ponds and within ponds across years. In the later analyses on climate, we will include pond as random effect and cycle will be instead represented by the respective annual climate data. 
```{r lmAnalyses}
#function to get statistics (same as above)
library(tidyverse)
library(mgcv)
library(broom)
library(flextable)
library(here)

run_temporal_network_analysis <- function(data, file_name, labels_vec = NULL) {
  
  # Prepare Data
  # Ensure cycle is treated as both numeric (for GAM) and factor (for potential grouping)
  data <- data %>%
    mutate(cycleC = as.factor(cycle),
           cycle = as.numeric(cycle))
  
  # Define the 15 metrics used in your network framework
  metrics <- c("Unique_Nodes", "Connectance", "betweenness", 
               "modularity", "mean_Distance", "meanDurAv", "freqStateAv", 
               "skewDur", "TotalEdges", "ave_Edge_weight", "skewEdge", 
               "B", "event.rate", "time_memory")
  
  # Run GAM Interaction Analysis
  res_gam <- map_dfr(metrics, function(response) {
    
    # Fit the Full Model: Global trend + Pond-specific deviations (fs basis)
    # k=4 is used as per your study design to avoid over-smoothing 24 years
    form_full <- as.formula(paste(response, "~ pond + s(cycle, k = 4) + s(cycle, pond, bs = 'fs', k = 4)"))
    mod_full <- gam(form_full, data = data, method = "ML")
    
    a_gam <- anova(mod_full)
    
    # Extract Parametric and Smooth terms
    bind_rows(
      data.frame(term = "pond", 
                 statistic = a_gam$pTerms.table["pond", "F"], 
                 p.value = a_gam$pTerms.table["pond", "p-value"]),
      data.frame(term = "cycle", 
                 statistic = a_gam$s.table["s(cycle)", "F"], 
                 p.value = a_gam$s.table["s(cycle)", "p-value"]),
      data.frame(term = "pond:cycle", 
                 statistic = a_gam$s.table[str_detect(rownames(a_gam$s.table), "cycle,pond"), "F"], 
                 p.value = a_gam$s.table[str_detect(rownames(a_gam$s.table), "cycle,pond"), "p-value"])
    ) %>% 
    mutate(response = response)
  })
  
  #great table output
  # Use labels if provided, otherwise keep raw metric names
  if(!is.null(labels_vec)) {
    res_gam <- res_gam %>% mutate(response = labels_vec[response])
  }
  
  sig_table <- res_gam %>%
    select(response, term, statistic, p.value) %>%
    mutate(p.value = round(p.value, 4), 
           statistic = round(statistic, 2)) %>%
    pivot_wider(
      names_from = term, 
      values_from = c(statistic, p.value),
      names_glue = "{term}_{.value}",
      names_vary = "slowest"
    ) %>%
    select(response, starts_with("pond_"), starts_with("cycle_"), contains(":"))

  # 4. Create and Save Flextable
  ft <- sig_table %>%
    flextable() %>%
    separate_header(split = "_") %>%
    labelizor(labels = c("statistic" = "F", "p.value" = "p"), part = "header") %>%
    theme_booktabs() %>%
    bold(part = "header") %>%
    align(align = "center", part = "all") %>%
    set_caption(paste("Table: Pond and Cycle effect for", file_name)) %>%
    autofit()
  
  save_as_docx(ft, path = here("Output", paste0(file_name, ".docx")))
  
  return(ft)
}
```
now run it with all data sets
```{r}
run_temporal_network_analysis(Transition_metrics_g, "Anova_PondCycle", variable_labels)
run_temporal_network_analysis(Transition_metrics_g1, "Anova_PondCycle_G1", variable_labels)
run_temporal_network_analysis(Transition_metrics_g2, "Anova_PondCycle_G2", variable_labels)
run_temporal_network_analysis(Transition_metrics_g3, "Anova_PondCycle_G3", variable_labels)

```


