---
title: "Sequential network states"
author: "Volker Rudolf"
date: "2025-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Goals
The goal of this analysis is to turn species phenologies into a network of community state transitions. When species co-occur, they form a community state, and each species combination is a unique state, which can occur more than once. Each state has a start and ending date, and transitions into a new state. This creates a network that we will analyze to better understand how communities change

# Load packages
```{r, warning=FALSE, message=FALSE}
 #Load packages
  #list of packages that are needed
  pack<-c("tidyverse",
          "tsna",
          "ndtv",
          "knitr",
          "here",
          "igraph",
          "ggraph"
          )
  #check whether packages are installed and load them, if not get and then load
 package.check <- lapply(pack, FUN = function(x){
    if (!require(x, character.only = TRUE)) 
      {
        install.packages(x, dependencies = TRUE)
        library(x, character.only = TRUE)
      }
    })
   #search() #verify that packages are loaded

```

#Get & transform data into network

## get the data
We assume the data comes in the form of "Dynamic Nodes" from the "network Code", i.e. each observation (row) indicates when (e.g. the day) a species appears (onset) and disappears (terminus). This can happen multiple times for a given species. So there are at least 3 columns: 1) species ID, 2) onset, 3) terminus
```{r}
#get previously curated data
#raw data
node_dynamic<-read_csv(here("./output/frogs_node_dynamic.csv"))%>%select("pond", "cycle", "onset", "terminus", "species")%>%relocate(species,.before=onset)%>%subset(cycle<23)

#get number of species per pond x cycle

sp_num<-node_dynamic%>%
  group_by(pond, cycle)%>%
  summarize(No_Spec=n_distinct(species),.groups="drop")
write_csv(sp_num, here("./output/frog_pond_cycle_NoSpec.csv"))  

#data but ignoring gaps of 1, 2 or 3 days between calling of the same species.
node_dynamic_g1<-read_csv(here("./output/frogs_node_dynamic_gap1.csv"))%>%select("pond", "cycle", "onset", "terminus", "species")%>%relocate(species,.before=onset)%>%subset(cycle<23)
node_dynamic_g2<-read_csv(here("./output/frogs_node_dynamic_gap2.csv"))%>%select("pond", "cycle", "onset", "terminus", "species")%>%relocate(species,.before=onset)%>%subset(cycle<23)
node_dynamic_g3<-read_csv(here("./output/frogs_node_dynamic_gap3.csv"))%>%select("pond", "cycle", "onset", "terminus", "species")%>%relocate(species,.before=onset)%>%subset(cycle<23)

```

##organize data into community states
We will create a new data frame, that includes column for each unique species combination, and additional columns that indicate when the state starts and ends. Note, in same cases this might include state without species, which we call "none" here.
```{r}
generate_state_transitions <- function(df) {
    # Validate input
  if (nrow(df) == 0) {
    return(data.frame(state = character(), onset = integer(), terminus = integer()))
  }
  
  required_cols <- c("onset", "terminus", "species")
  if (!all(required_cols %in% names(df))) {
    stop(paste("Missing required columns:", paste(setdiff(required_cols, names(df)), collapse = ", ")))
  }
  
  # Get all unique transition points (both onset and terminus)
  transition_points <- sort(unique(c(min(df$onset), df$onset, df$terminus + 1)))
  
  # Initialize results list
  results <- vector("list", length(transition_points) - 1)
  
  # Process each time window between transition points
  for (i in seq_along(transition_points)[-length(transition_points)]) {
    current_day <- transition_points[i]
    next_transition <- transition_points[i + 1]
    
    # Find species present during this window
    present_species <- df$species[df$onset <= current_day & df$terminus >= current_day]
    
    # Create state string - if no species present, use "none"
    state <- if (length(present_species) == 0) "none" else paste(sort(present_species), collapse = ",")
    
    results[[i]] <- data.frame(
      state = state,
      onset = current_day,
      terminus = next_transition - 1
    )
  }
  
  # Combine results
  result_df <- do.call(rbind, results)
  
  # Add final "none" state if needed
  max_terminus <- max(df$terminus)
  if (tail(result_df$terminus, 1) < max_terminus) {
    final_none <- data.frame(
      state = "none",
      onset = tail(result_df$terminus, 1) + 1,
      terminus = max_terminus
    )
    result_df <- rbind(result_df, final_none)
  }
  
  return(result_df)
}
# Main function to process community states for any dataset
process_community_states <- function(node_dynamic_data, 
                                   grouping_vars = c("pond", "cycle"),
                                   output_file = NULL,
                                   add_duration = TRUE,
                                   add_metrics = TRUE) {
  
  # Validate inputs
  if (nrow(node_dynamic_data) == 0) {
    stop("Input data is empty")
  }
  
  required_cols <- c("onset", "terminus", "species")
  missing_cols <- setdiff(required_cols, names(node_dynamic_data))
  if (length(missing_cols) > 0) {
    stop(paste("Missing required columns:", paste(missing_cols, collapse = ", ")))
  }
  
  missing_group_cols <- setdiff(grouping_vars, names(node_dynamic_data))
  if (length(missing_group_cols) > 0) {
    stop(paste("Missing grouping columns:", paste(missing_group_cols, collapse = ", ")))
  }
  
  cat("Processing community states...\n")
  cat(paste("Grouping by:", paste(grouping_vars, collapse = ", "), "\n"))
  
  # Process all group combinations at once
  community_states <- node_dynamic_data %>%
    group_by(across(all_of(grouping_vars))) %>%
    group_split() %>%
    lapply(function(group) {
      transitions <- generate_state_transitions(group)
      
      # Add grouping variables to the result
      for (var in grouping_vars) {
        transitions[[var]] <- unique(group[[var]])
      }
      
      transitions
    }) %>%
    do.call(rbind, .)
  
  # Reset row names
  rownames(community_states) <- NULL
  
  # Add duration and other metrics if requested
  if (add_duration) {
    community_states <- community_states %>%
      mutate(duration = terminus - onset + 1)
  }
  
  if (add_metrics) {
    community_states <- community_states %>%
      mutate(
        n_species = ifelse(state == "none", 0, lengths(strsplit(state, ","))),
        state_id = as.numeric(factor(state))
      )
  }
  
# Reorder columns to put grouping variables first
  col_order <- c(grouping_vars, setdiff(names(community_states), grouping_vars))
  community_states <- community_states[, col_order]
  
  # Sort the results
  community_states <- community_states %>%
    arrange(across(all_of(c(grouping_vars, "onset"))))
  
  # Save to file if specified
  if (!is.null(output_file)) {
    write_csv(community_states, output_file)
    cat(paste("Results saved to:", output_file, "\n"))
  }
  
  # Print summary
  cat(paste("Total community states generated:", nrow(community_states), "\n"))
  
  if (add_metrics) {
    cat("Summary statistics:\n")
    summary_stats <- community_states %>%
      summarise(
        unique_states = n_distinct(state),
        avg_duration = round(mean(duration, na.rm = TRUE), 2),
        max_species = max(n_species, na.rm = TRUE),
        empty_states = sum(state == "none", na.rm = TRUE),
        .groups = "drop"
      )
    print(summary_stats)
  }
  
  return(community_states)
}

# Convenience wrapper for the most common use case (pond + cycle grouping)
process_pond_cycle_states <- function(node_dynamic_data, 
                                    output_suffix = NULL,
                                    output_dir = here("output")) {
  
  # Create output filename if suffix provided
  output_file <- if (!is.null(output_suffix)) {
    file.path(output_dir, paste0("frog_CommunityStates", output_suffix, ".csv"))
  } else {
    NULL
  }
  
  # Process with pond and cycle grouping
  result <- process_community_states(
    node_dynamic_data = node_dynamic_data,
    grouping_vars = c("pond", "cycle"),
    output_file = output_file,
    add_duration = TRUE,
    add_metrics = TRUE
  )
  
  return(result)
}

community_states <- process_pond_cycle_states(node_dynamic, output_suffix = "")

community_states_g1 <- process_pond_cycle_states(node_dynamic_g1,output_suffix = "_g1")

community_states_g2 <- process_pond_cycle_states(node_dynamic_g2, output_suffix = "_g2")

community_states_g3 <- process_pond_cycle_states(node_dynamic_g3, output_suffix = "_g3")
```

##create transition links between states
Now that we know when states are present, we can connect states. An important consideration here is how to deal with empty communities. e.g. if communities are empty between states and the period is short, do we model transitions including empty state, or just between communities? Depends on the question and code should be modified based on those. For now, we include all empty states.
```{r state transitions, message=FALSE}
community_states<-read_csv(here("./output/frog_CommunityStates.csv"))
community_states_g1<-read_csv(here("./output/frog_CommunityStates_g1.csv"))
community_states_g2<-read_csv(here("./output/frog_CommunityStates_g3.csv"))
community_states_g3<-read_csv(here("./output/frog_CommunityStates_g3.csv"))

# Create transitions data frame function
get_community_transitions <- function(data) {
  data %>%
    group_by(pond, cycle) %>%
    arrange(pond, cycle, onset) %>%
    mutate(
      starting_state = lag(state),
      day = onset
    ) %>%
    filter(!is.na(starting_state)) %>%
    select(pond, cycle, starting_state, end_state = state, day) %>%
    ungroup()
}

#get transitions
community_transitions<-get_community_transitions(community_states)
community_transitions_g1<-get_community_transitions(community_states_g1)
community_transitions_g2<-get_community_transitions(community_states_g2)
community_transitions_g3<-get_community_transitions(community_states_g3)



#save results
write_csv(community_transitions, here("./output/frog_transit_states.csv"))
write_csv(community_transitions_g1, here("./output/frog_transit_states_g1.csv"))
write_csv(community_transitions_g2, here("./output/frog_transit_states_g2.csv"))
write_csv(community_transitions_g3, here("./output/frog_transit_states_g3.csv"))
```

#create transition networks
Now we will use the igraph package to create graphs to visualize the networks
```{r transition networks}
community_transitions<-read_csv(here("./output/frog_transit_states.csv"))
community_transitions_g1<-read_csv(here("./output/frog_transit_states_g1.csv"))
community_transitions_g2<-read_csv(here("./output/frog_transit_states_g2.csv"))
community_transitions_g3<-read_csv(here("./output/frog_transit_states_g3.csv"))


#let's run a test for a simple pond, note it will assume first two columns are the node pairs for the edges
testP<-community_transitions_g3%>%
  subset(pond==4 & cycle==10, select=c(-pond, -cycle))

test.network <- graph_from_data_frame(d = testP, directed = T)

#basic igraph
plot(test.network, layout=layout.circle, main="circle")

#fancy version, that uses different layout and colors nodes and edges by duration and time of year
ggraph(test.network, layout = layout_as_star(
    test.network, center = which.max(degree(test.network))[[1]]
)) +
    geom_edge_fan(
      aes(color = day),
        arrow = arrow(length = unit(0.2, "inches")),
        end_cap = circle(0.215, "inches"),
        strength = 4,
        width = 1.5,
        alpha = 0.4
    ) +
    geom_node_point(size=6, color="gray") +
    geom_node_text(aes(label = name), size = 2, repel = TRUE) +
    scale_edge_color_gradient(
        low = "gray24",
        high = "greenyellow",
    ) 

```
Now let's create networks for all pond x cycle combinations
```{r}
# Create a list of networks, one for each pond x cycle combination
community_networks_fun <- function(data){
  # Get the grouping keys for naming
  group_keys <- data %>%
    group_by(pond, cycle) %>%
    group_keys()
  
  # Create networks and names simultaneously
  networks <- data %>%
    group_by(pond, cycle) %>%
    group_split() %>%
    map(function(group_data) {
      edges <- group_data %>%
        select(starting_state, end_state)
      graph_from_data_frame(d = edges, directed = TRUE)
    }) %>%
    set_names(paste0("pond_", group_keys$pond, "_cycle_", group_keys$cycle))
  
  return(networks)
  }

community_networks<- community_networks_fun(community_transitions)
community_networks_g1<- community_networks_fun(community_transitions_g1)
community_networks_g2<- community_networks_fun(community_transitions_g2)
community_networks_g3<- community_networks_fun(community_transitions_g3)

#save data
saveRDS(community_networks,file=here("./output/frog_community_networks.rds"))
saveRDS(community_networks_g1,file=here("./output/frog_community_networks_g1.rds"))
saveRDS(community_networks_g2,file=here("./output/frog_community_networks_g2.rds"))
saveRDS(community_networks_g3,file=here("./output/frog_community_networks_g3.rds"))

```

#Analysis
We can look at this in a lot of different ways. 

##Community states
First, we can just look at community states,either at the distributions, or average duration of a given state, or unique states etc.
```{r}
community_states<-read_csv(here("./output/frog_CommunityStates.csv"))
#now look at averages of states and frequency

com_state_Ave<-community_states%>%
  group_by(pond, cycle,state)%>%
  summarize(meanDur=mean(duration),
            varDur=sd(duration),
            freqSt=n())

com_state_Ave_g1<-community_states_g1%>%
  group_by(pond, cycle,state)%>%
  summarize(meanDur=mean(duration),
            varDur=sd(duration),
            freqSt=n())
com_state_Ave_g3<-community_states_g3%>%
  group_by(pond, cycle,state)%>%
  summarize(meanDur=mean(duration),
            varDur=sd(duration),
            freqSt=n())

#and plot them
ggplot(subset(com_state_Ave, meanDur>4), aes(y=meanDur, x=cycle, group=pond,color=factor(pond)))+
  stat_summary(fund.data=mean_se)+theme_bw()+stat_smooth(se=F)

ggplot(com_state_Ave, aes(x=freqSt))+
  geom_histogram()+theme_bw()+
  facet_grid(vars(pond), vars(cycle))

ggplot(com_state_Ave, aes(x=freqSt, y=meanDur))+
  geom_point()+theme_bw()+
  stat_smooth(method=lm, se=FALSE)+
  facet_grid(vars(pond), vars(cycle))


# compare it to exponential distibution
lambda_df <- community_states %>%
  group_by(pond) %>%
  summarise(
    lambda = 1 / mean(duration),
    .groups = "drop"
  )

# Choose a max x-value (duration range)
max_duration <- max(community_states$duration)

# Create curve points for each pond
exp_curve <- lambda_df %>%
  mutate(data = map(lambda, ~ {
    tibble(
      duration = seq(0, max_duration, by = 0.1),
      density = dexp(seq(0, max_duration, by = 0.1), rate = .x)
    )
  })) %>%
  unnest(data)

ggplot() +
  geom_line(data = exp_curve, aes(x = duration, y = density),
            color = "darkgreen", size = 1.2, alpha=0.5) +
  geom_histogram(data = community_states, aes(x = duration, y = ..density..),
                 binwidth = 1, fill = "black") +
  facet_grid(~pond, scales = "free_y") +
  theme_bw()
```

lets calculate if duration and frequency of occurrence of a state are correlated
```{r}
corStat_dur_freq <- com_state_Ave %>%
  group_by(pond, cycle) %>%
  summarise(correlation = cor(meanDur, freqSt, method = "kendall",),
            obs=n(),
            .groups = "drop")

corStat_dur_freq_g1 <- com_state_Ave_g1 %>%
  group_by(pond, cycle) %>%
  summarise(correlation = cor(meanDur, freqSt, method = "kendall",),
            obs=n(),
            .groups = "drop")
corStat_dur_freq_g3 <- com_state_Ave_g3 %>%
  group_by(pond, cycle) %>%
  summarise(correlation = cor(meanDur, freqSt, method = "kendall",),
            obs=n(),
            .groups = "drop")

ggplot(corStat_dur_freq, aes(x=pond, y=correlation, color=cycle, size=obs))+
  geom_point()+
  scale_size(range = c(1, 5)) +
  stat_summary(fun.data=mean_se, show.legend = FALSE)+
  ylab("Correlation between duration and frequency of states")+
  theme_bw()

ggplot(corStat_dur_freq_g1, aes(x=pond, y=correlation, color=cycle, size=obs))+
  geom_point()+
  scale_size(range = c(1, 5)) +
  stat_summary(fun.data=mean_se, show.legend = FALSE)+
  ylab("Correlation between duration and frequency of states")+
  theme_bw()

ggplot(corStat_dur_freq_g3, aes(x=pond, y=correlation, color=cycle, size=obs))+
  geom_point()+
  scale_size(range = c(1, 5)) +
  stat_summary(fun.data=mean_se, show.legend = FALSE)+
  ylab("Correlation between duration and frequency of states")+
  theme_bw()



```


let's see how the number of unique community states varies
```{r}
unique_states<-community_states%>%
  group_by(pond, cycle)%>%
  summarize(No_states=length(unique(state)))
ggplot(unique_states, aes(x=cycle, y=No_states,color=pond, group=pond))+
  geom_point()+
  geom_smooth(se=F)+
  theme_bw()

unique_states_g1<-community_states_g1%>%
  group_by(pond, cycle)%>%
  summarize(No_states=length(unique(state)))
ggplot(unique_states_g1, aes(x=cycle, y=No_states,color=pond, group=pond))+
  geom_point()+
  geom_smooth(se=F)+
  theme_bw()

unique_states_g3<-community_states_g3%>%
  group_by(pond, cycle)%>%
  summarize(No_states=length(unique(state)))
ggplot(unique_states_g3, aes(x=cycle, y=No_states,color=pond, group=pond))+
  geom_point()+
  geom_smooth(se=F)+
  theme_bw()


```


#Network Structure Analysis
Finally, we can look at the whole structure of the transition network. We can do this as a temporal networks, or transform it into a static Markov-chain network
The second option essentially ways the transition between nodes by the frequency at which the transitions occur. Note this looses all temporal information.

##Tansition graphs
```{r}

#function to turn transition matrix into a markov transition matrix that gives frequency of transitions between states
trans_markov<-function(data){
  data%>%
  group_by(pond, cycle,starting_state,end_state)%>%
  summarize(weight=n())%>%
  ungroup()
}

trans_markov_matrix<-trans_markov(community_transitions)
trans_markov_matrix_g1<-trans_markov(community_transitions_g1)
trans_markov_matrix_g2<-trans_markov(community_transitions_g2)
trans_markov_matrix_g3<-trans_markov(community_transitions_g3)


# save results
write_csv(trans_markov_matrix, here("./output/frog_trans_markov.csv"))
write_csv(trans_markov_matrix_g1, here("./output/frog_trans_markov_g1.csv"))
write_csv(trans_markov_matrix_g2, here("./output/frog_trans_markov_g2.csv"))
write_csv(trans_markov_matrix_g3, here("./output/frog_trans_markov_g3.csv"))

#calculated weighed networks using iGraph package
weighted_networks <- function(data){
  grouped_data<-data%>%group_by(pond, cycle) %>%group_split()
  
  graphs<-lapply(grouped_data,function(group_data) {
    # Create edges from starting_state to end_state
    edges <- group_data %>%
      select(starting_state, end_state, weight)
    # Create the graph
    graph_from_data_frame(d = edges, directed = TRUE)
    })
  
  # Name each network in the list with "pond_cycle"
  names(graphs) <- data%>%
  group_by(pond, cycle) %>%
  group_keys() %>%
  mutate(name = paste0("pond_", pond, "_cycle_", cycle)) %>%
  pull(name)
  return(graphs)
  }

community_networks_weighted<- weighted_networks(trans_markov_matrix)
community_networks_weighted_g1<- weighted_networks(trans_markov_matrix_g1)
community_networks_weighted_g2<- weighted_networks(trans_markov_matrix_g2)
community_networks_weighted_g3<- weighted_networks(trans_markov_matrix_g3)

#Let's see how this looks ilke
plot(community_networks_weighted[["pond_1_cycle_2"]], layout=layout.circle, edge.width = E(community_networks_weighted[["pond_1_cycle_2"]])$weight*0.5,main = "Pond 1 Cycle 2 Network")
plot(community_networks_weighted[["pond_1_cycle_11"]], layout=layout.circle, edge.width = E(community_networks_weighted[["pond_1_cycle_11"]])$weight*0.5, main = "Pond 1 Cycle 11 Network")
plot(community_networks_weighted[["pond_1_cycle_22"]], layout=layout.circle, edge.width = E(community_networks_weighted[["pond_1_cycle_22"]])$weight*0.5, main = "Pond 1 Cycle 22 Network")


plot(community_networks_weighted[["pond_3_cycle_7"]], layout=layout.circle, edge.width = E(community_networks_weighted[["pond_3_cycle_7"]])$weight*0.5,main = "Pond 3 Cycle 7 Network")
plot(community_networks_weighted[["pond_3_cycle_11"]], layout=layout.circle, edge.width = E(community_networks_weighted[["pond_3_cycle_11"]])$weight*0.5,main = "Pond 3 Cycle 11 Network")


plot(community_networks_weighted[["pond_6_cycle_7"]], layout=layout.circle, edge.width = E(community_networks_weighted[["pond_6_cycle_7"]])$weight*0.5,main = "Pond 6 Cycle 7 Network")
plot(community_networks_weighted[["pond_6_cycle_11"]], layout=layout.circle, edge.width = E(community_networks_weighted[["pond_6_cycle_11"]])$weight*0.5,main = "Pond 6 Cycle 11 Network")

plot(community_networks_weighted[["pond_5_cycle_7"]], layout=layout.circle, edge.width = E(community_networks_weighted[["pond_5_cycle_7"]])$weight*0.5,main = "Pond 5 Cycle 7 Network")
plot(community_networks_weighted[["pond_5_cycle_11"]], layout=layout.circle, edge.width = E(community_networks_weighted[["pond_5_cycle_11"]])$weight*0.5,main = "Pond 5 Cycle 11 Network")


#save data
saveRDS(community_networks_weighted,file=here("./output/frog_community_networks_weighted.rds"))
saveRDS(community_networks_weighted_g1,file=here("./output/frog_community_networks_weighted_g1.rds"))
saveRDS(community_networks_weighted_g2,file=here("./output/frog_community_networks_weighted_g2.rds"))
saveRDS(community_networks_weighted_g3,file=here("./output/frog_community_networks_weighted_g3.rds"))
```
##whole network statistics on weighted static graphs
We can now calculate a number of statistics on basic "static" weighted transition graphs, with or without weights using the basic igraph package, like betweeness(), closseness(), strength(sums up edge weights), degree(), centrality (e.g. harmonic_centrality())

##Transition-Network metrics 
This is the ratio of actual links vs. possible links. So we would assume that all states can be reached from all states. But, in our case we can have multiple edges occur multiple times. So we can either ignore those, or think about how to incorporate them 
```{r}
#load data
community_networks_weighted<-readRDS(here("./output/frog_community_networks_weighted.rds"))
community_networks_weighted_g1<-readRDS(here("./output/frog_community_networks_weighted_g1.rds"))
community_networks_weighted_g2<-readRDS(here("./output/frog_community_networks_weighted_g2.rds"))
community_networks_weighted_g3<-readRDS(here("./output/frog_community_networks_weighted_g3.rds"))

#This function calculates various network metrics for a list of igraph objects
analyze_network_metrics <- function(network_list, join_data = NULL) {
  # Get network count
  network_count <- length(network_list)
  
  # Initialize results dataframe
  results <- data.frame(
    Pond_Cycle = character(network_count),
    Connectance = numeric(network_count),
    Unique_Nodes = numeric(network_count),
    totalEdges = numeric(network_count),
    modularity = numeric(network_count),
    mean_Distance = numeric(network_count),
    harm.centrality = numeric(network_count),
    betweenness = numeric(network_count),
    stringsAsFactors = FALSE
  )
  
  # Loop through and fill the dataframe by index
  for(i in seq_len(network_count)) {
    name <- names(network_list)[i]
    graph <- network_list[[name]]
    
    # Skip if there's any problem with the graph
    if(!is(graph, "igraph")) {
      results[i, ] <- NA
      next
    }
    
     # Check for trivial or empty graphs
    if (gorder(graph) < 2 || gsize(graph) == 0) {
      warning("Graph too small or empty: ", name)
      results[i, ] <- NA
      next
    }
    
    # Fill in the values for this row
    results$Pond_Cycle[i] <- name
    results$Unique_Nodes[i] <- gorder(graph)
    results$totalEdges[i] <- gsize(graph)
    
    # Calculate connectance (directed graph formula)
    results$Connectance[i] <- results$totalEdges[i] / 
      (results$Unique_Nodes[i] * (results$Unique_Nodes[i] - 1))
    
    # Calculate modularity using walktrap algorithm
    results$modularity[i] <- modularity(cluster_walktrap(graph))
    
    # Calculate mean distance
    results$mean_Distance[i] <- mean_distance(graph)
    
    # Calculate centralization measures
    results$harm.centrality[i] <- mean(harmonic_centrality(graph, mode = "all",  normalized = TRUE)) #harmonic sets disconnect =0 not Inf
    results$betweenness[i] <- centr_betw(graph)$centralization
  }
  
  # Extract pond and cycle information from names
  results <- results %>%
    mutate(
      pond = as.factor(sub("pond_([^_]+)_cycle_.*", "\\1", Pond_Cycle)),
      cycle = as.numeric(sub(".*_cycle_([^_]+)", "\\1", Pond_Cycle))
    )
  
  # Join with additional data if provided
  if(!is.null(join_data)) {
    results <- inner_join(results, join_data, by = intersect(names(results), names(join_data)))
  }
  
  # Remove any rows with NA values (from skipped networks)
  #results <- results%>%filter(!is.na)
  results <- results %>%
    mutate(across(everything(), ~replace_na(., NA)))
  return(results)
}

sp_num<-read_csv(here("./output/frog_pond_cycle_NoSpec.csv"))  %>%mutate(pond=as.factor(pond), cycle=as.numeric(cycle))

state_connectance_w <- analyze_network_metrics(community_networks_weighted, sp_num)

state_connectance_w_g1 <- analyze_network_metrics(community_networks_weighted_g1, sp_num)

state_connectance_w_g2 <- analyze_network_metrics(community_networks_weighted_g2, sp_num)

state_connectance_w_g3 <- analyze_network_metrics(community_networks_weighted_g3, sp_num)
 

```
Let's look at some patterns again
```{r}
ggplot(state_connectance_w, aes(x=cycle, y=betweenness, group=pond,color=pond))+
  geom_point()+
  geom_line()+
  theme_bw()
ggplot(state_connectance_w, aes(x=cycle, y=Unique_Nodes, group=pond,color=pond))+
  geom_point()+
  geom_line()+
  theme_bw()

ggplot(state_connectance_w, aes(y=log(Connectance), x=Unique_Nodes, group=pond,color=pond))+
  geom_point()+
  stat_smooth(se=F)+
  theme_bw()


#calculate all possible unique states for a given number of species
n_species=3:12 #possible  species pool, 12 is max
max_size=9 #maximum number of co-occuring species we ever observed

total_combinations <- sapply(n_species, function(n) {
    max_k <- min(n, max_size)
    sum(choose(n, 2:n))
    }
    )

maxSpecComb <- data.frame(
  SpeciesPool = n_species,
  Unique_Nodes = total_combinations
)

ggplot(state_connectance_w,aes(x=No_Spec, y=log(Unique_Nodes)))+
  stat_smooth(se=F)+
  geom_point(aes(color=pond))+
  geom_smooth(data=state_connectance_w_g1,aes(x=No_Spec, y=log(Unique_Nodes)),se=F, color="black")+
  geom_smooth(data=state_connectance_w_g3,aes(x=No_Spec, y=log(Unique_Nodes)), se=F,color="green")+
  geom_line(data=maxSpecComb, aes(x=SpeciesPool, y=log(Unique_Nodes)),color="darkgrey", linetype=2)+
  scale_x_continuous(breaks = seq(0, 12, by = 1))+
  theme_classic()

ggplot(state_connectance_w,aes(x=cycle, y=(Connectance)))+
  stat_smooth(se=F)+
  geom_point(aes(color=cycle))+
  scale_x_continuous(breaks = seq(0, 22, by = 2))+
  facet_wrap(~pond)+
  theme_bw()

```


#Temporal network 
We can also transform this data into a temporal network using the ndtv package 
##create temporal network
```{r temporal network,results='hide', message=FALSE, warning=FALSE}
# Function to create temporal networks from a single dataset
temporal_networks <- function(transitions_data, states_data = NULL) {
  
  # Helper function to create a temporal network for a specific pond and cycle
  create_temporal_network <- function(transitions, states, pond_id, cycle_id) {
    # Filter data for the specific pond and cycle
    sub_transitions <- transitions %>% 
      filter(pond == pond_id, cycle == cycle_id)
    
    sub_states <- states %>% 
      filter(pond == pond_id, cycle == cycle_id)
    
    # Get unique states (including "none")
    unique_states <- as.character(unique(sub_states$state))
    
    # Create vertex spells dataframe
    vertex_spells <- data.frame(
      vertex.id = integer(),
      onset = numeric(),
      terminus = numeric(),
      stringsAsFactors = FALSE
    )
    
    for (i in 1:nrow(sub_states)) {
      state <- sub_states$state[i]
      
      # Get node ID
      vertex_id <- which(unique_states == state)
      
      if (length(vertex_id) > 0) {
        spell_row <- data.frame(
          vertex.id = vertex_id,
          onset = sub_states$onset[i],
          terminus = sub_states$terminus[i]
        )
        vertex_spells <- rbind(vertex_spells, spell_row)
      }
    }
    
    # Create edge spells dataframe
    edge_spells <- data.frame(
      tail = integer(),
      head = integer(),
      onset = numeric(),
      terminus = numeric(),
      stringsAsFactors = FALSE
    )
    
    for (i in 1:nrow(sub_transitions)) {
      start <- sub_transitions$starting_state[i]
      end <- sub_transitions$end_state[i]
      
      # Find node IDs
      tail_id <- which(unique_states == start)
      head_id <- which(unique_states == end)
      
      if (length(tail_id) > 0 && length(head_id) > 0) {
        edge_row <- data.frame(
          tail = tail_id,
          head = head_id,
          onset = sub_transitions$day[i],
          terminus = sub_transitions$day[i] + 1
        )
        edge_spells <- rbind(edge_spells, edge_row)
      }
    }
    
    # Check if we have any valid data
    if (nrow(vertex_spells) == 0) {
      warning(paste("No valid vertex spells for pond", pond_id, "cycle", cycle_id))
      return(NULL)
    }
    
    # Create the dynamic network
    if (nrow(edge_spells) == 0) {
      # Create network with only vertices if no edges
      warning(paste("No valid edge spells for pond", pond_id, "cycle", cycle_id))
      dyn_net <- networkDynamic(
        vertex.spells = vertex_spells[c("onset", "terminus", "vertex.id")]
      )
    } else {
      # Create network with both vertices and edges
      dyn_net <- networkDynamic(
        vertex.spells = vertex_spells[c("onset", "terminus", "vertex.id")],
        edge.spells = edge_spells[c("onset", "terminus", "tail", "head")]
      )
    }
    
    # Store node name mapping
    node_mapping <- setNames(seq_along(unique_states), unique_states)
    
    return(list(network = dyn_net, node_mapping = node_mapping))
  }
  
  # Use provided states data or assume community_states exists in environment
  if (is.null(states_data)) {
    if (exists("community_states")) {
      states_data <- community_states
    } else {
      stop("states_data must be provided or 'community_states' must exist in the environment")
    }
  }
  
  # Get unique pond-cycle combinations
  pond_cycles <- unique(transitions_data[, c("pond", "cycle")])
  
  # Create networks for each pond-cycle combination
  temporal_networks_list <- list()
  
  for (i in 1:nrow(pond_cycles)) {
    pond_id <- pond_cycles$pond[i]
    cycle_id <- pond_cycles$cycle[i]
    
    network_name <- paste0("pond_", pond_id, "_cycle_", cycle_id)
    
    result <- create_temporal_network(transitions_data, states_data, pond_id, cycle_id)
    
    if (!is.null(result)) {
      temporal_networks_list[[network_name]] <- result
    }
    
    cat("Created network for", network_name, "\n")
  }
  
  return(temporal_networks_list)
}

#get data
community_transitions<-read_csv(here("./output/frog_transit_states.csv"))
community_transitions_g1<-read_csv(here("./output/frog_transit_states_g1.csv"))
community_transitions_g2<-read_csv(here("./output/frog_transit_states_g2.csv"))
community_transitions_g3<-read_csv(here("./output/frog_transit_states_g3.csv"))

#run function with data
temporal_trans_networks <- temporal_networks(community_transitions)
temporal_trans_networks_g1 <- temporal_networks(community_transitions_g1)
temporal_trans_networks_g2 <- temporal_networks(community_transitions_g2)
temporal_trans_networks_g3 <- temporal_networks(community_transitions_g3)


#save networks
saveRDS(temporal_trans_networks,here("./output/frog_temp_transit_network.rds"))
saveRDS(temporal_trans_networks_g1,here("./output/frog_temp_transit_network_g1.rds"))
saveRDS(temporal_trans_networks_g2,here("./output/frog_temp_transit_network_g2.rds"))
saveRDS(temporal_trans_networks_g3,here("./output/frog_temp_transit_network_g3.rds"))

```

## analyze temporal structure
```{r}
#get reachable nodes
# Extract the network and node mapping
test_p1_c1 <- temporal_trans_networks[["pond_1_cycle_1"]]
dyn_net <- test_p1_c1$network
node_mapping <- test_p1_c1$node_mapping  # Ensure node mapping is retrieved

# Get the vertex ids (start vertices)
vertex_ids <- seq_along(node_mapping)

# Initialize a list to store shortest paths
shortest_paths_list <- list()

# Loop over each vertex and compute the shortest forward paths
for (v in vertex_ids) {
  # Compute shortest path from vertex v to all others
  path_matrix <- tPath(dyn_net, v = v)  # Specify the starting vertex
     path_lengths <- path_matrix$tdist
    
    # Store the result in the list
    shortest_paths_list[[names(node_mapping)[v]]] <- setNames(path_lengths, names(node_mapping))
  }
print(shortest_paths_list) #issue, we assume that any node can be connected via the none state, but doesn't always make sense, so we can remove the none state to see if certain community states can be reached at all. 


# Compute temporal reachability matrix
reach_matrix <- tReach(dyn_net)

# Ensure reach_matrix is a proper matrix
if (is.null(dim(reach_matrix))) {  
  # Handle case where tReach() returns a vector
  reachable_nodes_list <- list(names(node_mapping)[which(reach_matrix > 0)])
  names(reachable_nodes_list) <- names(node_mapping)[1]  # Assign first node name
} else {
  # Handle case where tReach() returns a matrix
  reachable_nodes_list <- apply(reach_matrix, 1, function(x) {
    as.character(names(node_mapping)[which(x > 0)])  # Ensure character format
  })
  names(reachable_nodes_list) <- names(node_mapping)
}

# Print the corrected list of reachable nodes
print(reachable_nodes_list)
```
##Burstiness
We will calcluate temporal distribution of events, in this case when a new state appears. This gives us a good idea of when and how changes are clumped in time or not. Values=1 indicate perfect (periodic) spacing, = 0 random, >0 clumping. Note, we need to decide if we include empty states in this calculation
```{r}
#get data, we'll just use basic community state data for simplicity
community_states<-read_csv(here("./output/frog_CommunityStates.csv"))
community_statesNonone<-read_csv(here("./output/frog_CommunityStates.csv"))%>%subset(state!="none")
community_states_g1<-read_csv(here("./output/frog_CommunityStates_g1.csv"))
community_states_g2<-read_csv(here("./output/frog_CommunityStates_g2.csv"))
community_states_g3<-read_csv(here("./output/frog_CommunityStates_g3.csv"))

# Function to compute burstiness coefficient B
calculate_burstiness <- function(df) {
  df %>%
    group_by(pond, cycle) %>%  # Group by pond and cycle
    arrange(onset) %>%  # Ensure events are ordered by time
    summarise(
      mean_t = mean(diff(onset), na.rm = TRUE),
      sd_t = sd(diff(onset), na.rm = TRUE),
      B = ifelse(sd_t + mean_t > 0, (sd_t - mean_t) / (sd_t + mean_t), NA_real_),  # Burstiness formula
      .groups = "drop"
    )
}
# Example usage
burstiness_results<-calculate_burstiness(community_states)
burstiness_resultsNonone<-calculate_burstiness(community_statesNonone)
burstiness_results_g1<-calculate_burstiness(community_states_g1)
burstiness_results_g2<-calculate_burstiness(community_states_g2)
burstiness_results_g3<-calculate_burstiness(community_states_g3)




#let's look at it
ggplot(burstiness_results, aes(x=pond,y=B, color=cycle))+
  stat_summary(fun.data=mean_se, size=1.5)+
  geom_point(alpha=0.5)+
  labs(x="Pond", y="Burstiness, B")+
  theme_bw()

ggplot(burstiness_results, aes(x=cycle,y=B, color=pond, group=as.factor(pond)))+
  geom_point(alpha=0.5)+
  stat_smooth(se=F)+
  labs(x="Cycle", y="Burstiness, B (excluding empty states)")+
  theme_bw()


#let's graph both so we can see if including none state alters general patter (it doesnt', just shifts curves down a bit)
ggplot(burstiness_results, aes(x=cycle,y=B, color=as.factor(pond)))+
  stat_smooth(data=burstiness_results, aes(x=cycle,y=B, color=as.factor(pond)), linetype=1, se=F)+
  geom_point(alpha=0.5)+
  stat_smooth(data=burstiness_resultsNonone, aes(x=cycle,y=B, color=as.factor(pond)), linetype=2, se=F)+
  stat_smooth(data=burstiness_results_g1, aes(x=cycle,y=B),color="blue",  se=F)+
  stat_smooth(data=burstiness_results_g3, aes(x=cycle,y=B),color="darkgrey",  se=F)+
  labs(x="Pond", y="Burstiness, B")+
  facet_wrap(~pond)+
  theme_bw()

#lets see if burstiness is correlated over time across ponds
library(ggcorrplot)
wide_data <- burstiness_results %>%select(pond,B,cycle)%>%
  pivot_wider(names_from = pond, values_from = B)%>%
  arrange(cycle)%>%select(-cycle)
p.mat<-cor_pmat(wide_data)
cor.mat<-cor(wide_data)
ggcorrplot(cor.mat,type="lower",
           outline.color="white",
           ggtheme=ggplot2::theme_classic,colors = c("#6D9EC1", "white", "#E46726"),
           lab=T,
           p.mat=p.mat)


```

##Memory
```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(purrr)

# Calculate memory metrics for a given pond and cycle
calculate_memory <- function(data, pond_val, cycle_val) {
  # Sort by onset time
  data <- data %>% arrange(onset)
  
  # Calculate time gaps between consecutive states
  data <- data %>%
    mutate(
      prev_state = lag(state),
      prev_terminus = lag(terminus),
      time_gap = onset - lag(terminus),
      state_transition = paste(lag(state), "->", state)
    )
  
  # Calculate state frequencies
  state_counts <- table(data$state)
  state_freq <- state_counts / nrow(data)
  
  # Calculate autocorrelation for durations
  ac <- tryCatch({
    acf_result <- acf(data$duration, lag.max = min(10, nrow(data)-1), plot = FALSE)
    ac_values <- as.numeric(acf_result$acf)
    names(ac_values) <- paste0("ac_lag_", 0:(length(ac_values)-1))
    ac_values
  }, error = function(e) {
    ac_values <- rep(NA, 11)
    names(ac_values) <- paste0("ac_lag_", 0:10)
    ac_values
  })
  
  # Calculate memory metrics
  # 1. Markov memory: How well previous state predicts current state
  state_repeats <- sum(data$state == data$prev_state, na.rm = TRUE) / max(1, (nrow(data) - 1))
  
  # 2. Time memory: Correlation between consecutive durations
  time_memory <- tryCatch({
    if(nrow(data) > 1 && sd(data$duration, na.rm = TRUE) > 0) {
      durations1 <- data$duration[-length(data$duration)]
      durations2 <- data$duration[-1]
      if(sd(durations1, na.rm = TRUE) > 0 && sd(durations2, na.rm = TRUE) > 0) {
        cor(durations1, durations2, use = "complete.obs")
      } else {
        NA_real_
      }
    } else {
      NA_real_
    }
  }, error = function(e) {
    NA_real_
  })
  
  # 4. Mean return time (how long until a state recurs)
  return_times_df <- tryCatch({
    data %>%
      group_by(state) %>%
      mutate(
        next_occurrence = lead(onset),
        return_time = ifelse(!is.na(next_occurrence), next_occurrence - onset, NA)
      ) %>%
      summarise(mean_return_time = mean(return_time, na.rm = TRUE), .groups = "drop")
  }, error = function(e) {
    data.frame(state = character(), mean_return_time = numeric())
  })
  
  # Convert return times to a more compact representation
  return_times_str <- if(nrow(return_times_df) > 0) {
    paste(return_times_df$state, round(return_times_df$mean_return_time, 2), 
          sep = ":", collapse = "; ")
  } else {
    NA_character_
  }
  
  # Return as a single-row data frame with all metrics
  result_df <- data.frame(
    # Don't include pond and cycle here - they're already in the outer data frame
    markov_memory = state_repeats,
    time_memory = time_memory
    )
  return(result_df)
}

# Apply memory calculation to each pond and cycle combination
memory_analysis <- function(community_states) {
  results <- community_states %>%
    group_by(pond, cycle) %>%
    nest() %>%
    mutate(
      memory_metrics = pmap(
        list(data = data, pond_val = pond, cycle_val = cycle),
        function(data, pond_val, cycle_val) {
          calculate_memory(data, pond_val, cycle_val)
        }
      )
    )
  
  #Unnest the results
  results_df <- results %>%
    select(-data) %>%
    unnest(memory_metrics)
  
  return(results_df)
}

# Summarize memory metrics across ponds and cycles
summarize_memory <- function(memory_results) {
  summary_stats <- memory_results %>%
    summarise(across(where(is.numeric), 
                    list(mean = ~mean(.x, na.rm = TRUE),
                         sd = ~sd(.x, na.rm = TRUE),
                         min = ~min(.x, na.rm = TRUE),
                         max = ~max(.x, na.rm = TRUE)),
                    .names = "{.col}_{.fn}"))
  
  return(summary_stats)
}

memory_results <- memory_analysis(community_states)
memory_results_g1 <- memory_analysis(community_states_g1)
memory_results_g2 <- memory_analysis(community_states_g2)
memory_results_g3 <- memory_analysis(community_states_g3)

burst_mem<-left_join(burstiness_results,memory_results)
burst_mem_g1<-left_join(burstiness_results_g1,memory_results_g1)
burst_mem_g2<-left_join(burstiness_results_g2,memory_results_g2)
burst_mem_g3<-left_join(burstiness_results_g3,memory_results_g3)

write_csv(burst_mem,here("./output/frog_comm_state_burstmem.csv"))
write_csv(burst_mem_g1,here("./output/frog_comm_state_burstmem_g1.csv"))
write_csv(burst_mem_g2,here("./output/frog_comm_state_burstmem_g2.csv"))
write_csv(burst_mem_g3,here("./output/frog_comm_state_burstmem_g3.csv"))

#combine with other network metrics
Astate_connectance_w <- left_join(
  state_connectance_w %>% mutate(pond = as.character(pond), cycle = as.character(cycle)),
  burst_mem %>% mutate(pond = as.character(pond), cycle = as.character(cycle)),
  by = c("pond", "cycle"))
Astate_connectance_w_g1 <- left_join(
  state_connectance_w_g1 %>% mutate(pond = as.character(pond), cycle = as.character(cycle)),
  burst_mem_g1 %>% mutate(pond = as.character(pond), cycle = as.character(cycle)),
  by = c("pond", "cycle"))
Astate_connectance_w_g2 <- left_join(
  state_connectance_w_g2 %>% mutate(pond = as.character(pond), cycle = as.character(cycle)),
  burst_mem_g2 %>% mutate(pond = as.character(pond), cycle = as.character(cycle)),
  by = c("pond", "cycle"))
Astate_connectance_w_g3 <- left_join(
  state_connectance_w_g3 %>% mutate(pond = as.character(pond), cycle = as.character(cycle)),
  burst_mem_g3 %>% mutate(pond = as.character(pond), cycle = as.character(cycle)),
  by = c("pond", "cycle"))

write_csv(Astate_connectance_w, here("./output/frog_transition_matrix_metrics.csv"))
write_csv(Astate_connectance_w_g1, here("./output/frog_transition_matrix_metrics_g1.csv"))
write_csv(Astate_connectance_w_g2, here("./output/frog_transition_matrix_metrics_g2.csv"))
write_csv(Astate_connectance_w_g3, here("./output/frog_transition_matrix_metrics_g3.csv"))


ggplot(memory_results, aes(x=pond, y=time_memory, color=cycle))+
  stat_summary(fun.data=mean_se, size=1.5, alpha=0.5)+
  geom_point()+
  theme_bw()

ggplot(memory_results, aes(x=cycle, y=time_memory, group=pond,color=pond))+
  geom_point()+
  geom_line(se=F)+
  geom_line(data=memory_results_g1, aes(x=cycle, y=time_memory, group=pond,color=pond),color="black", se=F)+
  geom_line(data=memory_results_g2, aes(x=cycle, y=time_memory, group=pond,color=pond),color="darkgrey",linetype=2, se=F)+
  geom_line(data=memory_results_g3, aes(x=cycle, y=time_memory, group=pond,color=pond),color="lightgrey", se=F)+
  theme_bw()+
  facet_wrap(~pond)

#test for differences
modelM<-lm(time_memory~(cycle)*as.factor(pond), data=memory_results)
library(car)
Anova(modelM)
#so differences in memory across ponds, but no consistent temporal effect. Higher order polynomials didn't change that
#no time effect
```
let's add the burstiness and memory data to the rest of the network data.


##Edge diversity  
Transition between certain state pairs can occur more common, indicating a back and forth, while other transitions may be rare. We can think of this in term of either distributions of edges, weights, or evenness. If all transitions occur equally, then eveness of edges should be high, if some transitions are dominant and occur much more frequently, than eveness is low. So in simplest case, we can calcuclate this with Pielou's eveness index.
```{r}
#load edge frequency matrix
trans_freq<-read_csv(here("./output/frog_trans_markov.csv"))
trans_freq_g1<-read_csv(here("./output/frog_trans_markov_g1.csv"))
trans_freq_g2<-read_csv(here("./output/frog_trans_markov_g2.csv"))
trans_freq_g3<-read_csv(here("./output/frog_trans_markov_g3.csv"))

#calculate eveness of transitions
calc_evenness<-function(df){
  df%>%
    group_by(pond, cycle)%>%
    summarize(
      total_weight=sum(weight),
      S=n_distinct(paste(starting_state,end_state,sep="->")), #unique state transitions
      H=-sum((weight/total_weight)*log(weight/total_weight), na.rm=TRUE), #shannon's diversity index
      J=H/log(S),
      .groups="drop"
    )
}
Trans_evenness <- calc_evenness(trans_freq)%>%mutate(pond = as.character(pond), cycle = as.character(cycle))
print(Trans_evenness)

Trans_evenness_g1 <- calc_evenness(trans_freq_g1)%>%mutate(pond = as.character(pond), cycle = as.character(cycle))
Trans_evenness_g2 <- calc_evenness(trans_freq_g2)%>%mutate(pond = as.character(pond), cycle = as.character(cycle))
Trans_evenness_g3 <- calc_evenness(trans_freq_g3)%>%mutate(pond = as.character(pond), cycle = as.character(cycle))


Tansition_metrics_w <- left_join(Astate_connectance_w, Trans_evenness)
Tansition_metrics_w_g1 <- left_join(Astate_connectance_w_g1, Trans_evenness_g1)
Tansition_metrics_w_g2 <- left_join(Astate_connectance_w_g2, Trans_evenness_g2)
Tansition_metrics_w_g3 <- left_join(Astate_connectance_w_g3, Trans_evenness_g3)

write_csv(Tansition_metrics_w, here("./output/frog_transition_matrix_metrics.csv"))
write_csv(Tansition_metrics_w_g1, here("./output/frog_transition_matrix_metrics_g1.csv"))
write_csv(Tansition_metrics_w_g2, here("./output/frog_transition_matrix_metrics_g2.csv"))
write_csv(Tansition_metrics_w_g3, here("./output/frog_transition_matrix_metrics_g3.csv"))

ggplot(Trans_evenness, aes(x=pond, y=J, color=cycle))+
  stat_summary(fun.data=mean_se, size=1.5, alpha=0.5)+
  geom_point()+
  theme_bw()

```



use tna package to look at sequential/transition networks
```{r}
library("tna")
```
```{r}
#we can analyze directed weighted newtorks with tna package which is set up to graph and analyze directed marcov chain networks. So frequenc here would be seen as probabilities, which is the same as weights in igraph. It has a lot of easy to use build in metrics and nice plot options that can make our lives easier. Ideally this shoudl start with data that's showing each transition, but no other column (like days).  un our case, that's community_transitions data set

community_transitions<-read_csv("output/frog_transit_states.csv")

#convert into tna format, note that requires that we only have two columns of head and tail of transitions

#quick example for single graph got one pond_cycle combo
testTNA<-community_transitions%>%subset(pond==4 & cycle==10)%>%select(starting_state, end_state)
#convert to tna format
tna_T<-tna(testTNA)
options(scipen = 0, digits=2)
summary(tna_T)
plot(tna_T)
cent<-centralities(tna_T)
plot(cent)


communities <- communities(tna_T)
plot(communities, method="fast_greedy", pieBorder=FALSE)
```