---
title: "Historical Climate analysis"
author: "Volker Rudolf"
date: "2026-02-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Goals
We will analyze historical climate (~100years) to determine extreme years.


## Load packages & data
```{r, warning=FALSE, message=FALSE}
 #Load packages
  #list of packages that are needed
  pack<-c("tidyverse",
          "knitr",
          "here",
          "lubridate",
          "paletteer",
          "ggpubr",
          "patchwork",
          "factoextra"
          )
  #check whether packages are installed and load them, if not get and then load
 package.check <- lapply(pack, FUN = function(x){
    if (!require(x, character.only = TRUE)) 
      {
        install.packages(x, dependencies = TRUE)
        library(x, character.only = TRUE)
      }
    })
   #search() #verify that packages are loaded
```

#load Temperature and precipitation data
Note the full data set goes back to 1906, but early records are incomplete.

```{r}
#get and organize raw daily climate observations from weather station
hclimate<-read_csv(here("data","LufkinClimatedata1906_2025.csv"))%>%select(DATE,PRCP,TMAX,TMIN)
hclimate$DATE<-as.Date(hclimate$DATE,format="%m/%d/%Y")
hclimate<-hclimate%>%mutate(year = year(DATE), month=month(DATE), day=day(DATE),doy=as.numeric(format(DATE, "%j")))

#now this is usually on annual basis, but we need to align it with the "frog cycle" we used for the community analysis
hclimate<-hclimate%>%subset(year<2025)%>%
  mutate(cycle=case_when(
  doy>339~year+1,
  .default=year
))
hclimate<-hclimate%>%mutate(cycle_day=case_when(
  doy>=340~doy-339,
  .default=doy+27
))

sumclimaHY<-hclimate%>%group_by(cycle)%>%
  summarize(meanMaxT=mean(TMAX,na.rm = TRUE), 
            meanMinT=mean(TMIN,na.rm = TRUE),
            hotdays=sum(TMAX > 35,na.rm = TRUE),
            meanPrec=mean(PRCP,na.rm=TRUE),
            noPrec=sum(PRCP==0,na.rm=TRUE),
            
            recDays=length(DATE)
            )

#check how many data points there are per year
ggplot(sumclimaHY, aes(x=cycle, y=recDays))+geom_point()+theme_bw()

incomplete_years <- sumclimaHY %>%
  filter(recDays < 364)%>%pull(cycle)
print(incomplete_years)

#so we see there are several 9 years with incomplete data including one with 306 and one with 335 days.  

c_sumclimaHY<-sumclimaHY%>%filter(!(cycle %in% incomplete_years))
```
so we need to exclude years that don't have enough entries, that includes first year

# Extreme climate years
Now we will look at temperature and rainfall to identify extreme years. We will look at average Max Day temperature (meanMaxT), and number of hotdays above 100F, and average daily precipitation (meanPrec), and days without (0 prec) rain (noPrec). We will get average and 1 standard deviation across the 50 years and then look at potential outlier outside of that range.
```{r}
cs_sumclimaHY<-c_sumclimaHY%>%subset(cycle>1972)%>%
  mutate(
    mMaxTY=mean(meanMaxT),
    mMinTY=mean(meanMinT),
    mPrecY=mean(meanPrec),
    mhotdaysY=mean(hotdays),
    mNoPrec=mean(noPrec),
    eMaxTY=mean(meanMaxT)+sd(meanMaxT),
    eMinTY=mean(meanMinT)-sd(meanMinT),
    ePrecY=mean(meanPrec)-sd(meanPrec),
    eNoPrec=mean(mNoPrec)+sd(noPrec),
    ehotdaysY=mean(hotdays)+sd(hotdays),
    )%>%
  mutate(exthighTemp=ifelse(meanMaxT>eMaxTY, "Yes", "No"),
         extlowTemp=ifelse(meanMinT<eMinTY, "Yes", "No"),
         extlongHeat=ifelse(hotdays>ehotdaysY, "Yes", "No"),
         extDry=ifelse(meanPrec<ePrecY, "Yes", "No"),
         extDryDays=ifelse(noPrec>eNoPrec, "Yes", "No"),
         extHot=ifelse(exthighTemp=="Yes" | extlongHeat=="Yes", "Yes", "No"),
         extDryT=ifelse(extDry=="Yes" | extDryDays=="Yes", "Yes", "No"))


#make plots (Insert of Figure 5)
HETmaxplot<-ggplot(subset(cs_sumclimaHY),aes(x=cycle,y=meanMaxT))+
  annotate("rect", xmin = 2000.5, xmax = 2025, ymin = -Inf, ymax = Inf, 
           color="black",linewidth=0.5,fill = NA, linetype="solid") +
  geom_ribbon(aes(ymin = mMaxTY-sd(meanMaxT), ymax = eMaxTY), alpha = 0.3, fill = "grey") +
  geom_segment(aes(x = 1974, xend = 2024, y = mMaxTY, yend = mMaxTY), 
             linetype = "dashed", color = "black")+
  geom_line()+
  geom_point(aes(color=ifelse(cycle>2000,exthighTemp,"No")), size=3)+
  annotate("rect", xmin = 1973, xmax = 2000.5, ymin = -Inf, ymax = Inf, 
           fill="grey",alpha=0.35) +
  labs(y="Daily high temperature (\u00b0C)", x="Annual Cycle")+
  scale_color_manual(name = "Extreme event",
    values = c("No" = "black", "Yes" = "red"))+
  scale_x_continuous(
    limits = c(1973, NA), 
    breaks = seq(from = 1974, to = max(cs_sumclimaHY$cycle), by = 5))+
  theme_classic()+
  theme(legend.position = c(0.15, 0.9),
        axis.text.x = element_text(size = 10), 
        axis.text.y = element_text(size = 10), 
        axis.title = element_text(size = 14),)+theme(legend.position = "none")

HEhotdayplot<-ggplot(subset(cs_sumclimaHY),aes(x=cycle,y=hotdays))+
   annotate("rect", xmin = 2000.5, xmax = 2025, ymin = -Inf, ymax = Inf, 
           color="black",linewidth=0.5,fill = NA, linetype="solid") +
  geom_ribbon(aes(ymin = mhotdaysY-sd(hotdays), ymax = ehotdaysY), alpha = 0.3, fill = "grey") +
  geom_segment(aes(x = 1974, xend = 2024, y = mhotdaysY, yend = mhotdaysY), 
             linetype = "dashed", color = "black")+
  geom_line()+
  geom_point(aes(color=ifelse(cycle>2000,extlongHeat,"No")), size=3)+
  annotate("rect", xmin = 1973, xmax = 2000.5, ymin = -Inf, ymax = Inf, 
           fill="grey",alpha=0.3) +labs(y="Daily high temperature (\u00b0C)", x="Annual Cycle")+
  labs(y="Total hot days (>35\u00b0C)", x="Annual Cycle")+
  scale_color_manual(name = "Extreme event",
    values = c("No" = "black", "Yes" = "red"))+
  scale_x_continuous(
    limits = c(1973, NA), 
    breaks = seq(from = 1974, to = max(cs_sumclimaHY$cycle), by = 5))+
  theme_classic()+
  theme(legend.position = "none")+
  theme(legend.position = c(0.15, 0.9),
        axis.text.x = element_text(size = 10), 
        axis.text.y = element_text(size = 10), 
        axis.title = element_text(size = 14))
  
HEprecplot<-ggplot(subset(cs_sumclimaHY),aes(x=cycle,y=meanPrec))+
  annotate("rect", xmin = 2000.5, xmax = 2025, ymin = -Inf, ymax = Inf, 
           color="black",linewidth=0.5,fill = NA, linetype="solid") +
  geom_ribbon(aes(ymax = mPrecY+sd(meanPrec), ymin = ePrecY), alpha = 0.3, fill = "grey") +
  geom_segment(aes(x = 1974, xend = 2024, y = mPrecY, yend = mPrecY), 
             linetype = "dashed", color = "black")+
  geom_line()+
  geom_point(aes(color=ifelse(cycle>2000,extDry, "No")), size=3)+
  annotate("rect", xmin = 1973, xmax = 2000.5, ymin = -Inf, ymax = Inf, 
           fill="grey",alpha=0.3) +labs(y="Daily high temperature (\u00b0C)", x="Annual Cycle")+
  labs(y="Precipitation (mm)", x="Annual Cycle")+
  scale_color_manual(name = "Extreme event",
    values = c("No" = "black", "Yes" = "red"))+
  scale_x_continuous(
    limits = c(1973, NA), 
    breaks = seq(from = 1974, to = max(cs_sumclimaHY$cycle), by = 5))+
  theme_classic()+
  theme(legend.position = "none")+
  theme(legend.position = c(0.15, 0.9),
        axis.text.x = element_text(size = 10), 
        axis.text.y = element_text(size = 10), 
        axis.title = element_text(size = 14))
  
HEnorainplot<-ggplot(subset(cs_sumclimaHY),aes(x=cycle,y=noPrec))+
  annotate("rect", xmin = 2000.5, xmax = 2025, ymin = -Inf, ymax = Inf, 
           color="black",linewidth=0.5,fill = NA, linetype="solid") +
  geom_ribbon(aes(ymin = mNoPrec-sd(noPrec), ymax = mNoPrec+sd(noPrec)), alpha = 0.3, fill = "grey")+
  geom_segment(aes(x = 1974, xend = 2024, y = mNoPrec, yend = mNoPrec), 
             linetype = "dashed", color = "black")+
  geom_line()+
  geom_point(aes(color=ifelse(cycle>2000,extDryDays,"No")), size=3)+
  annotate("rect", xmin = 1973, xmax = 2000.5, ymin = -Inf, ymax = Inf, 
           fill="grey",alpha=0.3) +labs(y="Daily high temperature (\u00b0C)", x="Annual Cycle")+
  labs(y="Total days without rain", x="Annual Cycle")+
  scale_color_manual(name = "Extreme event",
    values = c("No" = "black", "Yes" = "red"))+
  scale_x_continuous(
    limits = c(1973, NA), 
    breaks = seq(from = 1974, to = max(cs_sumclimaHY$cycle), by = 5))+
  theme_classic()+
  theme(legend.position = "none")+
  theme(legend.position = c(0.15, 0.9),
        axis.text.x = element_text(size = 10), 
        axis.text.y = element_text(size = 10), 
        axis.title = element_text(size = 14))
  
Hcombined_plot <- (HETmaxplot + HEhotdayplot) / (HEprecplot + HEnorainplot)+
  plot_layout(guides="collect")& 
  theme(
    legend.position = "none",
    #axis.text = element_text(size = 8), 
    #axis.title = element_text(size = 9, face="bold")
  )+theme(legend.position="top")

Hcombined_plot
ggsave(here("Figures","extremeYears50.png"),plot=Hcombined_plot,dpi=600,width = 30, height = 15, units = "cm")
```
now let's save the extreme year info to combine it with the other climate data
```{r}
#first need to align cycle same way as previously
cs_sumclimaHY.r<-cs_sumclimaHY%>%
  select(cycle,exthighTemp,extlongHeat,extDry,extDryDays,extHot,extDry,extDryT)%>%
  subset(cycle>2000)%>%
  mutate(year=cycle,cycle=cycle-2000)

write.csv(cs_sumclimaHY.r, here("output","extremeyearsHist.csv"), row.names = FALSE)
```

