---
title: "NetworkPlots"
author: "Volker Rudolf"
date: "2025-12-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
#Goals
The goal of this analysis is to turn species phenologies into a network of community state transitions. When species co-occur, they form a community state, and each species combination is a unique state, which can occur more than once. Each state has a start and ending date, and transitions into a new state. This creates a network that we will analyze to better understand how communities change

### Load packages
```{r, warning=FALSE, message=FALSE}
 #Load packages
  #list of packages that are needed
  pack<-c("tidyverse",
          "tsna",
          "ndtv",
          "knitr",
          "here",
          "igraph",
          "ggraph"
          )
  #check whether packages are installed and load them, if not get and then load
 package.check <- lapply(pack, FUN = function(x){
    if (!require(x, character.only = TRUE)) 
      {
        install.packages(x, dependencies = TRUE)
        library(x, character.only = TRUE)
      }
    })
   #search() #verify that packages are loaded

```

Load data & get variables to plot
```{r}
#simple transition matrix has each state, with start and end date and duration
community_states<-readRDS(here("./output/frog_CommunityStates.rds"))

#calculate average node duration and record species in a state
node_data_overall <- community_states %>%
  group_by(pond, cycle, state) %>%
  summarise(
    avg_duration = mean(duration, na.rm = TRUE),
    .groups = 'drop'
  )

# Calculate Edge Weights (Frequency) per Group
edge_data_grouped <- community_states %>%
  group_by(pond, cycle) %>%
  mutate(
    from = state,
    to = lead(state)
  ) %>%
  ungroup() %>%
  filter(!is.na(to)) %>%
  group_by(pond, cycle, from, to) %>%
  summarise(
    frequency = n(),
    .groups = 'drop'
  )

# Separate the data frames into a list of nested data frames (one per group)
nested_dataEW <- edge_data_grouped %>%
  group_by(pond, cycle) %>%
  nest(edges = c(from, to, frequency))

node_data_nested <- node_data_overall %>%
  group_by(pond, cycle) %>%
  nest(nodes = c(state, avg_duration))

# Join the nested node and edge data
graph_data_list <- inner_join(nested_dataEW, node_data_nested, by = c("pond", "cycle"))

```

Graph the networks
```{r}
# Function to Create and Plot Graph
create_and_plot_graphD <- function(pond, cycle, edges, nodes) {
  # Prep the vertices data frame for igraph
  all_states <- unique(c(nodes$state, edges$from, edges$to))
  vertices_df <- data.frame(state = all_states) %>%
    left_join(nodes, by = "state") %>%
    replace_na(list(avg_duration = 0))

  # Create igraph object
  graph <- graph_from_data_frame(
    d = edges,
    vertices = vertices_df,
    directed = TRUE
  )

  # Plotting parameters
  size_scale <- 25

  p <- ggraph(graph, layout = 'nicely') +
    geom_edge_fan(
      aes(alpha = frequency, width = log(frequency), color="#F05039"),
      arrow = arrow(length = unit(3, 'mm')),
      end_cap = circle(5, 'mm')
    ) +
    scale_edge_width_continuous(range = c(0.5, 2)) +
    scale_edge_alpha_continuous(range = c(0.5, 1)) +
    #scale_edge_color_distiller(palette = "Reds", direction = 1) +
    
    geom_node_point(
      aes(size = log(avg_duration)*2),
      color = "skyblue"
    ) +
    geom_node_text(
      aes(label = name),
      repel = TRUE,
      fontface = "bold",
      size = 4
    ) +
    theme_graph() +
    theme(legend.position = "none")

  
  # Return the plot object
  return(p)
}

#Apply function for  just for one pond cycle combination
Target_pond<-4
Target_cycle<-23
single_plot_data <- graph_data_list %>%
  filter(pond == Target_pond, cycle ==Target_cycle)

print(single_plot_data$plot)

# Apply the function to all groups
graph_data_list <- graph_data_list %>%
  mutate(
    plot = pmap(
      .l = list(pond, cycle, edges, nodes),
      .f = create_and_plot_graphD
    )
  )
# View/Save the Plots
# You can now access and display/save individual plots:
# graph_data_list$plot[[1]] # Displays the first plot
# ggsave(filename = "Pond1_Cycle1_Graph.png", plot = graph_data_list$plot[[1]])

# Example: Display the first plot
print(graph_data_list$plot[[150]])

```
Alternatively, plot each transition
```{r}
#select focal pond and cycle
Focal_pond<-4
Focal_cycle<-23
testP<-community_transitions%>%
  subset(pond==Focal_pond & cycle==Focal_cycle)%>%
  select(starting_state,end_state,day)

#get duration of states
state_durations<-testP%>%
  arrange(day)%>%
  group_by(starting_state)%>%
  summarize(duration=mean(c(diff(day),1)),.groups="drop")

#create nodes dataframe with duration info
nodes <- data.frame(name = unique(c(testP$starting_state, testP$end_state)))%>%
  left_join(state_durations, by = c("name" = "starting_state")) %>%
  replace_na(list(duration = 1))


state_durations <- community_transitions_g3 %>%
  subset(pond == 8 & cycle == 10) %>%
  arrange(day) %>%
  group_by(starting_state) %>%
  summarize(avg_duration = mean(c(diff(day), 1)), .groups = 'drop')

# Create network with node attributes
test.network <- graph_from_data_frame(d = testP, vertices = nodes, directed = TRUE)

            
#visualize newtork with node size by duration and edge by date

network_plot<-ggraph(test.network, layout = layout_as_star(
    test.network, center = which.max(degree(test.network))[[1]]
)) +
    geom_edge_fan(
      aes(color = day),
        arrow = arrow(length = unit(0.12, "inches"), type="closed", angle=25),
      end_gap=circle(0.25,"inches"),
        end_cap = circle(0.1, "inches"),
        strength = 3,
        spread=2,
        width = 1.2,
        alpha = .8
    ) +
    geom_node_point(aes(size=duration), color="#5B2C6F") +
    #geom_node_text(aes(label = name), size = 2, repel = TRUE) +
    scale_edge_color_gradient(
        low = "#2C3E50",
        high = "#A8E6CF"
    ) +
  scale_size_continuous(range=c(3,12))+
  theme_void()
network_plot
ggsave(here("Figures","network_plot.png"),plot=network_plot,dpi=600,width = 12, height = 8, units = "cm")
```
Now let's to this for all years for a specific pond
```{r}
# Function to create network plot for a given pond and cycle
# Get all cycles for pond x
focalpond<-4

#get max duration to scale sizes in plots consistently
max_dur <- community_transitions %>%
  subset(pond==focalpond)%>% #if we want to only plot a susbet of ponds
  group_by(pond, cycle, starting_state) %>%
  summarize(avg_duration = mean(c(diff(day), 1)), .groups = 'drop') %>%
  pull(avg_duration) %>%
  max(na.rm = TRUE)

plot_network <- function(pond_num, cycle_num) {
  # Calculate state duration
  state_durations <- community_transitions %>%
    subset(pond == pond_num & cycle == cycle_num) %>%
    arrange(day) %>%
    group_by(starting_state) %>%
    summarize(avg_duration = mean(c(diff(day), 1)), .groups = 'drop')
  
  # Prepare edge data
  testP <- community_transitions %>%
    subset(pond == pond_num & cycle == cycle_num, 
           select = c(starting_state, end_state, day))
  
  # Create nodes dataframe
  nodes <- data.frame(name = unique(c(testP$starting_state, testP$end_state))) %>%
    left_join(state_durations, by = c("name" = "starting_state")) %>%
    replace_na(list(avg_duration = 1))
  
  # Create network
  test.network <- graph_from_data_frame(d = testP, vertices = nodes, directed = TRUE)
  
  # Create plot
  ggraph(test.network, layout = layout_as_star(
      test.network, center = which.max(degree(test.network))[[1]]
  )) +
      geom_edge_fan(
        aes(edge_colour = day),
          arrow = arrow(length = unit(0.15, "inches"), type = "open"),
          end_cap = circle(0.25, "inches"),
          strength = 1.8,
          spread = 3,
          width = 1.2,
          alpha = 0.6
      ) +
          #geom_node_text(aes(label = name), size = 2.5, repel = TRUE) +
      scale_edge_color_gradient(
          low = "#2C3E50",
          high = "#A8E6CF",
          name = "Day",
          limits=c(1,365)
          ) +
      geom_node_point(aes(size = avg_duration), 
                      fill = "#5B2C6F",
                      color = "white",
                      stroke = 1,
                      shape = 21) +
      scale_size_continuous(
        range = c(3, 10), 
        name = "Avg Duration",
        limits=c(1,max_dur)) +
      theme_void() + 
      labs(title = paste("Cycle", cycle_num))
}


cycles_pond1 <- community_transitions %>%
  filter(pond == focalpond) %>%
  pull(cycle) %>%
  unique() %>%
  sort()

# Create all plots
plot_list <- lapply(cycles_pond1, function(c) plot_network(focalpond, c))

# Combine using patchwork
library(patchwork)
wrap_plots(plot_list, ncol = 5)  # Adjust ncol as needed

#plot a  subset 3 dry years (top row) vs 3 normal years

extY_comp_plot<-wrap_plots(plot_list[c(6,11,22,1 ,16,24)], ncol = 3)+
  plot_layout(guide="collect")&
  theme(legend.position="right",
        legend.box.just="right",
        legend.justification="right")
#save all six plots
ggsave(here("Figures","extremeY_comp_network_plot.png"),plot=extY_comp_plot,dpi=600,width = 12, height = 8, units = "in")

#plot each individually
P1Y1<-wrap_plots(plot_list[c(1)])
P1Y6<-wrap_plots(plot_list[c(6)])
P1Y11<-wrap_plots(plot_list[c(11)])
P1Y16<-wrap_plots(plot_list[c(16)])
P1Y22<-wrap_plots(plot_list[c(22)])
P1Y24<-wrap_plots(plot_list[c(24)])

ggsave(here("Figures","network_plot_p1y1.png"),plot=P1Y1,dpi=600,width = 6, height = 4, units = "in")


```






playing around with tna and qgraph, but it's not ideal
```{r}
test.network<-community_networks_weighted[["pond_3_cycle_11"]]
tna.net<-tna(test.network)

original_weight_matrix <- as.matrix(
    igraph::as_adjacency_matrix(
        graph = test.network,
        type = "both",        # Include all edges (or "directed" if preferred)
        attr = "weight", # <-- **CRITICAL: Uses the edge attribute value**
        sparse = FALSE        # Returns a standard R matrix
    )
)
tna.net$weights<-(original_weight_matrix)
plot(tna.net, 
     edge.labels=FALSE,
     cut=2,
     edge.wdith.max=6,
     edge.width.min=1)
```

