---
title: "Robustness analysis"
author: "Volker Rudolf"
date: "2026-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Goals
We will examine how climate variables are linked to different aspects of community state transitions. THis is focused on frog calling data from 8 different ponds across 22 years. Climate data is taking from local weather stations, and frog calling data was converted to transition matrix indicating how community states (i.e. a specific combination of species) transition over time within a year.

## Load packages & data
```{r, warning=FALSE, message=FALSE}
 #Load packages
  #list of packages that are needed
  pack<-c("tidyverse",
          "knitr",
          "here",
          "car",
          "lme4",
          "flextable",
          "stringr",
          "ggpubr",
          "broom.mixed",
          "lmerTest",
          "rlang",
          "ggrepel"
          )
  #check whether packages are installed and load them, if not get and then load
 package.check <- lapply(pack, FUN = function(x){
    if (!require(x, character.only = TRUE)) 
      {
        install.packages(x, dependencies = TRUE)
        library(x, character.only = TRUE)
      }
    })
   #search() #verify that packages are loaded

```


Let's load the community matrix and climate data (see "Community transition code.Rdm" and "Climate data preparation.Rdm")
```{r get data, message=F}
#get summary metrics of the transition matrix
state_connectance_metrics<-read_rds(here("./output/frog_transition_matrix_metrics.rds"))#raw data transition matrix 
#we have 3 additional version of the transition matrix that ignore a gap of 1, 2, or 3 day in the calling of a species and treat it as a continuous period. 
state_connectance_metrics_g1<-read_rds(here("output","frog_transition_matrix_metrics_g1.rds")) 
state_connectance_metrics_g2<-read_rds(here("output","frog_transition_matrix_metrics_g2.rds"))
state_connectance_metrics_g3<-read_rds(here("output","frog_transition_matrix_metrics_g3.rds"))

#get annual climate data summary metrics 
sumclima<-read_csv(here("output","frog_climate_data_annual_scaled.csv"))%>%subset(cycle<25)

#combine climate and network statistics
netclim <- reduce(list(sumclima, state_connectance_metrics), full_join) %>%
  select(pond, cycle, Pond_Cycle, everything())
netclim_g1 <- reduce(list(sumclima, state_connectance_metrics_g1), full_join) %>%
  select(pond, cycle, Pond_Cycle, everything())
netclim_g2 <- reduce(list(sumclima, state_connectance_metrics_g2), full_join) %>%
  select(pond, cycle, Pond_Cycle, everything())
netclim_g3 <- reduce(list(sumclima, state_connectance_metrics_g3), full_join) %>%
  select(pond, cycle, Pond_Cycle, everything())

```

##PCA analysis
```{r metric_WeatherPCA}
climatePCA<-read_csv(here("./output/frog_WeatherPCA.csv"))
netclimPC<-left_join(netclim,climatePCA)
netclimPC_g1<-left_join(netclim_g1,climatePCA)
netclimPC_g2<-left_join(netclim_g2,climatePCA)
netclimPC_g3<-left_join(netclim_g3,climatePCA)

#get variables to analyze
variable.for.extanalysis<-c("Unique_Nodes","Connectance", "modularity","betweenness", "mean_Distance", "freqStateAv",    "meanDurAv","skewDur","TotalEdges","ave_Edge_weight","skewEdge", "B","time_memory","event.rate")

run_pca_analysis <- function(data_input, vars_to_analyze, file_name) {
  
  # 1. Internal Label Map
  # Add any other "pretty names" here as needed.
  variable_labels <- c(
    "Connectance" = "Connectance (Log)",
    "mean_Distance" = "Mean Distance (Log)"
  )

  # 2. Scale and log-transform
  # We log these two specifically, then scale all variables in the list
  data_scaled <- data_input %>% 
    mutate(across(c("Connectance", "mean_Distance"), ~log(. + 0.001))) # added small offset in case of zeros
  
  data_scaled[vars_to_analyze] <- lapply(data_scaled[vars_to_analyze], scale)
  
  # 3. Analyze the data (LMMs)
  resPCA <- map_dfr(vars_to_analyze, function(response) {
    formula_str <- paste(response, "~ PC1 + PC2 + (1 | pond)")
    model <- lmerTest::lmer(as.formula(formula_str), data = data_scaled)
    
    broom.mixed::tidy(model, effects = "fixed", conf.int = TRUE) %>%
      mutate(response = response) %>%
      filter(!str_detect(term, "(Intercept)"))
  })
  
  # 4. Create significance pattern table
  sig_tablePCA <- resPCA %>%
    # FIXED LOGIC: If the variable has a label, use it. Otherwise, keep the original name.
    mutate(response = ifelse(response %in% names(variable_labels), 
                             variable_labels[response], 
                             response)) %>% 
    # Ensure uniqueness before pivoting
    distinct(response, term, .keep_all = TRUE) %>% 
    select(response, term, p.value, statistic) %>%
    mutate(
      p.value = round(p.value, 4),
      statistic = round(statistic, 2)
    ) %>%
    pivot_wider(
      names_from = term, 
      values_from = c(statistic, p.value),
      names_glue = "{term}_{.value}",
      names_vary = "slowest"
    )
  
  # 5. Create flextable and Save
  sig_table_flextable <- sig_tablePCA %>%
    flextable() %>%
    set_caption(paste("Table: Significance Patterns -", file_name)) %>%
    theme_booktabs() %>%
    bold(part = "header") %>%
    bold(i = ~ PC1_p.value < 0.05, j = ~ PC1_p.value + PC1_statistic) %>% # Bonus: bold significant PC1
    bold(i = ~ PC2_p.value < 0.05, j = ~ PC2_p.value + PC2_statistic) %>% # Bonus: bold significant PC2
    align(align = "center", part = "all") %>%
    autofit()
  
  save_as_docx(sig_table_flextable, path = here("Output", paste0(file_name, ".docx")))
  
  return(resPCA)
}
results_1 <- run_pca_analysis(netclimPC_g1, variable.for.extanalysis, "AnovaPCATable_gap_1")
results_2 <- run_pca_analysis(netclimPC_g2, variable.for.extanalysis, "AnovaPCATable_gap_2")
results_3 <- run_pca_analysis(netclimPC_g3, variable.for.extanalysis, "AnovaPCATable_gap_3")
```

##PCA analysis effect size plot
```{r}
variable.for.extanalysis<-c("Unique_Nodes","Connectance", "modularity","betweenness", "mean_Distance", "freqStateAv",    "meanDurAv","skewDur","TotalEdges","ave_Edge_weight","skewEdge", "time_memory","B","event.rate")

#scale data to be comparable but first log-transform two variables to improve lmer model fit
currentPCdataS<-currentPCdata%>% 
  mutate(across(c("Connectance", "mean_Distance"), log))

currentPCdataS[variable.for.extanalysis] <- lapply(currentPCdata[variable.for.extanalysis], scale)
#analyze the data

resPCA <- map_dfr(variable.for.extanalysis, function(response) {
  # Fit the mixed model for this response
  model <- lmerTest::lmer(as.formula(paste(response,"~ PC1+PC2 +(1 | pond)")), data = currentPCdataS)
  

  # Tidy model output: fixed effects only, with confidence intervals
  broom.mixed::tidy(model, effects = "fixed", conf.int = TRUE) %>%
    mutate(response = response)%>%
    filter(!str_detect(term,"(Intercept)"))
  }
  )

# Create significance pattern table
sig_tablePCA <- resPCA %>%
  mutate(response=variable_labels[response])%>%
  select(response, term,p.value, statistic) %>%
  mutate(
    p.value = round(p.value, 4),
    statistic=round(statistic,2)
    ) %>%
  pivot_wider(
    names_from = term, 
    values_from = c(statistic,p.value),
    names_glue="{term}_{.value}",
    names_vary="slowest"
  )



###plot
resPCA_prep <- resPCA %>%
  mutate(significant = p.value < 0.05)

# Add a category column to your data based on your grouping
resPCA_prep <- resPCA_prep %>%
  mutate(category = case_when(
    response %in% c("Unique_Nodes",  "Connectance", 
                    "betweenness", "mean_Distance", "modularity") ~ "Network",
    response %in% c("meanDurAv", "freqStateAv", "skewDur") ~ "Node",
    response %in% c("ave_Edge_weight","TotalEdges", "skewEdge") ~ "Edge",
    response %in% c("event.rate", "B", "time_memory") ~ "Spacing",
    TRUE ~ "Other"
  )) %>%
  # Set category order (top to bottom as they appear in your list)
  mutate(category = factor(category, 
                          levels = c("Network", "Node", "Edge", "Spacing"))) %>%
  # Set response variable order within categories
  mutate(response = factor(response, levels = my_grouped_y_order))


# Create background shading by category (spans entire category height)
background_data <- data.frame(
  category = factor(c("Network", "Node", "Edge", "Spacing"),
                   levels = c("Network", "Node", "Edge", "Spacing")),
  fill_color = c("white", "grey95", "white", "grey95")
) %>%
  # Expand to include both terms
  tidyr::crossing(term = unique(resPCA_prep$term))

# Create the plot with faceting by category
coef.plotPCA <- resPCA_prep %>%
  ggplot(aes(x = estimate, y = response, color = term)) +
  # Add background rectangles that span entire category
  geom_rect(data = background_data,
            aes(fill = fill_color),
            xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf,
            inherit.aes = FALSE, alpha = 1) + 
  scale_fill_identity() +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray50", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high, alpha = significant), 
                 height = 0.0, linewidth = 1) +
  geom_point(aes(alpha = significant), size = 3) +
  scale_alpha_manual(values = c("FALSE" = 0.3, "TRUE" = 1), guide = "none") +
  # Facet by both category (rows) and term (columns)
  facet_grid(category ~ term, 
             scales = "free_y",      # Free y-axis for each category
             space = "free_y",       # Adjust panel height based on number of items
             switch = "y",           # Put category labels on the left
             labeller = labeller(
               term = c("PC1" = "Weather PCA1", "PC2" = "Weather PCA2"),
               category = label_value)) +
  scale_y_discrete(
    drop = TRUE,  # Drop unused factor levels in each facet
    labels = variable_labels) +
  labs(x = "Standardized coefficients and 95% CI", y = "") + 
  scale_color_manual(
    values = c("PC1" = "#33B8CC", "PC2" = "#CCAA33"), 
    guide = "none") +
  theme_pubr(base_size = 16) +
  theme(
    strip.background = element_blank(),  # Remove box around facet labels
    strip.text.x = element_text(size = 18, face = "bold"),
    strip.text.y.left = element_text(size = 16, face = "bold", angle = 90),  # Rotate category labels
    strip.placement = "outside",
    panel.spacing = unit(0.3, "lines"),      # Space between categories
    panel.border = element_blank(),  # Remove boxes around panels
    axis.text.x = element_text(size = 14), 
    axis.text.y = element_text(size = 16), 
    axis.title = element_text(size = 18)
  )

coef.plotPCA
ggsave(here("Figures","PCAWeather.effect.png"),plot=coef.plotPCA,dpi=600,width = 30, height = 15, units = "cm")
```

```{r}
run_full_pca_workflow <- function(data_input, vars_to_analyze, file_name) {
  
  # 1. Internal Mappings
  # Define your labels and the y-axis order here
  variable_labels <- c(
    "Connectance" = "Connectance (Log)",
    "mean_Distance" = "Mean Distance (Log)"
  )
  
  # Set the order for the Y-axis (Top to Bottom)
  y_order <- c("Unique_Nodes", "Connectance", "modularity", "betweenness", "mean_Distance",
               "freqStateAv", "meanDurAv", "skewDur", "TotalEdges", "ave_Edge_weight", 
               "skewEdge", "B", "time_memory", "event.rate")

  # 2. Data Transformation
  data_scaled <- data_input %>% 
    mutate(across(c("Connectance", "mean_Distance"), ~log(. + 0.001)))
  
  data_scaled[vars_to_analyze] <- lapply(data_scaled[vars_to_analyze], scale)
  
  # 3. Mixed Model Analysis
  resPCA <- map_dfr(vars_to_analyze, function(response) {
    model <- lmerTest::lmer(as.formula(paste(response, "~ PC1 + PC2 + (1 | pond)")), data = data_scaled)
    broom.mixed::tidy(model, effects = "fixed", conf.int = TRUE) %>%
      mutate(response = response) %>%
      filter(!str_detect(term, "(Intercept)"))
  })

  # 4. Plot Preparation
  resPCA_prep <- resPCA %>%
    mutate(
      significant = p.value < 0.05,
      # Category grouping logic
      category = case_when(
        response %in% c("Unique_Nodes", "Connectance", "betweenness", "mean_Distance", "modularity") ~ "Network",
        response %in% c("meanDurAv", "freqStateAv", "skewDur") ~ "Node",
        response %in% c("ave_Edge_weight","TotalEdges", "skewEdge") ~ "Edge",
        response %in% c("event.rate", "B", "time_memory") ~ "Spacing",
        TRUE ~ "Other"
      ),
      category = factor(category, levels = c("Network", "Node", "Edge", "Spacing")),
      response = factor(response, levels = rev(y_order)) # rev() so top of list is top of plot
    )

  # Background Shading Data
  background_data <- expand.grid(
    category = factor(c("Network", "Node", "Edge", "Spacing"), levels = levels(resPCA_prep$category)),
    term = unique(resPCA_prep$term)
  ) %>%
    mutate(fill_color = ifelse(category %in% c("Node", "Spacing"), "grey95", "white"))

  # 5. Generate Plot
  coef_plot <- ggplot(resPCA_prep, aes(x = estimate, y = response, color = term)) +
    geom_rect(data = background_data, aes(fill = fill_color),
              xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf,
              inherit.aes = FALSE, alpha = 1) + 
    scale_fill_identity() +
    geom_vline(xintercept = 0, linetype = "solid", color = "gray50", linewidth = 0.5) +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high, alpha = significant), 
                   height = 0, linewidth = 1) +
    geom_point(aes(alpha = significant), size = 3) +
    scale_alpha_manual(values = c("FALSE" = 0.3, "TRUE" = 1), guide = "none") +
    facet_grid(category ~ term, scales = "free_y", space = "free_y", switch = "y",
               labeller = labeller(term = c("PC1" = "Weather PCA1", "PC2" = "Weather PCA2"))) +
    scale_y_discrete(labels = function(x) ifelse(x %in% names(variable_labels), variable_labels[x], x)) +
    scale_color_manual(values = c("PC1" = "#33B8CC", "PC2" = "#CCAA33"), guide = "none") +
    labs(x = "Standardized coefficients and 95% CI", y = "") +
    theme_pubr(base_size = 16) +
    theme(
      strip.background = element_blank(),
      strip.text.x = element_text(size = 18, face = "bold"),
      strip.text.y.left = element_text(size = 16, face = "bold", angle = 0),
      strip.placement = "outside",
      panel.spacing = unit(0.3, "lines"),
      axis.text.y = element_text(size = 12)
    )

  # 6. Save Plot
  ggsave(here("Figures", paste0("EffectSize_", file_name, ".png")), 
         plot = coef_plot, dpi = 600, width = 30, height = 20, units = "cm")
  
  return(coef_plot)
}

plot1 <- run_full_pca_workflow(netclimPC_g1, variable.for.extanalysis, "Gap_1")
plot2 <- run_full_pca_workflow(netclimPC_g2, variable.for.extanalysis, "Gap_2")
plot3 <- run_full_pca_workflow(netclimPC_g3, variable.for.extanalysis, "Gap_3")

```


For extreme weather
```{r}
run_extreme_clim_workflow <- function(netclim_data, vars_to_analyze, file_suffix) {
  
  # 1. Internal Settings & Labels
  variable_labels <- c(
    "Connectance" = "Connectance (Log)",
    "mean_Distance" = "Mean Distance (Log)"
  )
  
  y_order <- c("Unique_Nodes", "Connectance", "modularity", "betweenness", "mean_Distance",
               "freqStateAv", "meanDurAv", "skewDur", "TotalEdges", "ave_Edge_weight", 
               "skewEdge", "B", "time_memory", "event.rate")

  # 2. Data Preparation
  # Load and prep the raw climate data
  sumclima_raw <- read.csv(here("./output/frog_climate_data_annualsum.csv")) %>%
    mutate(across(c(exthighTemp, extlongHeat, extDry, extDryT, extHot), factor)) %>%
    select(-burstinessRain)

  # Join and Scale
  netclimR <- left_join(netclim_data, sumclima_raw, by = "cycle")
  netclimR[vars_to_analyze] <- lapply(netclimR[vars_to_analyze], scale)

  # 3. Analyze Data (Mixed Models)
  res <- map_dfr(vars_to_analyze, function(response) {
    # Using your specific extreme event formula
    model <- lmerTest::lmer(as.formula(paste(response, "~ extHot + extDryT + (1 | pond)")), data = netclimR)
    
    broom.mixed::tidy(model, effects = "fixed", conf.int = TRUE) %>%
      mutate(response = response) %>%
      filter(!str_detect(term, "(Intercept)"))
  })

  # 4. Plot Preparation
  res_prepE <- res %>%
    filter(!response %in% c("No_Spec")) %>%
    mutate(
      significant = p.value < 0.05,
      category = case_when(
        response %in% c("Unique_Nodes", "Connectance", "betweenness", "mean_Distance", "modularity") ~ "Network",
        response %in% c("meanDurAv", "freqStateAv", "skewDur") ~ "Node",
        response %in% c("ave_Edge_weight", "TotalEdges", "skewEdge") ~ "Edge",
        response %in% c("event.rate", "B", "time_memory") ~ "Spacing",
        TRUE ~ "Other"
      ),
      category = factor(category, levels = c("Network", "Node", "Edge", "Spacing")),
      response = factor(response, levels = rev(y_order))
    )

  # Background Shading Data
  background_data <- tidyr::crossing(
    category = factor(c("Network", "Node", "Edge", "Spacing"), levels = levels(res_prepE$category)),
    term = unique(res_prepE$term)
  ) %>%
    mutate(fill_color = ifelse(category %in% c("Node", "Spacing"), "grey95", "white"))

  # 5. Create the Plot
  coef_plotE <- res_prepE %>%
    ggplot(aes(x = estimate, y = response, color = term)) +
    geom_rect(data = background_data, aes(fill = fill_color),
              xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf,
              inherit.aes = FALSE, alpha = 1) + 
    scale_fill_identity() +
    geom_vline(xintercept = 0, linetype = "solid", color = "gray50", linewidth = 0.5) +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high, alpha = significant), 
                   height = 0.0, linewidth = 1) +
    geom_point(aes(alpha = significant), size = 3) +
    scale_alpha_manual(values = c("FALSE" = 0.3, "TRUE" = 1), guide = "none") +
    facet_grid(category ~ term, scales = "free_y", space = "free_y", switch = "y",
               labeller = labeller(
                 term = c("extDryTYes" = "Extreme Dry", "extHotYes" = "Extreme Hot"),
                 category = label_value)) +
    scale_y_discrete(labels = function(x) ifelse(x %in% names(variable_labels), variable_labels[x], x)) +
    scale_color_manual(values = c("extDryTYes" = "#A65F32", "extHotYes" = "#CC3333"), guide = "none") +
    labs(x = "Standardized coefficients and 95% CI", y = "") +
    theme_pubr(base_size = 16) +
    theme(
      strip.background = element_blank(),
      strip.text.x = element_text(size = 18, face = "bold"),
      strip.text.y.left = element_text(size = 16, face = "bold", angle = 90),
      strip.placement = "outside",
      panel.spacing = unit(0.3, "lines"),
      axis.text.y = element_text(size = 12)
    )

  # 6. Save Plot
  ggsave(here("Figures", paste0("ExtremeClim_Effect_", file_suffix, ".png")), 
         plot = coef_plotE, dpi = 600, width = 30, height = 15, units = "cm")
  
  return(res) # Returns the results dataframe for further use
}
results_ext_B <- run_extreme_clim_workflow(netclim, variable.for.extanalysis, "Gap_0")
results_ext_1 <- run_extreme_clim_workflow(netclim_g1, variable.for.extanalysis, "Gap_1")
results_ext_2 <- run_extreme_clim_workflow(netclim_g1, variable.for.extanalysis, "Gap_2")
results_ext_3 <- run_extreme_clim_workflow(netclim_g1, variable.for.extanalysis, "Gap_3")

```

